---
format: gfm
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(targets)
tar_store <- NULL
if (dir.exists("_targets")) tar_store <- "_targets"

safe_read <- function(name) {
  if (!is.null(tar_store)) {
    val <- tryCatch(tar_read_raw(name, store = tar_store), error = function(e) NULL)
    if (!is.null(val)) return(val)
  }
  rds <- file.path("inst/extdata/vignettes", paste0(name, ".rds"))
  if (file.exists(rds)) return(readRDS(rds))
  NULL
}
```

# coMMpass

Reproducible analysis of the MMRF CoMMpass (Clinical Outcomes in MM to Personal Assessment of Genetic Profile) multiple myeloma study data using R and Nix.

## Overview

The [CoMMpass study](https://themmrf.org/finding-a-cure/personalized-treatment-approaches/) is a landmark longitudinal genomic-clinical study of **1,143 newly diagnosed multiple myeloma patients** collected between 2011-2016 with 8-year follow-up. This project provides:

- Reproducible Nix environment with all required R/Bioconductor packages
- Analysis workflows for RNA-seq, single-cell, and survival analysis
- Direct access to CoMMpass data via GDC API and AWS S3 (open access)

## Usage

### Quick Start (without Nix)

```{r install-code}
#| results: asis
code <- safe_read("readme_install_code")
if (!is.null(code)) cat("```r\n", code, "\n```\n", sep = "")
```

#### Pipeline status

```{r pipeline-status}
#| results: asis
ps <- safe_read("readme_pipeline_summary")
if (!is.null(ps)) {
  cat(paste0("The pipeline has **", ps$n_targets, " targets** ",
             "(total build time: ", ps$total_time, ").\n\n"))
  cat("Top 5 targets by build time:\n\n")
  cat(knitr::kable(ps$top5, format = "pipe"), sep = "\n")
}
```

#### Example: read a result

```{r km-structure}
#| results: asis
km_str <- safe_read("readme_km_structure")
if (!is.null(km_str)) {
  cat("```r\nkm <- tar_read(km_overall)\nstr(km, max.level = 1)\n")
  cat(paste0("#> ", km_str), sep = "\n")
  cat("\n```\n")
}
```

> **Note:** The pipeline downloads ~3.6 GB of RNA-seq data from GDC on first
> run. Subsequent runs skip downloads (`cue = "never"`). The `config` target
> sets `sample_limit = 200` by default — edit `R/tar_plans/plan_data_acquisition.R`
> to change this.

### Quick Start (with Nix — fully reproducible)

```{r nix-code}
#| results: asis
code <- safe_read("readme_nix_code")
if (!is.null(code)) cat("```bash\n", code, "\n```\n", sep = "")
```

### What Happens

1. **`default.sh`** checks if `default_dev.nix` needs regeneration
2. Runs **`default.R`** which:
   - Extracts package dependencies from `DESCRIPTION` (single source of truth)
   - Calls `rix::rix()` to generate `default_dev.nix`
3. Builds the Nix environment with `nix-build`
4. Creates a GC root symlink (prevents garbage collection)
5. Enters an interactive shell with all packages

## Data Access

| Access Level | Data Type | Requirements | Used by this pipeline |
|---|---|---|---|
| **[GDC Data Portal](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS)** | Clinical, RNA-seq (995 cases) | None (open access) | Yes |
| **[AWS Open Data](https://registry.opendata.aws/mmrf-commpass/)** | RNA-seq gene expression | None (AWS CLI) | Yes |
| **[MMRF Researcher Gateway](https://research.themmrf.org/)** | FISH, PFS, treatment response | Free registration (pending) | No (12 targets blocked) |
| **dbGaP Controlled** | Raw sequences (BAM/FASTQ) | Institutional IRB | No |

## Project Structure

```{r project-tree}
#| results: asis
pt <- safe_read("readme_project_tree")
if (!is.null(pt)) {
  cat("```\n")
  for (i in seq_len(nrow(pt$tree))) {
    cat(sprintf("%-30s # %s\n", pt$tree$Path[i], pt$tree$Description[i]))
  }
  cat("```\n\n")
  cat(paste0("**", pt$counts$r_files, "** R source files, ",
             "**", pt$counts$vignettes, "** vignettes, ",
             "**", pt$counts$rds, "** pre-computed RDS, ",
             "**", pt$counts$tests, "** test files.\n"))
}
```

## Key References

- [Keats JJ et al. Blood 2013](https://doi.org/10.1182/blood.V122.21.532.532) — CoMMpass interim analysis
- [Nature Genetics 2024](https://www.nature.com/articles/s41588-024-01853-0) — Molecular profiling of 1,143 patients
- [Nature Cancer 2025](https://www.nature.com/articles/s43018-025-01072-4) — Bone marrow immune cell atlas

## Glossary

See the [Glossary](https://johngavin.github.io/coMMpass-analysis/articles/glossary.html) vignette for definitions of terms used throughout this project.

## Requirements

- [Nix](https://nixos.org/download.html) package manager (for reproducible environment)
- [Cachix](https://cachix.org/) (optional, for faster builds)

```{r cachix-code}
#| results: asis
code <- safe_read("readme_cachix_code")
if (!is.null(code)) cat("```bash\n", code, "\n```\n", sep = "")
```

## License

MIT
