---
title: "01. Data Acquisition & Status"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01. Data Acquisition & Status}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
library(targets)
library(dplyr)
library(ggplot2)
library(stringr)
has_plotly <- requireNamespace("plotly", quietly = TRUE)
has_DT <- requireNamespace("DT", quietly = TRUE)
has_SE <- requireNamespace("SummarizedExperiment", quietly = TRUE)
if (has_plotly) library(plotly)
if (has_DT) library(DT)

# Source utility functions
if (file.exists("../R/utils.R")) {
  source("../R/utils.R")
} else if (file.exists("R/utils.R")) {
  source("R/utils.R")
}

key <- "TAR_PROJECT"
proj_dir <- Sys.getenv(key)
targets_available <- FALSE

if (proj_dir != "" && dir.exists(file.path(proj_dir, "_targets"))) {
  tar_config_set(store = file.path(proj_dir, "_targets"))
  targets_available <- TRUE
} else if (dir.exists("_targets")) {
  targets_available <- TRUE
}

# Helper to safely read targets (returns NULL if not available)
safe_tar_read <- function(name) {
  if (!targets_available) return(NULL)
  tryCatch(tar_read(name), error = function(e) NULL)
}

# Resolve project root for file paths
project_root <- if (proj_dir != "") proj_dir else "."
```

## Overview

This report summarizes the initial data acquisition step for the MMRF CoMMpass study. We access Open Access RNA-seq data from the GDC S3 bucket (`s3://gdc-mmrf-commpass-phs000748-2-open/`) via the `targets` pipeline.

## Pipeline Configuration

```{r config}
config <- safe_tar_read("config")

if (!is.null(config)) {
  config_df <- data.frame(
    Parameter = c("Project ID", "Sample Limit", "Data Directory",
                  "Results Directory", "Random Seed"),
    Value = c(config$project_id, config$sample_limit, config$data_dir,
              config$results_dir, config$seed),
    stringsAsFactors = FALSE
  )
  knitr::kable(config_df, caption = "Pipeline Configuration")
} else {
  cat("*Pipeline configuration not available. Run `tar_make(config)` to generate.*\n")
}
```

## RNA-seq Data

RNA-seq gene expression data is downloaded from GDC as a SummarizedExperiment object.

```{r rnaseq}
# Read the raw_rnaseq target (returns file path to RDS)
rnaseq_path <- safe_tar_read("raw_rnaseq")

# Try to load the SummarizedExperiment
se_data <- NULL
if (!is.null(rnaseq_path)) {
  # Resolve path relative to project root
  full_path <- if (file.exists(rnaseq_path)) {
    rnaseq_path
  } else {
    file.path(project_root, rnaseq_path)
  }

  if (file.exists(full_path) && has_SE) {
    se_data <- tryCatch(readRDS(full_path), error = function(e) NULL)
  }
}

if (!is.null(se_data) && has_SE) {
  counts <- SummarizedExperiment::assay(se_data, "counts")

  cat("### RNA-seq Data Summary\n\n")
  cat("- **Samples:**", ncol(counts), "\n")
  cat("- **Genes:**", format_with_commas(nrow(counts)), "\n")
  cat("- **Total counts:**", format_with_commas(sum(counts)), "\n")
  cat("- **Median counts per sample:**",
      format_with_commas(round(median(colSums(counts)))), "\n")
  cat("- **Genes detected (>0 in any sample):**",
      format_with_commas(sum(rowSums(counts) > 0)), "\n")
  cat("- **Sparsity (% zero entries):**",
      sprintf("%.1f%%", 100 * mean(counts == 0)), "\n\n")

  # Per-sample summary
  sample_stats <- data.frame(
    Sample = colnames(counts),
    Total_Counts = format_with_commas(colSums(counts)),
    Genes_Detected = format_with_commas(colSums(counts > 0)),
    Median_Count = round(apply(counts, 2, median)),
    Max_Count = format_with_commas(apply(counts, 2, max)),
    stringsAsFactors = FALSE
  )

  knitr::kable(sample_stats,
               col.names = c("Sample", "Total Counts", "Genes Detected",
                             "Median Count", "Max Count"),
               caption = "Per-Sample Expression Summary")

  # Count distribution plot
  cat("\n### Count Distribution\n\n")
  log_counts <- log10(counts + 1)
  mean_expr <- rowMeans(log_counts)

  if (has_plotly) {
    p <- plot_ly(
      x = mean_expr,
      type = "histogram",
      nbinsx = 50,
      marker = list(color = "steelblue"),
      name = "Gene Expression"
    ) %>%
      layout(
        title = "Distribution of Mean Gene Expression",
        xaxis = list(title = "log10(counts + 1)"),
        yaxis = list(title = "Number of Genes"),
        bargap = 0.05
      )
    p
  } else {
    hist_data <- data.frame(mean_expression = mean_expr)
    p <- ggplot(hist_data, aes(x = mean_expression)) +
      geom_histogram(bins = 50, fill = "steelblue", color = "white") +
      labs(title = "Distribution of Mean Gene Expression",
           x = "log10(counts + 1)", y = "Number of Genes") +
      theme_minimal()
    print(p)
  }
} else if (!is.null(rnaseq_path)) {
  cat("*RNA-seq file path found (`", rnaseq_path,
      "`) but data could not be loaded.*\n", sep = "")
} else {
  cat("*RNA-seq data not available. Run `tar_make(raw_rnaseq)` to download.*\n")
}
```

## Clinical Data

Clinical metadata retrieved from GDC provides patient demographics, disease characteristics, and outcomes.

```{r clinical}
# Try to load clinical data
# First try clinical_data_clean (from analysis pipeline)
clinical <- safe_tar_read("clinical_data_clean")

# Fall back to loading from the clinical_data directory path
if (is.null(clinical)) {
  clinical_dir <- safe_tar_read("clinical_data")

  if (!is.null(clinical_dir) && is.character(clinical_dir)) {
    # clinical_data target returns a directory path - load the RDS or CSV from it
    dir_path <- if (dir.exists(clinical_dir)) {
      clinical_dir
    } else {
      file.path(project_root, clinical_dir)
    }

    if (dir.exists(dir_path)) {
      rds_file <- file.path(dir_path, "clinical_data.rds")
      csv_file <- file.path(dir_path, "clinical_data.csv")

      if (file.exists(rds_file)) {
        clinical <- tryCatch(readRDS(rds_file), error = function(e) NULL)
      } else if (file.exists(csv_file)) {
        clinical <- tryCatch(read.csv(csv_file), error = function(e) NULL)
      }
    }
  }
}

if (!is.null(clinical) && is.data.frame(clinical)) {
  # Ensure unique column names
  names(clinical) <- make.unique(names(clinical))

  # Summary statistics
  cat("### Clinical Data Overview\n\n")
  cat("- **Total patients:**", format_with_commas(nrow(clinical)), "\n")
  cat("- **Number of variables:**", ncol(clinical), "\n")
  cat("- **Data completeness:**", sprintf("%.1f%%", 100 * mean(!is.na(clinical))), "\n\n")

  # Key demographic variables
  cols_to_show <- c("project_id", "submitter_id", "gender", "vital_status",
                    "age_at_diagnosis", "race", "ethnicity")
  cols_available <- intersect(cols_to_show, names(clinical))

  if (length(cols_available) > 0) {
    cat("### Patient Demographics\n\n")

    # Gender distribution
    if ("gender" %in% names(clinical)) {
      gender_table <- table(clinical$gender, useNA = "ifany")
      gender_df <- data.frame(
        Gender = names(gender_table),
        Count = as.numeric(gender_table),
        Percentage = sprintf("%.1f%%", 100 * as.numeric(gender_table) / nrow(clinical))
      )
      knitr::kable(gender_df, caption = "Gender Distribution")
      cat("\n")
    }

    # Vital status
    if ("vital_status" %in% names(clinical)) {
      vital_table <- table(clinical$vital_status, useNA = "ifany")
      vital_df <- data.frame(
        Status = names(vital_table),
        Count = as.numeric(vital_table),
        Percentage = sprintf("%.1f%%", 100 * as.numeric(vital_table) / nrow(clinical))
      )
      knitr::kable(vital_df, caption = "Vital Status Distribution")
      cat("\n")
    }

    # Age at diagnosis
    if ("age_at_diagnosis" %in% names(clinical) && is.numeric(clinical$age_at_diagnosis)) {
      # Check if values look like days (>365) or years (<120)
      age_values <- clinical$age_at_diagnosis[!is.na(clinical$age_at_diagnosis)]
      is_days <- max(age_values, na.rm = TRUE) > 120
      age_years <- if (is_days) age_values / 365.25 else age_values

      cat("### Age at Diagnosis (years)\n\n")
      cat("- **Median:**", sprintf("%.1f", median(age_years, na.rm = TRUE)), "\n")
      cat("- **Range:**", sprintf("%.1f - %.1f",
                                   min(age_years, na.rm = TRUE),
                                   max(age_years, na.rm = TRUE)), "\n\n")

      if (has_plotly && length(age_years) > 0) {
        p <- plot_ly(
          x = age_years,
          type = "histogram",
          nbinsx = 30,
          marker = list(color = "steelblue"),
          name = "Age Distribution"
        ) %>%
          layout(
            title = "Age at Diagnosis Distribution",
            xaxis = list(title = "Age (years)"),
            yaxis = list(title = "Number of Patients"),
            bargap = 0.1
          )
        p
      }
    }

    # Sample of clinical data
    cat("\n### Sample Clinical Data\n\n")
    display_clinical <- clinical[seq_len(min(10, nrow(clinical))), cols_available, drop = FALSE]

    if (has_DT) {
      DT::datatable(display_clinical,
                options = list(pageLength = 5, scrollX = TRUE),
                caption = "Sample of Clinical Data")
    } else {
      knitr::kable(display_clinical, caption = "Sample of Clinical Data")
    }
  }

  # Missing data analysis
  cat("\n### Data Completeness Analysis\n\n")
  missing_summary <- data.frame(
    Variable = names(clinical),
    Missing_Count = colSums(is.na(clinical)),
    Missing_Percent = sprintf("%.1f%%", 100 * colSums(is.na(clinical)) / nrow(clinical)),
    stringsAsFactors = FALSE
  ) %>%
    arrange(desc(Missing_Count)) %>%
    head(20)

  knitr::kable(missing_summary,
               row.names = FALSE,
               caption = "Variables with Most Missing Data (Top 20)")
} else {
  cat("*Clinical data not available. Run `tar_make(clinical_data)` to retrieve.*\n")
}
```

## Quality Control

```{r qc}
qc <- safe_tar_read("qc_metrics")
filtered <- safe_tar_read("filtered_data")

if (!is.null(qc)) {
  cat("### QC Metrics Summary\n\n")
  qc_df <- data.frame(
    Metric = names(qc),
    Value = unlist(lapply(qc, function(x) {
      if (is.numeric(x) && length(x) == 1) format_with_commas(round(x, 2))
      else as.character(x)
    })),
    stringsAsFactors = FALSE
  )
  knitr::kable(qc_df, caption = "Quality Control Metrics")
}

if (!is.null(filtered) && has_SE) {
  filtered_counts <- SummarizedExperiment::assay(filtered, "counts")
  cat("\n### After Filtering\n\n")
  cat("- **Samples retained:**", ncol(filtered_counts), "\n")
  cat("- **Genes retained:**", format_with_commas(nrow(filtered_counts)), "\n")

  if (!is.null(se_data)) {
    orig_counts <- SummarizedExperiment::assay(se_data, "counts")
    cat("- **Genes removed:**",
        format_with_commas(nrow(orig_counts) - nrow(filtered_counts)),
        sprintf("(%.1f%%)", 100 * (1 - nrow(filtered_counts) / nrow(orig_counts))),
        "\n")
  }
}

if (is.null(qc) && is.null(filtered)) {
  cat("*QC data not available. Run `tar_make()` to generate.*\n")
}
```

## Reproducibility

<details>
<summary>Git Commit Info (click to expand)</summary>

```{r git-info, echo=FALSE}
if (requireNamespace("gert", quietly = TRUE)) {
  tryCatch({
    repo_path <- "."
    if (Sys.getenv("TAR_PROJECT") != "") repo_path <- Sys.getenv("TAR_PROJECT")

    info <- gert::git_info(repo = repo_path)
    log1 <- gert::git_log(max = 1, repo = repo_path)
    knitr::kable(data.frame(
      Item = c("Commit Hash", "Author", "Time", "Branch"),
      Value = c(info$commit, log1$author,
                as.character(log1$time), info$shorthand)
    ))
  }, error = function(e) {
    cat("Git info not available.")
  })
} else {
  cat("gert package not installed.")
}
```

</details>

<details>
<summary>Session Info (click to expand)</summary>

```{r session-info}
sessionInfo()
```

</details>
