---
title: "07. Causal Analysis"
format:
  html:
    code-fold: true
    code-summary: "Show code"
vignette: >
  %\VignetteIndexEntry{07. Causal Analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
in_pkgdown <- nzchar(Sys.getenv("IN_PKGDOWN"))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 8
)
if (!in_pkgdown) library(targets)

# Shared vignette utilities (safe_tar_read, show_target, render helpers)
utils_path <- system.file("vignette_utils.R", package = "coMMpass")
if (utils_path == "") {
  utils_path <- if (file.exists("../inst/vignette_utils.R")) "../inst/vignette_utils.R"
  else if (file.exists("inst/vignette_utils.R")) "inst/vignette_utils.R"
  else stop("Cannot find vignette_utils.R")
}
source(utils_path, local = TRUE)
```

## Overview

This vignette documents the **causal assumptions** underlying the CoMMpass
survival and differential expression analyses. We use directed acyclic graphs
(DAGs) to make these assumptions explicit, identify confounders, and check
whether current models adjust for the right variables.

::: {.callout-note}
DAGs encode domain knowledge, not statistical results. They represent our
*assumptions* about how variables relate causally, which then determines
what we need to adjust for in regression models.
:::

## Causal DAG

The CoMMpass DAG encodes relationships between cytogenetic risk, ISS stage,
treatment, gene expression, and overall survival. Latent (unmeasured)
variables — tumor biology and comorbidities — represent important confounders
that cannot be directly adjusted for.

```{r dag-plot, fig.height = 8, fig.width = 10, fig.cap = "CoMMpass causal DAG. Nodes represent measured and latent variables. Edges represent assumed causal effects. Exposure: cytogenetic_risk. Outcome: overall_survival. Latent variables (tumor_biology, comorbidities) represent unmeasured confounders."}
safe_tar_read("vig_causal_dag_plot")
```

### Key Causal Assumptions

1. **Cytogenetic risk → survival**: High-risk cytogenetics (del(17p), t(4;14),
   gain(1q)) directly shorten survival via treatment resistance and aggressive
   disease biology.

2. **ISS stage as mediator**: Cytogenetic risk influences ISS stage (via tumor
   burden), which in turn affects treatment decisions and survival.

3. **Treatment as collider**: Treatment selection depends on both disease
   severity (ISS, cytogenetics) and patient fitness (age, comorbidities).
   Conditioning on treatment can induce collider bias.

4. **Gene expression as downstream**: Expression patterns are downstream of
   both cytogenetic alterations and tumor biology. They mediate some of the
   effect on survival.

## Adjustment Sets

For each causal question, the DAG identifies the **minimal sufficient
adjustment set** — the smallest set of variables that blocks all backdoor
paths between exposure and outcome.

```{r adjustment-table}
safe_tar_read("vig_causal_adjustment_table")
```

## Model Adequacy Check

We compare the covariates used in current Cox proportional hazards models
against the DAG-implied adjustment sets.

```{r model-check-table}
safe_tar_read("vig_causal_model_check_table")
```

### Interpretation

- If **Sufficient = TRUE**, the model's covariates include at least one
  complete adjustment set — the causal effect estimate is valid under
  the DAG assumptions.
- If **Sufficient = FALSE**, there may be unblocked backdoor paths
  (residual confounding). The recommendation column suggests which
  variables to add.

::: {.callout-warning}
**Unmeasured confounders**: The DAG includes latent variables (tumor biology,
comorbidities) that cannot be adjusted for with available data. Causal
estimates remain subject to unmeasured confounding even with perfect
measured-variable adjustment.
:::

## Implications for Analysis

| Analysis | Causal Question | DAG Recommendation |
|----------|----------------|-------------------|
| DE analysis | Cytogenetic risk → expression | Adjust for age (confounder via comorbidities) |
| KM survival | Cytogenetic risk → OS | Stratify by ISS; do NOT condition on treatment |
| Cox regression | Cytogenetic risk → OS | Adjust for age, comorbidities; avoid treatment as covariate |
| Treatment effect | Treatment → OS | Adjust for age, ISS, cytogenetic risk |

## References

- Textor J, van der Zander B, Gilthorpe MS, Liśkiewicz M, Ellison GT (2016).
  Robust causal inference using directed acyclic graphs: the R package dagitty.
  *International Journal of Epidemiology*, 45(6), 1887-1894.
- Hernán MA, Robins JM (2020). *Causal Inference: What If*.
  Chapman & Hall/CRC.
- Pearl J (2009). *Causality*. Cambridge University Press.

```{r session-info}
sessionInfo()
```
