---
title: "02. Data Dictionary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02. Data Dictionary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
library(targets)
library(dplyr)
has_DT <- requireNamespace("DT", quietly = TRUE)
if (has_DT) library(DT)

# Source utility functions
if (file.exists("../R/utils.R")) {
  source("../R/utils.R")
} else if (file.exists("R/utils.R")) {
  source("R/utils.R")
}
if (file.exists("../R/data_dictionary.R")) {
  source("../R/data_dictionary.R")
} else if (file.exists("R/data_dictionary.R")) {
  source("R/data_dictionary.R")
}

# Discover targets store and project root
proj_dir <- Sys.getenv("TAR_PROJECT")
tar_store <- NULL
project_root <- "."

if (proj_dir != "" && dir.exists(file.path(proj_dir, "_targets"))) {
  tar_store <- file.path(proj_dir, "_targets")
  project_root <- proj_dir
} else if (dir.exists("_targets")) {
  tar_store <- "_targets"
  project_root <- "."
} else if (dir.exists("../_targets")) {
  tar_store <- "../_targets"
  project_root <- ".."
}

safe_tar_read <- function(name) {
  if (is.null(tar_store)) return(NULL)
  tryCatch(tar_read_raw(name, store = tar_store), error = function(e) NULL)
}
```

## Overview

> See the [Glossary](glossary.html) for term definitions and [units reference](glossary.html#units-reference) used throughout this project.

- Describes all variables in the MMRF CoMMpass dataset from the [GDC](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS)
- CoMMpass: longitudinal observational study of ~1,000 newly diagnosed multiple myeloma patients
- Variables span clinical demographics, biospecimen metadata, and RNA-seq gene expression

**Data sources:**

- [GDC Portal - MMRF-COMMPASS](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS)
- [GDC Data Dictionary](https://docs.gdc.cancer.gov/Data_Dictionary/viewer/)
- [MMRF CoMMpass Study](https://themmrf.org/research/commpass-study/)

## Complete Data Dictionary

```{r dictionary-table}
dd <- tryCatch(
  get_commpass_data_dictionary(),
  error = function(e) NULL
)

if (!is.null(dd)) {
  if (has_DT) {
    # Convert gdc_link URLs to short clickable links to save column width
    dd_display <- dd
    if ("gdc_link" %in% names(dd_display)) {
      dd_display$gdc_link <- ifelse(
        is.na(dd_display$gdc_link) | dd_display$gdc_link == "",
        "",
        paste0('<a href="', dd_display$gdc_link, '" target="_blank">GDC</a>')
      )
    }
    DT::datatable(
      dd_display,
      filter = "top",
      escape = FALSE,  # Allow HTML in gdc_link column
      options = list(
        pageLength = 20,
        scrollX = TRUE,
        autoWidth = FALSE,
        columnDefs = list(
          # Prevent wrapping on narrow columns
          list(className = "dt-nowrap", targets = "_all")
        )
      ),
      caption = htmltools::tags$caption(
        style = "caption-side: top; text-align: left;",
        paste0(
          "CoMMpass Data Dictionary - All Variables. ",
          "Click column headers to sort; use filters to search. ",
          "GDC links open the official data dictionary page."
        )
      )
    )
  } else {
    knitr::kable(dd, row.names = FALSE,
                 caption = "CoMMpass Data Dictionary - All Variables")
  }
} else {
  cat("*Data dictionary function not available.*\n")
}
```

## Clinical Data

- Patient demographics, disease characteristics, and outcomes
- Downloaded via `TCGAbiolinks::GDCquery_clinic()`
- All time/age variables in **days** (see [units reference](glossary.html#units-reference))

```{r clinical-summary}
# clinical_data target stores a directory path; load the RDS from it
clinical_dir <- safe_tar_read("clinical_data")
clinical <- NULL

if (!is.null(clinical_dir) && is.character(clinical_dir)) {
  rds_path <- file.path(project_root, clinical_dir, "clinical_data.rds")
  if (file.exists(rds_path)) {
    clinical <- tryCatch(readRDS(rds_path), error = function(e) NULL)
  }
}

if (!is.null(clinical) && is.data.frame(clinical)) {
  cat("### Summary\n\n")
  cat("- **Total patients:**", format_with_commas(nrow(clinical)), "\n")
  cat("- **Variables:**", ncol(clinical), "\n")
  cat("- **Data completeness:**",
      sprintf("%.1f%%", 100 * mean(!is.na(clinical))), "\n\n")

  # --- Column structure overview (str-like) ---
  cat("### Column Structure\n\n")

  col_info <- data.frame(
    Column = names(clinical),
    Type = vapply(clinical, function(x) class(x)[1], character(1)),
    Non_NA = vapply(clinical, function(x) sum(!is.na(x)), integer(1)),
    Pct_Complete = vapply(
      clinical,
      function(x) sprintf("%.0f%%", 100 * mean(!is.na(x))),
      character(1)
    ),
    N_Unique = vapply(
      clinical,
      function(x) length(unique(x[!is.na(x)])),
      integer(1)
    ),
    Example = vapply(clinical, function(x) {
      vals <- x[!is.na(x)]
      if (length(vals) == 0) return("(all NA)")
      if (is.numeric(vals)) {
        paste0(
          "min=", signif(min(vals), 3),
          " med=", signif(median(vals), 3),
          " max=", signif(max(vals), 3)
        )
      } else {
        top <- names(sort(table(vals), decreasing = TRUE))[
          1:min(3, length(unique(vals)))
        ]
        paste(top, collapse = ", ")
      }
    }, character(1)),
    stringsAsFactors = FALSE
  )

  # Add units to column names where known
  unit_map <- c(
    age_at_diagnosis = "age_at_diagnosis_days",
    age_at_index = "age_at_index_days",
    days_to_death = "days_to_death_days",
    days_to_last_follow_up = "days_to_last_follow_up_days",
    days_to_last_known_disease_status = "days_to_last_known_disease_status_days"
  )
  for (i in seq_len(nrow(col_info))) {
    if (col_info$Column[i] %in% names(unit_map)) {
      col_info$Column[i] <- unit_map[col_info$Column[i]]
    }
  }

  # Reorder columns by similarity
  id_cols <- grep("^(submitter|project|case|bcr|patient)", col_info$Column)
  demo_cols <- grep("(gender|sex|race|ethnicity|age_)", col_info$Column)
  time_cols <- grep("(days_|year_|date_)", col_info$Column)
  vital_cols <- grep("(vital|alive|dead|status|cause)", col_info$Column)
  disease_cols <- grep("(diagnosis|stage|grade|iss|tumor|site)", col_info$Column)

  ordered_idx <- unique(c(id_cols, demo_cols, time_cols, vital_cols,
                           disease_cols, seq_len(nrow(col_info))))
  col_info <- col_info[ordered_idx, ]
  rownames(col_info) <- NULL

  if (has_DT) {
    DT::datatable(
      col_info,
      filter = "top",
      options = list(pageLength = 20, scrollX = TRUE, autoWidth = FALSE),
      colnames = c("Column (units appended)", "R Type", "Non-NA",
                    "% Complete", "Unique Values", "Example / Range"),
      caption = htmltools::tags$caption(
        style = "caption-side: top; text-align: left;",
        paste0(
          "Structure of all ", ncol(clinical), " clinical variables for ",
          nrow(clinical), " patients. ",
          "Time variables are in DAYS (GDC convention). ",
          "Columns reordered: IDs > demographics > time > vital status > ",
          "disease > other. ",
          "See glossary for units reference."
        )
      )
    )
  } else {
    knitr::kable(
      col_info, row.names = FALSE,
      col.names = c("Column (units)", "R Type", "Non-NA",
                     "% Complete", "Unique", "Example / Range"),
      caption = paste0(
        "Structure of all ", ncol(clinical), " clinical variables. ",
        "Time variables are in DAYS."
      )
    )
  }
} else {
  cat("*Clinical data not available. Run the pipeline to download.*\n")
}
```

## Biospecimen Data

- Tissue samples collected from patients
- Includes sample type, tissue type, and preservation method

```{r biospecimen-summary}
# biospecimen lives in the same directory as clinical_data
biospecimen <- NULL
if (!is.null(clinical_dir) && is.character(clinical_dir)) {
  bio_rds <- file.path(project_root, clinical_dir, "biospecimen_data.rds")
  if (file.exists(bio_rds)) {
    biospecimen <- tryCatch(readRDS(bio_rds), error = function(e) NULL)
  }
}

if (!is.null(biospecimen) && is.data.frame(biospecimen)) {
  cat("- **Total records:**", format_with_commas(nrow(biospecimen)), "\n")
  cat("- **Variables:**", ncol(biospecimen), "\n\n")

  bio_cols <- data.frame(
    Column = names(biospecimen),
    Type = vapply(biospecimen, function(x) class(x)[1], character(1)),
    Non_NA = vapply(biospecimen, function(x) sum(!is.na(x)), integer(1)),
    Pct_Complete = vapply(
      biospecimen,
      function(x) sprintf("%.0f%%", 100 * mean(!is.na(x))),
      character(1)
    ),
    N_Unique = vapply(
      biospecimen,
      function(x) length(unique(x[!is.na(x)])),
      integer(1)
    ),
    stringsAsFactors = FALSE
  )

  if (has_DT) {
    DT::datatable(
      bio_cols,
      filter = "top",
      options = list(pageLength = 15, scrollX = TRUE),
      colnames = c("Column", "R Type", "Non-NA", "% Complete", "Unique Values"),
      caption = htmltools::tags$caption(
        style = "caption-side: top; text-align: left;",
        paste0(
          "Biospecimen data columns for ", nrow(biospecimen), " records. ",
          "Click column headers to sort; use filters to search."
        )
      )
    )
  } else {
    knitr::kable(bio_cols, row.names = FALSE,
                 col.names = c("Column", "R Type", "Non-NA",
                                "% Complete", "Unique"),
                 caption = "Biospecimen Data Columns")
  }
} else {
  cat("*Biospecimen data not available. Run the pipeline to download.*\n")
}
```

## RNA-seq Data

- RNA-seq gene expression generated using the STAR aligner
- Quantified against GENCODE annotations
- Available in multiple quantification types (raw counts, TPM, FPKM)
- See [units reference](glossary.html#units-reference) for quantification units

```{r rnaseq-summary}
se_data <- NULL
rnaseq_path <- safe_tar_read("raw_rnaseq")

if (!is.null(rnaseq_path)) {
  full_path <- if (file.exists(rnaseq_path)) rnaseq_path else file.path(project_root, rnaseq_path)
  if (file.exists(full_path) && requireNamespace("SummarizedExperiment", quietly = TRUE)) {
    se_data <- tryCatch(readRDS(full_path), error = function(e) NULL)
  }
}

if (!is.null(se_data) && requireNamespace("SummarizedExperiment", quietly = TRUE)) {
  cat("### Dimensions\n\n")
  cat("- **Samples:**", ncol(se_data), "\n")
  cat("- **Genes:**", format_with_commas(nrow(se_data)), "\n")
  cat("- **Assays:**", paste(SummarizedExperiment::assayNames(se_data), collapse = ", "), "\n\n")

  # Gene metadata summary
  gene_meta <- as.data.frame(SummarizedExperiment::rowData(se_data))
  if ("gene_type" %in% names(gene_meta)) {
    cat("### Gene Biotypes\n\n")
    type_counts <- sort(table(gene_meta$gene_type), decreasing = TRUE)
    type_df <- data.frame(
      Biotype = names(type_counts),
      Count = as.integer(type_counts),
      Percentage = sprintf("%.1f%%", 100 * as.numeric(type_counts) / nrow(gene_meta)),
      stringsAsFactors = FALSE
    )
    if (has_DT) {
      DT::datatable(
        type_df,
        filter = "top",
        options = list(pageLength = 15, scrollX = TRUE),
        caption = htmltools::tags$caption(
          style = "caption-side: top; text-align: left;",
          paste0(
            "Gene biotypes from GENCODE annotations. ",
            format_with_commas(nrow(gene_meta)),
            " total gene entries. ",
            "protein_coding genes are the primary analysis target."
          )
        )
      )
    } else {
      knitr::kable(head(type_df, 10), row.names = FALSE,
                   caption = "Top 10 Gene Biotypes")
    }
  }

  # Sample metadata columns
  sample_meta <- as.data.frame(SummarizedExperiment::colData(se_data))
  cat("\n### Sample Metadata Columns\n\n")
  sample_cols <- data.frame(
    Column = names(sample_meta),
    Type = vapply(sample_meta, function(x) class(x)[1], character(1)),
    N_Unique = vapply(
      sample_meta,
      function(x) length(unique(x[!is.na(x)])),
      integer(1)
    ),
    stringsAsFactors = FALSE
  )
  if (has_DT) {
    DT::datatable(
      sample_cols,
      filter = "top",
      options = list(pageLength = 15, scrollX = TRUE),
      caption = htmltools::tags$caption(
        style = "caption-side: top; text-align: left;",
        "RNA-seq sample metadata columns from the SummarizedExperiment colData."
      )
    )
  } else {
    knitr::kable(sample_cols, row.names = FALSE,
                 caption = "RNA-seq Sample Metadata Columns")
  }
} else {
  cat("*RNA-seq data not available. Run the pipeline to download.*\n")
}
```

## Column Name Mappings

GDC uses standardized column names. Common aliases used in analysis:

```{r column-mappings}
mappings <- data.frame(
  GDC_Column = c("submitter_id", "age_at_diagnosis", "vital_status",
                  "days_to_death", "days_to_last_follow_up",
                  "gender", "primary_diagnosis"),
  Common_Alias = c("patient_id", "age_days", "status",
                     "os_time (if dead)", "os_time (if alive)",
                     "sex", "diagnosis"),
  Notes = c("Unique patient identifier",
             "In **days**, not years (divide by 365.25)",
             "Alive/Dead",
             "Overall survival time",
             "Censoring time",
             "male/female",
             "ICD-O-3 morphology"),
  stringsAsFactors = FALSE
)

if (has_DT) {
  DT::datatable(
    mappings,
    filter = "top",
    options = list(pageLength = 10, scrollX = TRUE),
    colnames = c("GDC Column", "Common Alias", "Notes"),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left;",
      "GDC column name mappings to common aliases used in analysis code."
    )
  )
} else {
  knitr::kable(
    mappings, row.names = FALSE,
    col.names = c("GDC Column", "Common Alias", "Notes"),
    caption = "GDC column name mappings."
  )
}
```

## Units Reference

For the complete units reference table (age/time in days, expression quantification types), see the [Glossary units reference](glossary.html#units-reference).

Key rule: **All age and time variables are stored in DAYS by GDC.** This is the single most common source of errors in GDC analyses.

## Data Sources & Links

- **GDC Portal**: <https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS>
- **GDC Data Dictionary**: <https://docs.gdc.cancer.gov/Data_Dictionary/viewer/>
- **GDC mRNA Analysis Pipeline**: <https://docs.gdc.cancer.gov/Data/Bioinformatics_Pipelines/Expression_mRNA_Pipeline/>
- **MMRF CoMMpass Study**: <https://themmrf.org/research/commpass-study/>
- **GENCODE**: <https://www.gencodegenes.org/>
- **Ensembl**: <https://www.ensembl.org/>

For exploratory analysis using these variables, see the [EDA vignette](exploratory-analysis.html).

## R Code Examples

### Loading Data

```{r dd-load-data-code, results='asis'}
code <- safe_tar_read("code_dd_load_data")
if (!is.null(code)) {
  cat("<details>\n<summary>Show code</summary>\n\n```r\n")
  cat(paste(code, collapse = "\n"))
  cat("\n```\n</details>\n")
} else {
  cat("*Code example not available. Run `tar_make()` first.*\n")
}
```

### Exploring the Dictionary

```{r dd-explore-dict-code, results='asis'}
code <- safe_tar_read("code_dd_explore_dict")
if (!is.null(code)) {
  cat("<details>\n<summary>Show code</summary>\n\n```r\n")
  cat(paste(code, collapse = "\n"))
  cat("\n```\n</details>\n")
} else {
  cat("*Code example not available. Run `tar_make()` first.*\n")
}
```

## Reproducibility

<details>
<summary>Session Info (click to expand)</summary>

```{r session-info}
sessionInfo()
```

</details>
