---
title: "01. Data Acquisition & Status"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01. Data Acquisition & Status}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
library(targets)
library(dplyr)
library(ggplot2)
library(stringr)
has_plotly <- requireNamespace("plotly", quietly = TRUE)
has_DT <- requireNamespace("DT", quietly = TRUE)
has_SE <- requireNamespace("SummarizedExperiment", quietly = TRUE)
if (has_plotly) library(plotly)
if (has_DT) library(DT)

# Source utility functions
if (file.exists("../R/utils.R")) {
  source("../R/utils.R")
} else if (file.exists("R/utils.R")) {
  source("R/utils.R")
}

# Discover targets store and project root
proj_dir <- Sys.getenv("TAR_PROJECT")
tar_store <- NULL
project_root <- "."

if (proj_dir != "" && dir.exists(file.path(proj_dir, "_targets"))) {
  tar_store <- file.path(proj_dir, "_targets")
  project_root <- proj_dir
} else if (dir.exists("_targets")) {
  tar_store <- "_targets"
  project_root <- "."
} else if (dir.exists("../_targets")) {
  tar_store <- "../_targets"
  project_root <- ".."
}

# Helper to safely read targets (returns NULL if not available)
safe_tar_read <- function(name) {
  if (is.null(tar_store)) return(NULL)
  tryCatch(tar_read_raw(name, store = tar_store), error = function(e) NULL)
}

# Helper: wrap text at 80 columns
wrap80 <- function(text) {
  paste(strwrap(text, width = 80), collapse = "\n")
}
```

## Overview

> See the [Glossary](glossary.html) for term definitions (read count, library size, gene detection, outlier, etc.).

- RNA-seq and clinical data from [GDC](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS) via the [`targets` pipeline](https://github.com/JohnGavin/coMMpass-analysis/tree/main/R/tar_plans)
- Open Access RNA-seq from the GDC S3 bucket (`s3://gdc-mmrf-commpass-phs000748-2-open/`)
- Clinical metadata via `TCGAbiolinks::GDCquery_clinic()`

## Pipeline Configuration

```{r config}
config <- safe_tar_read("config")

if (!is.null(config)) {
  cat(wrap80(paste0(
    "The pipeline downloads RNA-seq and clinical data from the ",
    "GDC portal (https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS) ",
    "for the ", config$project_id, " project."
  )))
  cat("\n\n")
  cat(wrap80(paste0(
    "Sample limit: ", config$sample_limit,
    " patients (GDC has ~900; subsetting speeds local development). ",
    "In CI, this is capped at 10 samples ",
    "(see R/01_data_acquisition.R) to keep build times under ",
    "5 minutes."
  )))
  cat("\n\n")
  cat(wrap80(paste0(
    "Random seed: ", config$seed,
    " --- ensures reproducible patient selection when ",
    "random_sample = TRUE."
  )))
  cat("\n\n")
  cat(wrap80(paste0(
    "Data directory: ", config$data_dir,
    " | Results directory: ", config$results_dir
  )))
  cat("\n")
} else {
  cat("*Pipeline configuration not available. Run `tar_make(config)` to generate.*\n")
}
```

> **Note:** This vignette was built with `sample_limit = `r if (!is.null(config)) config$sample_limit else "?"`` patients. Numbers below reflect this subset, not the full ~900-patient CoMMpass cohort. For pipeline execution details, see [issue #46 (telemetry vignette)](https://github.com/JohnGavin/coMMpass-analysis/issues/46).

## RNA-seq Data {#rnaseq-data}

RNA-seq gene expression data is downloaded from GDC as a [SummarizedExperiment](https://bioconductor.org/packages/SummarizedExperiment/) object containing [STAR-Counts](https://docs.gdc.cancer.gov/Data/Bioinformatics_Pipelines/Expression_mRNA_Pipeline/).

```{r rnaseq}
# Read the raw_rnaseq target (returns file path to RDS)
rnaseq_path <- safe_tar_read("raw_rnaseq")

# Try to load the SummarizedExperiment
se_data <- NULL
if (!is.null(rnaseq_path)) {
  # Resolve path relative to project root
  full_path <- if (file.exists(rnaseq_path)) {
    rnaseq_path
  } else {
    file.path(project_root, rnaseq_path)
  }

  if (file.exists(full_path) && has_SE) {
    se_data <- tryCatch(readRDS(full_path), error = function(e) NULL)
  }
}

if (!is.null(se_data) && has_SE) {
  counts <- SummarizedExperiment::assay(se_data, "counts")

  cat("### RNA-seq Data Summary\n\n")
  cat("- **Samples:**", ncol(counts), "\n")
  cat("- **Genes:**", format_with_commas(nrow(counts)), "\n")
  cat("- **Total counts:**", format_with_commas(sum(counts)),
      "(sum of all read counts across all genes and samples)\n")
  cat("- **Median counts per sample:**",
      format_with_commas(round(median(colSums(counts)))),
      "(median [library size](glossary.html))\n")
  cat("- **Genes detected (>0 in any sample):**",
      format_with_commas(sum(rowSums(counts) > 0)), "\n")
  cat("- **Sparsity (% zero entries):**",
      sprintf("%.1f%%", 100 * mean(counts == 0)), "\n\n")

  # Per-sample summary
  sample_stats <- data.frame(
    Sample = colnames(counts),
    Total_Counts = colSums(counts),
    Genes_Detected = colSums(counts > 0),
    Median_Count = round(apply(counts, 2, median), 1),
    Max_Count = apply(counts, 2, max),
    stringsAsFactors = FALSE
  )

  if (has_DT) {
    DT::datatable(
      sample_stats,
      filter = "top",
      options = list(pageLength = 10, scrollX = TRUE),
      colnames = c("Sample", "Total Counts (library size)",
                    "Genes Detected (count > 0)",
                    "Median Count", "Max Count"),
      caption = htmltools::tags$caption(
        style = "caption-side: top; text-align: left;",
        paste0(
          "Per-sample RNA-seq summary for ", ncol(counts), " samples. ",
          "Total Counts = library size (sum of all read counts for one sample). ",
          "Genes Detected = number of genes with at least 1 mapped read. ",
          "Sample IDs are GDC barcodes. ",
          "Data: GDC STAR-Counts pipeline. ",
          "This table is NOT repeated in the QC section; see ",
          "Quality Control for outlier flags and size factors."
        )
      )
    )
  } else {
    knitr::kable(
      sample_stats, row.names = FALSE,
      col.names = c("Sample", "Total Counts (library size)",
                     "Genes Detected (count > 0)",
                     "Median Count", "Max Count"),
      caption = paste0(
        "Per-sample RNA-seq summary for ", ncol(counts), " samples. ",
        "Total Counts = library size (sum of all read counts for one sample). ",
        "Genes Detected = number of genes with at least 1 mapped read."
      )
    )
  }
} else if (!is.null(rnaseq_path)) {
  cat("*RNA-seq file path found (`", rnaseq_path,
      "`) but data could not be loaded.*\n", sep = "")
} else {
  cat("*RNA-seq data not available. Run `tar_make(raw_rnaseq)` to download.*\n")
}
```

```{r count-distribution, fig.cap = "Distribution of mean gene expression (log10 scale) across all genes. The three peaks correspond to distinct gene populations: (1) unexpressed/noise genes near zero, (2) lowly-expressed genes detected in a subset of samples, and (3) moderately-to-highly expressed housekeeping/structural genes. This trimodal pattern is expected for GENCODE-annotated RNA-seq data, which includes ~60k entries spanning protein-coding genes, lncRNAs, pseudogenes, and other biotypes with very different expression levels. Data: GDC STAR-Counts."}
if (!is.null(se_data) && has_SE) {
  counts <- SummarizedExperiment::assay(se_data, "counts")
  log_counts <- log10(counts + 1)
  mean_expr <- rowMeans(log_counts)

  # Annotate gene biotypes if available
  gene_meta <- as.data.frame(SummarizedExperiment::rowData(se_data))
  has_biotype <- "gene_type" %in% names(gene_meta)

  if (has_plotly) {
    if (has_biotype) {
      # Color by broad biotype category to explain the three peaks
      gene_meta$biotype_group <- dplyr::case_when(
        gene_meta$gene_type == "protein_coding" ~ "protein-coding",
        gene_meta$gene_type %in% c("lncRNA", "processed_pseudogene",
                                    "unprocessed_pseudogene",
                                    "transcribed_unprocessed_pseudogene",
                                    "transcribed_processed_pseudogene") ~
          "lncRNA / pseudogene",
        TRUE ~ "other (miRNA, snoRNA, etc.)"
      )
      plot_df <- data.frame(
        mean_expr = mean_expr,
        biotype = gene_meta$biotype_group,
        stringsAsFactors = FALSE
      )
      p <- plot_ly(alpha = 0.6) %>%
        add_histogram(
          data = dplyr::filter(plot_df, biotype == "protein-coding"),
          x = ~mean_expr, name = "protein-coding",
          marker = list(color = "#0066CC"),
          nbinsx = 50
        ) %>%
        add_histogram(
          data = dplyr::filter(plot_df, biotype == "lncRNA / pseudogene"),
          x = ~mean_expr, name = "lncRNA / pseudogene",
          marker = list(color = "#DC3545"),
          nbinsx = 50
        ) %>%
        add_histogram(
          data = dplyr::filter(plot_df, biotype == "other (miRNA, snoRNA, etc.)"),
          x = ~mean_expr, name = "other (miRNA, snoRNA, etc.)",
          marker = list(color = "#6C757D"),
          nbinsx = 50
        ) %>%
        layout(
          barmode = "overlay",
          title = list(text = paste0(
            "Distribution of Mean Gene Expression by Biotype<br>",
            "<sup>", format_with_commas(nrow(counts)), " genes, ",
            ncol(counts), " samples</sup>"
          )),
          xaxis = list(title = "Mean log10(counts + 1)"),
          yaxis = list(title = "Number of Genes"),
          bargap = 0.05
        )
      p
    } else {
      p <- plot_ly(
        x = mean_expr,
        type = "histogram",
        nbinsx = 50,
        marker = list(color = "steelblue"),
        name = "Gene Expression"
      ) %>%
        layout(
          title = list(text = paste0(
            "Distribution of Mean Gene Expression<br>",
            "<sup>", format_with_commas(nrow(counts)), " genes, ",
            ncol(counts), " samples — trimodal: ",
            "unexpressed | lowly-expressed | housekeeping</sup>"
          )),
          xaxis = list(title = "Mean log10(counts + 1)"),
          yaxis = list(title = "Number of Genes"),
          bargap = 0.05
        )
      p
    }
  } else {
    hist_data <- data.frame(mean_expression = mean_expr)
    ggplot(hist_data, aes(x = mean_expression)) +
      geom_histogram(bins = 50, fill = "steelblue", color = "white") +
      labs(
        title = "Distribution of Mean Gene Expression",
        subtitle = paste0(
          format_with_commas(nrow(counts)), " genes, ",
          ncol(counts), " samples"
        ),
        x = "Mean log10(counts + 1)", y = "Number of Genes",
        caption = paste0(
          "Three peaks: unexpressed/noise genes (~0), ",
          "lowly-expressed genes (subset of samples), ",
          "housekeeping/structural genes. ",
          "Data: GDC STAR-Counts"
        )
      ) +
      theme_minimal()
  }
}
```

## Clinical Data

Clinical metadata from [GDC](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS) provides patient demographics, disease characteristics, and outcomes. See the [data dictionary](data-dictionary.html) for variable definitions and the [glossary](glossary.html) for units.

```{r clinical}
# Try to load clinical data
# First try clinical_data_clean (from analysis pipeline)
clinical <- safe_tar_read("clinical_data_clean")

# Fall back to loading from the clinical_data directory path
if (is.null(clinical)) {
  clinical_dir <- safe_tar_read("clinical_data")

  if (!is.null(clinical_dir) && is.character(clinical_dir)) {
    dir_path <- if (dir.exists(clinical_dir)) {
      clinical_dir
    } else {
      file.path(project_root, clinical_dir)
    }

    if (dir.exists(dir_path)) {
      rds_file <- file.path(dir_path, "clinical_data.rds")
      csv_file <- file.path(dir_path, "clinical_data.csv")

      if (file.exists(rds_file)) {
        clinical <- tryCatch(readRDS(rds_file), error = function(e) NULL)
      } else if (file.exists(csv_file)) {
        clinical <- tryCatch(read.csv(csv_file), error = function(e) NULL)
      }
    }
  }
}

if (!is.null(clinical) && is.data.frame(clinical)) {
  # Ensure unique column names
  names(clinical) <- make.unique(names(clinical))
  n_patients <- nrow(clinical)

  # Summary statistics
  cat("### Clinical Data Overview\n\n")
  cat("- **Total patients:**", format_with_commas(n_patients), "\n")
  cat("- **Number of variables:**", ncol(clinical), "\n")
  cat("- **Data completeness:**",
      sprintf("%.1f%%", 100 * mean(!is.na(clinical))), "\n\n")

  # --- Column structure overview (str-like) ---
  cat("### Column Structure\n\n")
  cat("All ", ncol(clinical), " columns with type, completeness, and ",
      "example values. Columns grouped by similarity.\n\n", sep = "")

  # Build str-like summary
  col_info <- data.frame(
    Column = names(clinical),
    Type = vapply(clinical, function(x) class(x)[1], character(1)),
    Non_NA = vapply(clinical, function(x) sum(!is.na(x)), integer(1)),
    Pct_Complete = vapply(
      clinical,
      function(x) sprintf("%.0f%%", 100 * mean(!is.na(x))),
      character(1)
    ),
    N_Unique = vapply(
      clinical,
      function(x) length(unique(x[!is.na(x)])),
      integer(1)
    ),
    Example = vapply(clinical, function(x) {
      vals <- x[!is.na(x)]
      if (length(vals) == 0) return("(all NA)")
      if (is.numeric(vals)) {
        paste0(
          "min=", signif(min(vals), 3),
          " med=", signif(median(vals), 3),
          " max=", signif(max(vals), 3)
        )
      } else {
        top <- names(sort(table(vals), decreasing = TRUE))[1:min(3, length(unique(vals)))]
        paste(top, collapse = ", ")
      }
    }, character(1)),
    stringsAsFactors = FALSE
  )

  # Add units to column names where known
  unit_map <- c(
    age_at_diagnosis = "age_at_diagnosis_days",
    age_at_index = "age_at_index_days",
    days_to_death = "days_to_death_days",
    days_to_last_follow_up = "days_to_last_follow_up_days",
    days_to_last_known_disease_status = "days_to_last_known_disease_status_days"
  )
  for (i in seq_len(nrow(col_info))) {
    if (col_info$Column[i] %in% names(unit_map)) {
      col_info$Column[i] <- unit_map[col_info$Column[i]]
    }
  }

  # Reorder columns by similarity: IDs first, then demographics,
  # then time variables, then disease, then treatment, then rest
  id_cols <- grep("^(submitter|project|case|bcr|patient)", col_info$Column)
  demo_cols <- grep("(gender|sex|race|ethnicity|age_)", col_info$Column)
  time_cols <- grep("(days_|year_|date_)", col_info$Column)
  vital_cols <- grep("(vital|alive|dead|status|cause)", col_info$Column)
  disease_cols <- grep("(diagnosis|stage|grade|iss|tumor|site)", col_info$Column)

  ordered_idx <- unique(c(id_cols, demo_cols, time_cols, vital_cols,
                           disease_cols, seq_len(nrow(col_info))))
  col_info <- col_info[ordered_idx, ]
  rownames(col_info) <- NULL

  if (has_DT) {
    DT::datatable(
      col_info,
      filter = "top",
      options = list(
        pageLength = 20,
        scrollX = TRUE,
        autoWidth = FALSE
      ),
      colnames = c("Column (units appended)", "R Type", "Non-NA",
                    "% Complete", "Unique Values", "Example / Range"),
      caption = htmltools::tags$caption(
        style = "caption-side: top; text-align: left;",
        paste0(
          "Structure of all ", ncol(clinical), " clinical variables for ",
          n_patients, " patients. ",
          "Time variables are in DAYS (GDC convention). ",
          "Columns reordered: IDs > demographics > time > vital status > ",
          "disease > other. ",
          "See data dictionary for full definitions and ",
          "glossary for units reference."
        )
      )
    )
  } else {
    knitr::kable(
      col_info, row.names = FALSE,
      col.names = c("Column (units)", "R Type", "Non-NA",
                     "% Complete", "Unique", "Example / Range"),
      caption = paste0(
        "Structure of all ", ncol(clinical), " clinical variables. ",
        "Time variables are in DAYS."
      )
    )
  }
} else {
  cat("*Clinical data not available. Run `tar_make(clinical_data)` to retrieve.*\n")
}
```

### Data Completeness

```{r missing-data}
missing <- safe_tar_read("eda_missing_summary")

if (!is.null(missing)) {
  if (isTRUE(missing$all_complete)) {
    cat(missing$caption, "\n")
  } else {
    cat(paste0(
      missing$n_complete, " of ", missing$n_vars,
      " variables are fully complete. ",
      "Only variables **with** missing data are shown below. ",
      "See [data dictionary](data-dictionary.html) for variable definitions.\n\n"
    ))
    if (has_DT) {
      DT::datatable(
        missing$missing_table,
        filter = "top",
        options = list(pageLength = 15, scrollX = TRUE),
        caption = htmltools::tags$caption(
          style = "caption-side: top; text-align: left;",
          paste0(
            "Variables with missing data (", nrow(missing$missing_table),
            " of ", missing$n_vars, " total). ",
            missing$n_complete, " variables are fully complete and omitted."
          )
        )
      )
    } else {
      knitr::kable(
        missing$missing_table, row.names = FALSE,
        caption = paste0(
          "Variables with missing data (", nrow(missing$missing_table),
          " of ", missing$n_vars, " total). ",
          missing$n_complete, " variables are fully complete and omitted."
        )
      )
    }
  }
} else if (!is.null(clinical) && is.data.frame(clinical)) {
  # Fallback: compute inline if target not available
  missing_counts <- colSums(is.na(clinical))
  has_missing <- missing_counts > 0
  if (sum(has_missing) == 0) {
    cat("All", ncol(clinical), "variables are fully complete (0% missing).\n")
  } else {
    missing_df <- data.frame(
      Variable = names(missing_counts[has_missing]),
      Missing_Count = as.integer(missing_counts[has_missing]),
      Missing_Percent = sprintf("%.1f%%",
        100 * missing_counts[has_missing] / nrow(clinical)),
      stringsAsFactors = FALSE
    )
    missing_df <- missing_df[order(-missing_df$Missing_Count), ]
    n_complete <- sum(!has_missing)
    if (has_DT) {
      DT::datatable(
        missing_df,
        filter = "top",
        options = list(pageLength = 15, scrollX = TRUE),
        caption = htmltools::tags$caption(
          style = "caption-side: top; text-align: left;",
          paste0(
            "Variables with missing data. ",
            n_complete, " of ", ncol(clinical),
            " variables are fully complete and omitted."
          )
        )
      )
    } else {
      knitr::kable(
        missing_df, row.names = FALSE,
        caption = paste0(
          "Variables with missing data. ",
          n_complete, " of ", ncol(clinical),
          " variables are fully complete and omitted."
        )
      )
    }
  }
} else {
  cat("*Clinical data not available. Run `tar_make(clinical_data)` to retrieve.*\n")
}
```

## Quality Control {#quality-control}

Quality control identifies outlier samples based on [library size](glossary.html) and [gene detection](glossary.html) rate. See [`R/02_quality_control.R`](https://github.com/JohnGavin/coMMpass-analysis/blob/main/R/02_quality_control.R) for filtering criteria.

The per-sample expression summary (Total Counts, Genes Detected) is shown in the [RNA-seq Data](#rnaseq-data) section above. This section adds QC-specific metrics: size factors, count dispersion (MAD), and outlier flags.

**Variable definitions** (see also [Glossary](glossary.html)):

- **Total Counts** --- [library size](glossary.html): sum of all [read counts](glossary.html) mapped to genes in one sample
- **Detected Genes** --- number of genes with at least 1 mapped [read](glossary.html) (count > 0)
- **Median Count** --- median read count across all genes in a sample (most genes have 0 or low counts, so this is typically 0)
- **MAD Count** --- [median absolute deviation](https://en.wikipedia.org/wiki/Median_absolute_deviation) of gene counts within a single sample (measures spread of the count distribution for that sample)
- **Size Factor** --- library size / median library size across all samples (values near 1.0 = typical; <<1 = under-sequenced; >>1 = over-sequenced)
- **Outlier** --- flagged `Yes` if the sample falls in the **bottom 5th percentile** of either library size OR genes detected across the `r if (!is.null(config)) config$sample_limit else "N"` samples^[The outlier flag is binary (Yes/No) because it answers: "Is this sample in the bottom 5% tail?" The continuous metrics (Total Counts, Genes Detected) inform the threshold. MAD and Median are computed per-sample across all `r if (!is.null(se_data) && has_SE) format_with_commas(nrow(SummarizedExperiment::assay(se_data, "counts"))) else "~60k"` genes in that sample. The 5th-percentile threshold is computed across all `r if (!is.null(config)) config$sample_limit else "N"` samples.]

```{r qc}
qc_summary <- safe_tar_read("eda_qc_summary")

if (!is.null(qc_summary)) {
  cat("### QC Metrics Summary\n\n")
  cat(paste0(
    "- **Samples assessed:** ", qc_summary$n_samples, "\n",
    "- **Outliers flagged:** ", qc_summary$n_outliers, "\n\n"
  ))
  if (has_DT) {
    DT::datatable(
      qc_summary$table,
      filter = "top",
      options = list(pageLength = 15, scrollX = TRUE),
      caption = htmltools::tags$caption(
        style = "caption-side: top; text-align: left;",
        qc_summary$caption
      )
    )
  } else {
    knitr::kable(
      qc_summary$table, row.names = FALSE,
      caption = qc_summary$caption
    )
  }
} else {
  # Fallback: try raw qc_metrics target
  qc <- safe_tar_read("qc_metrics")
  if (!is.null(qc) && is.data.frame(qc)) {
    cat("### QC Metrics Summary\n\n")
    display_qc <- data.frame(
      Sample = qc$sample,
      `Total Counts` = format(qc$total_counts, big.mark = ","),
      `Detected Genes` = format(qc$detected_genes, big.mark = ","),
      `Median Count` = sprintf("%.1f", qc$median_count),
      `MAD Count` = sprintf("%.1f", qc$mad_count),
      `Size Factor` = sprintf("%.3f", qc$size_factor),
      Outlier = ifelse(qc$is_outlier, "Yes", "No"),
      check.names = FALSE,
      stringsAsFactors = FALSE
    )
    if (has_DT) {
      DT::datatable(
        display_qc,
        filter = "top",
        options = list(pageLength = 15, scrollX = TRUE),
        caption = htmltools::tags$caption(
          style = "caption-side: top; text-align: left;",
          paste0(
            "Per-sample QC metrics for ", nrow(qc), " samples. ",
            sum(qc$is_outlier), " outlier(s) flagged (bottom 5% by ",
            "library size or genes detected). ",
            "MAD = median absolute deviation of gene counts within ",
            "each sample."
          )
        )
      )
    } else {
      knitr::kable(
        display_qc, row.names = FALSE,
        caption = paste0(
          "Per-sample QC metrics for ", nrow(qc), " samples. ",
          sum(qc$is_outlier), " outlier(s) flagged."
        )
      )
    }
  } else {
    cat("*QC data not available. Run `tar_make()` to generate.*\n")
  }
}
```

```{r filtered}
filtered <- safe_tar_read("filtered_data")

if (!is.null(filtered) && has_SE) {
  filtered_counts <- SummarizedExperiment::assay(filtered, "counts")
  cat("\n### After Filtering\n\n")
  cat("- **Samples retained:**", ncol(filtered_counts), "\n")
  cat("- **Genes retained:**", format_with_commas(nrow(filtered_counts)), "\n")

  if (!is.null(se_data)) {
    orig_counts <- SummarizedExperiment::assay(se_data, "counts")
    cat("- **Genes removed:**",
        format_with_commas(nrow(orig_counts) - nrow(filtered_counts)),
        sprintf("(%.1f%%)", 100 * (1 - nrow(filtered_counts) / nrow(orig_counts))),
        "\n")
  }
}
```

## Reproducibility

<details>
<summary>Git Commit Info (click to expand)</summary>

```{r git-info, echo=FALSE}
if (requireNamespace("gert", quietly = TRUE)) {
  tryCatch({
    repo_path <- "."
    if (Sys.getenv("TAR_PROJECT") != "") repo_path <- Sys.getenv("TAR_PROJECT")

    info <- gert::git_info(repo = repo_path)
    log1 <- gert::git_log(max = 1, repo = repo_path)
    knitr::kable(data.frame(
      Item = c("Commit Hash", "Author", "Time", "Branch"),
      Value = c(info$commit, log1$author,
                as.character(log1$time), info$shorthand)
    ), row.names = FALSE)
  }, error = function(e) {
    cat("Git info not available.")
  })
} else {
  cat("gert package not installed.")
}
```

</details>

<details>
<summary>Session Info (click to expand)</summary>

```{r session-info}
sessionInfo()
```

</details>
