---
title: "01. Data Acquisition & Status"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01. Data Acquisition & Status}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
library(targets)
library(dplyr)
library(ggplot2)
library(stringr)
has_plotly <- requireNamespace("plotly", quietly = TRUE)
has_DT <- requireNamespace("DT", quietly = TRUE)
has_SE <- requireNamespace("SummarizedExperiment", quietly = TRUE)
if (has_plotly) library(plotly)
if (has_DT) library(DT)

# Source utility functions
if (file.exists("../R/utils.R")) {
  source("../R/utils.R")
} else if (file.exists("R/utils.R")) {
  source("R/utils.R")
}

# Discover targets store and project root
proj_dir <- Sys.getenv("TAR_PROJECT")
tar_store <- NULL
project_root <- "."

if (proj_dir != "" && dir.exists(file.path(proj_dir, "_targets"))) {
  tar_store <- file.path(proj_dir, "_targets")
  project_root <- proj_dir
} else if (dir.exists("_targets")) {
  tar_store <- "_targets"
  project_root <- "."
} else if (dir.exists("../_targets")) {
  tar_store <- "../_targets"
  project_root <- ".."
}

# Helper to safely read targets (returns NULL if not available)
safe_tar_read <- function(name) {
  if (is.null(tar_store)) return(NULL)
  tryCatch(tar_read_raw(name, store = tar_store), error = function(e) NULL)
}
```

## Overview

This report summarizes the initial data acquisition step for the MMRF CoMMpass study. We access Open Access RNA-seq data from the GDC S3 bucket (`s3://gdc-mmrf-commpass-phs000748-2-open/`) via the [`targets` pipeline](https://github.com/JohnGavin/coMMpass-analysis/tree/main/R/tar_plans).

## Pipeline Configuration

```{r config}
config <- safe_tar_read("config")

if (!is.null(config)) {
  is_ci <- Sys.getenv("CI") == "true" || Sys.getenv("GITHUB_ACTIONS") == "true"
  cat(paste0(
    "The pipeline downloads RNA-seq and clinical data from the ",
    "[GDC portal](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS) ",
    "for the **", config$project_id, "** project.\n\n"
  ))
  cat(paste0(
    "- **Sample limit:** ", config$sample_limit,
    " patients (GDC has ~900; subsetting speeds local development). ",
    "In CI, this is [capped at 10 samples]",
    "(https://github.com/JohnGavin/coMMpass-analysis/blob/main/",
    "R/01_data_acquisition.R#L44-L47) to keep build times under 5 minutes.\n"
  ))
  cat(paste0(
    "- **Random seed:** ", config$seed,
    " --- ensures reproducible patient selection when `random_sample = TRUE`.\n"
  ))
  cat(paste0(
    "- **Data directory:** `", config$data_dir,
    "` | **Results directory:** `", config$results_dir, "`\n"
  ))
} else {
  cat("*Pipeline configuration not available. Run `tar_make(config)` to generate.*\n")
}
```

> **Note:** This vignette was built with `sample_limit = `r if (!is.null(config)) config$sample_limit else "?"`` patients. Numbers below reflect this subset, not the full ~900-patient CoMMpass cohort. For pipeline execution details, see [issue #46 (telemetry vignette)](https://github.com/JohnGavin/coMMpass-analysis/issues/46).

## RNA-seq Data

RNA-seq gene expression data is downloaded from GDC as a [SummarizedExperiment](https://bioconductor.org/packages/SummarizedExperiment/) object containing [STAR-Counts](https://docs.gdc.cancer.gov/Data/Bioinformatics_Pipelines/Expression_mRNA_Pipeline/).

```{r rnaseq}
# Read the raw_rnaseq target (returns file path to RDS)
rnaseq_path <- safe_tar_read("raw_rnaseq")

# Try to load the SummarizedExperiment
se_data <- NULL
if (!is.null(rnaseq_path)) {
  # Resolve path relative to project root
  full_path <- if (file.exists(rnaseq_path)) {
    rnaseq_path
  } else {
    file.path(project_root, rnaseq_path)
  }

  if (file.exists(full_path) && has_SE) {
    se_data <- tryCatch(readRDS(full_path), error = function(e) NULL)
  }
}

if (!is.null(se_data) && has_SE) {
  counts <- SummarizedExperiment::assay(se_data, "counts")

  cat("### RNA-seq Data Summary\n\n")
  cat("- **Samples:**", ncol(counts), "\n")
  cat("- **Genes:**", format_with_commas(nrow(counts)), "\n")
  cat("- **Total counts:**", format_with_commas(sum(counts)), "\n")
  cat("- **Median counts per sample:**",
      format_with_commas(round(median(colSums(counts)))), "\n")
  cat("- **Genes detected (>0 in any sample):**",
      format_with_commas(sum(rowSums(counts) > 0)), "\n")
  cat("- **Sparsity (% zero entries):**",
      sprintf("%.1f%%", 100 * mean(counts == 0)), "\n\n")

  # Per-sample summary
  sample_stats <- data.frame(
    Sample = colnames(counts),
    Total_Counts = format_with_commas(colSums(counts)),
    Genes_Detected = format_with_commas(colSums(counts > 0)),
    Median_Count = sprintf("%.1f", apply(counts, 2, median)),
    Max_Count = format_with_commas(apply(counts, 2, max)),
    stringsAsFactors = FALSE
  )

  knitr::kable(
    sample_stats,
    col.names = c("Sample", "Total Counts", "Genes Detected",
                   "Median Count", "Max Count"),
    row.names = FALSE,
    caption = paste0(
      "Per-sample expression summary for ", ncol(counts), " samples. ",
      "Total Counts = library size, Genes Detected = genes with >0 counts. ",
      "Sample IDs are GDC barcodes. ",
      "Data: [GDC STAR-Counts](https://docs.gdc.cancer.gov/Data/",
      "Bioinformatics_Pipelines/Expression_mRNA_Pipeline/)."
    )
  )
} else if (!is.null(rnaseq_path)) {
  cat("*RNA-seq file path found (`", rnaseq_path,
      "`) but data could not be loaded.*\n", sep = "")
} else {
  cat("*RNA-seq data not available. Run `tar_make(raw_rnaseq)` to download.*\n")
}
```

```{r count-distribution, fig.cap = "Distribution of mean gene expression (log10 scale) across all genes. Bimodal shape reflects expressed vs. unexpressed genes. Data: GDC STAR-Counts."}
if (!is.null(se_data) && has_SE) {
  counts <- SummarizedExperiment::assay(se_data, "counts")
  log_counts <- log10(counts + 1)
  mean_expr <- rowMeans(log_counts)

  if (has_plotly) {
    p <- plot_ly(
      x = mean_expr,
      type = "histogram",
      nbinsx = 50,
      marker = list(color = "steelblue"),
      name = "Gene Expression"
    ) %>%
      layout(
        title = list(text = paste0(
          "Distribution of Mean Gene Expression<br>",
          "<sup>", format_with_commas(nrow(counts)), " genes, ",
          ncol(counts), " samples</sup>"
        )),
        xaxis = list(title = "log10(counts + 1)"),
        yaxis = list(title = "Number of Genes"),
        bargap = 0.05
      )
    p
  } else {
    hist_data <- data.frame(mean_expression = mean_expr)
    ggplot(hist_data, aes(x = mean_expression)) +
      geom_histogram(bins = 50, fill = "steelblue", color = "white") +
      labs(
        title = "Distribution of Mean Gene Expression",
        subtitle = paste0(
          format_with_commas(nrow(counts)), " genes, ",
          ncol(counts), " samples"
        ),
        x = "log10(counts + 1)", y = "Number of Genes",
        caption = "Data: GDC STAR-Counts"
      ) +
      theme_minimal()
  }
}
```

## Clinical Data

Clinical metadata retrieved from [GDC](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS) provides patient demographics, disease characteristics, and outcomes. See the [data dictionary](data-dictionary.html) for variable definitions.

```{r clinical}
# Try to load clinical data
# First try clinical_data_clean (from analysis pipeline)
clinical <- safe_tar_read("clinical_data_clean")

# Fall back to loading from the clinical_data directory path
if (is.null(clinical)) {
  clinical_dir <- safe_tar_read("clinical_data")

  if (!is.null(clinical_dir) && is.character(clinical_dir)) {
    dir_path <- if (dir.exists(clinical_dir)) {
      clinical_dir
    } else {
      file.path(project_root, clinical_dir)
    }

    if (dir.exists(dir_path)) {
      rds_file <- file.path(dir_path, "clinical_data.rds")
      csv_file <- file.path(dir_path, "clinical_data.csv")

      if (file.exists(rds_file)) {
        clinical <- tryCatch(readRDS(rds_file), error = function(e) NULL)
      } else if (file.exists(csv_file)) {
        clinical <- tryCatch(read.csv(csv_file), error = function(e) NULL)
      }
    }
  }
}

if (!is.null(clinical) && is.data.frame(clinical)) {
  # Ensure unique column names
  names(clinical) <- make.unique(names(clinical))
  n_patients <- nrow(clinical)

  # Summary statistics
  cat("### Clinical Data Overview\n\n")
  cat("- **Total patients:**", format_with_commas(n_patients), "\n")
  cat("- **Number of variables:**", ncol(clinical), "\n")
  cat("- **Data completeness:**", sprintf("%.1f%%", 100 * mean(!is.na(clinical))), "\n\n")

  # Key demographic variables
  cols_to_show <- c("project_id", "submitter_id", "gender", "vital_status",
                    "age_at_diagnosis", "race", "ethnicity")
  cols_available <- intersect(cols_to_show, names(clinical))

  if (length(cols_available) > 0) {
    cat("### Patient Demographics\n\n")

    # Gender distribution
    if ("gender" %in% names(clinical)) {
      gender_table <- table(clinical$gender, useNA = "ifany")
      gender_df <- data.frame(
        Gender = names(gender_table),
        Count = as.numeric(gender_table),
        Percentage = sprintf("%.1f%%", 100 * as.numeric(gender_table) / n_patients)
      )
      knitr::kable(
        gender_df, row.names = FALSE,
        caption = paste0(
          "Gender distribution for ", n_patients, " patients. ",
          "Data: [GDC clinical](https://portal.gdc.cancer.gov/",
          "projects/MMRF-COMMPASS). ",
          "See [data dictionary](data-dictionary.html) for definitions."
        )
      )
      cat("\n")
    }

    # Vital status
    if ("vital_status" %in% names(clinical)) {
      vital_table <- table(clinical$vital_status, useNA = "ifany")
      vital_df <- data.frame(
        Status = names(vital_table),
        Count = as.numeric(vital_table),
        Percentage = sprintf("%.1f%%", 100 * as.numeric(vital_table) / n_patients)
      )
      knitr::kable(
        vital_df, row.names = FALSE,
        caption = paste0(
          "Vital status for ", n_patients, " patients. ",
          "Alive = censored at last follow-up. ",
          "See [data dictionary](data-dictionary.html)."
        )
      )
      cat("\n")
    }

    # Age at diagnosis
    if ("age_at_diagnosis" %in% names(clinical) && is.numeric(clinical$age_at_diagnosis)) {
      age_values <- clinical$age_at_diagnosis[!is.na(clinical$age_at_diagnosis)]
      is_days <- max(age_values, na.rm = TRUE) > 120
      age_years <- if (is_days) age_values / 365.25 else age_values

      cat("### Age at Diagnosis (years)\n\n")
      cat("- **Median:**", sprintf("%.1f", median(age_years, na.rm = TRUE)), "\n")
      cat("- **Range:**", sprintf("%.1f - %.1f",
                                   min(age_years, na.rm = TRUE),
                                   max(age_years, na.rm = TRUE)), "\n\n")
    }

    # Sample of clinical data
    cat("\n### Sample Clinical Data\n\n")
    display_clinical <- clinical[seq_len(min(10, n_patients)), cols_available, drop = FALSE]

    if (has_DT) {
      DT::datatable(
        display_clinical,
        options = list(pageLength = 5, scrollX = TRUE),
        caption = paste0(
          "First ", min(10, n_patients), " of ", n_patients,
          " patients showing key demographics. ",
          "age_at_diagnosis is in days (divide by 365.25 for years). ",
          "See [data dictionary](data-dictionary.html) for all variables."
        )
      )
    } else {
      knitr::kable(
        display_clinical, row.names = FALSE,
        caption = paste0(
          "First ", min(10, n_patients), " of ", n_patients,
          " patients showing key demographics."
        )
      )
    }
  }
} else {
  cat("*Clinical data not available. Run `tar_make(clinical_data)` to retrieve.*\n")
}
```

### Data Completeness

```{r missing-data}
missing <- safe_tar_read("eda_missing_summary")

if (!is.null(missing)) {
  if (isTRUE(missing$all_complete)) {
    cat(missing$caption, "\n")
  } else {
    cat(paste0(
      missing$n_complete, " of ", missing$n_vars,
      " variables are fully complete. ",
      "Only variables **with** missing data are shown below. ",
      "See [data dictionary](data-dictionary.html) for variable definitions.\n\n"
    ))
    knitr::kable(
      missing$missing_table, row.names = FALSE,
      caption = paste0(
        "Variables with missing data (", nrow(missing$missing_table),
        " of ", missing$n_vars, " total). ",
        missing$n_complete, " variables are fully complete and omitted. ",
        "See [data dictionary](data-dictionary.html) for definitions."
      )
    )
  }
} else if (!is.null(clinical) && is.data.frame(clinical)) {
  # Fallback: compute inline if target not available
  missing_counts <- colSums(is.na(clinical))
  has_missing <- missing_counts > 0
  if (sum(has_missing) == 0) {
    cat("All", ncol(clinical), "variables are fully complete (0% missing).\n")
  } else {
    missing_df <- data.frame(
      Variable = names(missing_counts[has_missing]),
      Missing_Count = as.integer(missing_counts[has_missing]),
      Missing_Percent = sprintf("%.1f%%", 100 * missing_counts[has_missing] / nrow(clinical)),
      stringsAsFactors = FALSE
    )
    missing_df <- missing_df[order(-missing_df$Missing_Count), ]
    n_complete <- sum(!has_missing)
    knitr::kable(
      missing_df, row.names = FALSE,
      caption = paste0(
        "Variables with missing data. ",
        n_complete, " of ", ncol(clinical),
        " variables are fully complete and omitted."
      )
    )
  }
} else {
  cat("*Clinical data not available. Run `tar_make(clinical_data)` to retrieve.*\n")
}
```

## Quality Control

Quality control identifies outlier samples based on library size and gene detection rate. Samples in the bottom 5% on either metric are flagged. See [`R/02_quality_control.R`](https://github.com/JohnGavin/coMMpass-analysis/blob/main/R/02_quality_control.R) for filtering criteria.

**Variable definitions:**

- **Total Counts** --- library size (sum of all read counts per sample)
- **Detected Genes** --- number of genes with at least 1 read
- **Median Count** --- median read count across all genes in a sample
- **MAD Count** --- median absolute deviation of counts (spread measure)
- **Size Factor** --- library size / median library size across all samples
- **Outlier** --- flagged if in bottom 5% by library size OR genes detected

```{r qc}
qc_summary <- safe_tar_read("eda_qc_summary")

if (!is.null(qc_summary)) {
  cat("### QC Metrics Summary\n\n")
  cat(paste0(
    "- **Samples assessed:** ", qc_summary$n_samples, "\n",
    "- **Outliers flagged:** ", qc_summary$n_outliers, "\n\n"
  ))
  knitr::kable(
    qc_summary$table, row.names = FALSE,
    caption = qc_summary$caption
  )
} else {
  # Fallback: try raw qc_metrics target
  qc <- safe_tar_read("qc_metrics")
  if (!is.null(qc) && is.data.frame(qc)) {
    cat("### QC Metrics Summary\n\n")
    display_qc <- data.frame(
      Sample = qc$sample,
      `Total Counts` = format(qc$total_counts, big.mark = ","),
      `Detected Genes` = format(qc$detected_genes, big.mark = ","),
      `Median Count` = sprintf("%.1f", qc$median_count),
      `MAD Count` = sprintf("%.1f", qc$mad_count),
      `Size Factor` = sprintf("%.3f", qc$size_factor),
      Outlier = ifelse(qc$is_outlier, "Yes", "No"),
      check.names = FALSE,
      stringsAsFactors = FALSE
    )
    knitr::kable(
      display_qc, row.names = FALSE,
      caption = paste0(
        "Per-sample QC metrics for ", nrow(qc), " samples. ",
        sum(qc$is_outlier), " outlier(s) flagged."
      )
    )
  } else {
    cat("*QC data not available. Run `tar_make()` to generate.*\n")
  }
}
```

```{r filtered}
filtered <- safe_tar_read("filtered_data")

if (!is.null(filtered) && has_SE) {
  filtered_counts <- SummarizedExperiment::assay(filtered, "counts")
  cat("\n### After Filtering\n\n")
  cat("- **Samples retained:**", ncol(filtered_counts), "\n")
  cat("- **Genes retained:**", format_with_commas(nrow(filtered_counts)), "\n")

  if (!is.null(se_data)) {
    orig_counts <- SummarizedExperiment::assay(se_data, "counts")
    cat("- **Genes removed:**",
        format_with_commas(nrow(orig_counts) - nrow(filtered_counts)),
        sprintf("(%.1f%%)", 100 * (1 - nrow(filtered_counts) / nrow(orig_counts))),
        "\n")
  }
}
```

## Reproducibility

<details>
<summary>Git Commit Info (click to expand)</summary>

```{r git-info, echo=FALSE}
if (requireNamespace("gert", quietly = TRUE)) {
  tryCatch({
    repo_path <- "."
    if (Sys.getenv("TAR_PROJECT") != "") repo_path <- Sys.getenv("TAR_PROJECT")

    info <- gert::git_info(repo = repo_path)
    log1 <- gert::git_log(max = 1, repo = repo_path)
    knitr::kable(data.frame(
      Item = c("Commit Hash", "Author", "Time", "Branch"),
      Value = c(info$commit, log1$author,
                as.character(log1$time), info$shorthand)
    ), row.names = FALSE)
  }, error = function(e) {
    cat("Git info not available.")
  })
} else {
  cat("gert package not installed.")
}
```

</details>

<details>
<summary>Session Info (click to expand)</summary>

```{r session-info}
sessionInfo()
```

</details>
