---
title: "02. Data Dictionary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02. Data Dictionary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
library(targets)
library(dplyr)
has_DT <- requireNamespace("DT", quietly = TRUE)
if (has_DT) library(DT)

# Source utility functions
if (file.exists("../R/utils.R")) {
  source("../R/utils.R")
} else if (file.exists("R/utils.R")) {
  source("R/utils.R")
}
if (file.exists("../R/data_dictionary.R")) {
  source("../R/data_dictionary.R")
} else if (file.exists("R/data_dictionary.R")) {
  source("R/data_dictionary.R")
}

# Discover targets store and project root
proj_dir <- Sys.getenv("TAR_PROJECT")
tar_store <- NULL
project_root <- "."

if (proj_dir != "" && dir.exists(file.path(proj_dir, "_targets"))) {
  tar_store <- file.path(proj_dir, "_targets")
  project_root <- proj_dir
} else if (dir.exists("_targets")) {
  tar_store <- "_targets"
  project_root <- "."
} else if (dir.exists("../_targets")) {
  tar_store <- "../_targets"
  project_root <- ".."
}

safe_tar_read <- function(name) {
  if (is.null(tar_store)) return(NULL)
  tryCatch(tar_read_raw(name, store = tar_store), error = function(e) NULL)
}
```

## Overview

This document describes all variables in the MMRF CoMMpass dataset as downloaded from the [Genomic Data Commons (GDC)](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS). The CoMMpass (Relating Clinical Outcomes in Multiple Myeloma to Personal Assessment of Genetic Profile) study is a longitudinal observational study of ~1,000 newly diagnosed multiple myeloma patients.

**Data sources:**

- [GDC Portal - MMRF-COMMPASS](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS)
- [GDC Data Dictionary](https://docs.gdc.cancer.gov/Data_Dictionary/viewer/)
- [MMRF CoMMpass Study](https://themmrf.org/research/commpass-study/)

## Complete Data Dictionary

```{r dictionary-table}
dd <- tryCatch(
  get_commpass_data_dictionary(),
  error = function(e) NULL
)

if (!is.null(dd)) {
  if (has_DT) {
    # Convert gdc_link URLs to short clickable links to save column width
    dd_display <- dd
    if ("gdc_link" %in% names(dd_display)) {
      dd_display$gdc_link <- ifelse(
        is.na(dd_display$gdc_link) | dd_display$gdc_link == "",
        "",
        paste0('<a href="', dd_display$gdc_link, '" target="_blank">GDC</a>')
      )
    }
    DT::datatable(
      dd_display,
      filter = "top",
      escape = FALSE,  # Allow HTML in gdc_link column
      options = list(
        pageLength = 20,
        scrollX = TRUE,
        autoWidth = FALSE,
        columnDefs = list(
          # Prevent wrapping on narrow columns
          list(className = "dt-nowrap", targets = "_all")
        )
      ),
      caption = "CoMMpass Data Dictionary - All Variables"
    )
  } else {
    knitr::kable(dd, row.names = FALSE, caption = "CoMMpass Data Dictionary")
  }
} else {
  cat("*Data dictionary function not available.*\n")
}
```

## Clinical Data

Clinical data includes patient demographics, disease characteristics, and outcomes. Downloaded via `TCGAbiolinks::GDCquery_clinic()`.

```{r clinical-summary}
# clinical_data target stores a directory path; load the RDS from it
clinical_dir <- safe_tar_read("clinical_data")
clinical <- NULL

if (!is.null(clinical_dir) && is.character(clinical_dir)) {
  rds_path <- file.path(project_root, clinical_dir, "clinical_data.rds")
  if (file.exists(rds_path)) {
    clinical <- tryCatch(readRDS(rds_path), error = function(e) NULL)
  }
}

if (!is.null(clinical) && is.data.frame(clinical)) {
  cat("### Summary\n\n")
  cat("- **Total patients:**", format_with_commas(nrow(clinical)), "\n")
  cat("- **Variables:**", ncol(clinical), "\n")
  cat("- **Data completeness:**",
      sprintf("%.1f%%", 100 * mean(!is.na(clinical))), "\n\n")

  # Column type overview
  col_types <- data.frame(
    Column = names(clinical),
    Type = sapply(clinical, function(x) class(x)[1]),
    Non_NA = sapply(clinical, function(x) sum(!is.na(x))),
    Pct_Complete = sprintf("%.0f%%", 100 * sapply(clinical, function(x) mean(!is.na(x)))),
    stringsAsFactors = FALSE
  )

  cat("### Column Inventory\n\n")
  if (has_DT) {
    DT::datatable(
      col_types,
      filter = "top",
      options = list(pageLength = 15, scrollX = TRUE),
      caption = "Clinical Data Columns"
    )
  } else {
    knitr::kable(col_types, caption = "Clinical Data Columns", row.names = FALSE)
  }

  # Sample rows
  cat("\n### Sample Data (first 5 rows, key columns)\n\n")
  key_cols <- intersect(
    c("submitter_id", "age_at_diagnosis", "gender", "race",
      "vital_status", "days_to_death", "days_to_last_follow_up",
      "primary_diagnosis"),
    names(clinical)
  )
  if (length(key_cols) > 0) {
    display <- head(clinical[, key_cols, drop = FALSE], 5)
    knitr::kable(display, row.names = FALSE,
                 caption = "Sample clinical records (first 5 rows, key columns). age_at_diagnosis is in days.")
  }
} else {
  cat("*Clinical data not available. Run the pipeline to download.*\n")
}
```

## Biospecimen Data

Biospecimen data describes the tissue samples collected from patients.

```{r biospecimen-summary}
# biospecimen lives in the same directory as clinical_data
biospecimen <- NULL
if (!is.null(clinical_dir) && is.character(clinical_dir)) {
  bio_rds <- file.path(project_root, clinical_dir, "biospecimen_data.rds")
  if (file.exists(bio_rds)) {
    biospecimen <- tryCatch(readRDS(bio_rds), error = function(e) NULL)
  }
}

if (!is.null(biospecimen) && is.data.frame(biospecimen)) {
  cat("- **Total records:**", format_with_commas(nrow(biospecimen)), "\n")
  cat("- **Variables:**", ncol(biospecimen), "\n\n")

  bio_cols <- data.frame(
    Column = names(biospecimen),
    Type = sapply(biospecimen, function(x) class(x)[1]),
    Non_NA = sapply(biospecimen, function(x) sum(!is.na(x))),
    Pct_Complete = sprintf("%.0f%%", 100 * sapply(biospecimen, function(x) mean(!is.na(x)))),
    stringsAsFactors = FALSE
  )

  if (has_DT) {
    DT::datatable(bio_cols, filter = "top",
                  options = list(pageLength = 10, scrollX = TRUE),
                  caption = "Biospecimen Data Columns")
  } else {
    knitr::kable(bio_cols, caption = "Biospecimen Data Columns", row.names = FALSE)
  }
} else {
  cat("*Biospecimen data not available. Run the pipeline to download.*\n")
}
```

## RNA-seq Data

RNA-seq gene expression data is generated using the STAR aligner and quantified against GENCODE annotations. Data is available in multiple quantification types.

```{r rnaseq-summary}
se_data <- NULL
rnaseq_path <- safe_tar_read("raw_rnaseq")

if (!is.null(rnaseq_path)) {
  full_path <- if (file.exists(rnaseq_path)) rnaseq_path else file.path(project_root, rnaseq_path)
  if (file.exists(full_path) && requireNamespace("SummarizedExperiment", quietly = TRUE)) {
    se_data <- tryCatch(readRDS(full_path), error = function(e) NULL)
  }
}

if (!is.null(se_data) && requireNamespace("SummarizedExperiment", quietly = TRUE)) {
  cat("### Dimensions\n\n")
  cat("- **Samples:**", ncol(se_data), "\n")
  cat("- **Genes:**", format_with_commas(nrow(se_data)), "\n")
  cat("- **Assays:**", paste(SummarizedExperiment::assayNames(se_data), collapse = ", "), "\n\n")

  # Gene metadata summary
  gene_meta <- as.data.frame(SummarizedExperiment::rowData(se_data))
  if ("gene_type" %in% names(gene_meta)) {
    cat("### Gene Biotypes\n\n")
    type_counts <- sort(table(gene_meta$gene_type), decreasing = TRUE)
    type_df <- data.frame(
      Biotype = names(type_counts),
      Count = as.integer(type_counts),
      Percentage = sprintf("%.1f%%", 100 * as.numeric(type_counts) / nrow(gene_meta)),
      stringsAsFactors = FALSE
    )
    knitr::kable(head(type_df, 10), caption = "Top 10 Gene Biotypes", row.names = FALSE)
  }

  # Sample metadata columns
  sample_meta <- as.data.frame(SummarizedExperiment::colData(se_data))
  cat("\n### Sample Metadata Columns\n\n")
  sample_cols <- data.frame(
    Column = names(sample_meta),
    Type = sapply(sample_meta, function(x) class(x)[1]),
    stringsAsFactors = FALSE
  )
  knitr::kable(sample_cols, caption = "RNA-seq Sample Metadata Columns", row.names = FALSE)
} else {
  cat("*RNA-seq data not available. Run the pipeline to download.*\n")
}
```

## Column Name Mappings

GDC uses standardized column names. Common aliases used in analysis:

| GDC Column | Common Alias | Notes |
|---|---|---|
| `submitter_id` | patient_id | Unique patient identifier |
| `age_at_diagnosis` | age_days | In **days**, not years |
| `vital_status` | status | Alive/Dead |
| `days_to_death` | os_time (if dead) | Overall survival time |
| `days_to_last_follow_up` | os_time (if alive) | Censoring time |
| `gender` | sex | male/female |
| `primary_diagnosis` | diagnosis | ICD-O-3 morphology |

## Units Reference

**Age and time variables are stored in DAYS by GDC.** This is the single most common source of errors in GDC analyses.

| Variable | Unit | Conversion |
|---|---|---|
| `age_at_diagnosis` | days | `age_years = age_at_diagnosis / 365.25` |
| `days_to_death` | days | Used directly in survival analysis |
| `days_to_last_follow_up` | days | Censoring time for survival |
| `days_to_last_known_disease_status` | days | Disease assessment time |
| `unstranded` / `stranded_*` | raw integer counts | Input for DESeq2/edgeR |
| `tpm_unstranded` | TPM | Cross-sample comparison |
| `fpkm_unstranded` | FPKM | Largely deprecated, use TPM |

## Data Sources & Links

- **GDC Portal**: <https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS>
- **GDC Data Dictionary**: <https://docs.gdc.cancer.gov/Data_Dictionary/viewer/>
- **GDC mRNA Analysis Pipeline**: <https://docs.gdc.cancer.gov/Data/Bioinformatics_Pipelines/Expression_mRNA_Pipeline/>
- **MMRF CoMMpass Study**: <https://themmrf.org/research/commpass-study/>
- **GENCODE**: <https://www.gencodegenes.org/>
- **Ensembl**: <https://www.ensembl.org/>

## R Code Examples

### Loading Data

```{r dd-load-data-code, results='asis'}
code <- safe_tar_read("code_dd_load_data")
if (!is.null(code)) {
  cat("<details>\n<summary>Show code</summary>\n\n```r\n")
  cat(paste(code, collapse = "\n"))
  cat("\n```\n</details>\n")
} else {
  cat("*Code example not available. Run `tar_make()` first.*\n")
}
```

### Exploring the Dictionary

```{r dd-explore-dict-code, results='asis'}
code <- safe_tar_read("code_dd_explore_dict")
if (!is.null(code)) {
  cat("<details>\n<summary>Show code</summary>\n\n```r\n")
  cat(paste(code, collapse = "\n"))
  cat("\n```\n</details>\n")
} else {
  cat("*Code example not available. Run `tar_make()` first.*\n")
}
```

## Reproducibility

<details>
<summary>Session Info (click to expand)</summary>

```{r session-info}
sessionInfo()
```

</details>
