---
title: "04. Differential Expression Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{04. Differential Expression Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)

safe_tar_read <- function(name) {
  tryCatch(
    targets::tar_read(name = name, store = targets::tar_config_get("store")),
    error = function(e) NULL
  )
}
```

## Overview

This vignette presents differential expression (DE) results from the CoMMpass
RNA-seq pipeline, comparing tumor vs normal or baseline vs relapse samples
using three complementary methods: **DESeq2**, **edgeR**, and **limma-voom**.

Visualizations include:

- **PCA** of VST-transformed counts for sample clustering
- **Volcano plot** showing fold-change vs significance
- **MA plot** showing mean expression vs fold-change
- **Heatmap** of top DE genes across samples
- **Method comparison** summarizing agreement across DE methods

> **Note:** This vignette was built in CI with `sample_limit=10`. Local builds
> default to 100 samples. Numbers below reflect the CI subset.

## Method Comparison

```{r method-summary}
de_summary <- safe_tar_read("de_method_summary")

if (!is.null(de_summary) && nrow(de_summary) > 0) {
  knitr::kable(
    de_summary,
    caption = paste0(
      "Differential expression results across methods. ",
      "Significance thresholds: |log2FC| > 1, adjusted p-value < 0.05. ",
      "DESeq2 uses Wald test with apeglm shrinkage; edgeR uses quasi-likelihood F-test; ",
      "limma uses empirical Bayes moderated t-test. ",
      "See [DESeq2 vignette](https://bioconductor.org/packages/DESeq2/) for details."
    ),
    col.names = c("Method", "Genes Tested", "Significant", "Up", "Down"),
    row.names = FALSE,
    align = "lrrrr"
  )
} else {
  cat("*DE method summary not available. Run `tar_make()` to build pipeline targets.*\n")
}
```

```{r method-barplot, fig.cap = "Number of significant DE genes identified by each method (|log2FC| > 1, padj < 0.05). Concordance across methods indicates robust signals. DESeq2 with apeglm shrinkage tends to be more conservative."}
if (!is.null(de_summary) && nrow(de_summary) > 0) {
  library(ggplot2)
  long <- rbind(
    data.frame(method = de_summary$method, direction = "Up",
               count = de_summary$n_up),
    data.frame(method = de_summary$method, direction = "Down",
               count = -de_summary$n_down)
  )
  ggplot(long, aes(x = method, y = count, fill = direction)) +
    geom_col() +
    scale_fill_manual(values = c("Up" = "#DC3545", "Down" = "#0066CC"),
                      name = "Direction") +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    labs(
      title = "DE Genes by Method",
      x = NULL,
      y = "Number of genes (down shown as negative)"
    ) +
    theme_minimal(base_size = 12)
}
```

## Principal Component Analysis

PCA is computed from the top 500 most variable genes after
[variance-stabilizing transformation](https://bioconductor.org/packages/DESeq2/)
(VST), which stabilizes variance across the expression range. VST is
preferred over raw counts or logCPM for exploratory visualization.

```{r pca-plot, fig.cap = "PCA of VST-transformed RNA-seq counts. Each point is a sample. Variance explained per PC shown on axes. Computed from top 500 most variable genes. See [DESeq2::vst()](https://bioconductor.org/packages/DESeq2/) for the transformation method."}
pca_data <- safe_tar_read("pca_data")

if (!is.null(pca_data) && nrow(pca_data$coords) > 0) {
  # Color by condition if available
  color_var <- NULL
  if ("condition" %in% names(pca_data$coords)) color_var <- "condition"

  p <- plot_pca(pca_data, color_by = color_var,
                title = "Sample PCA (VST)")
  print(p)
} else {
  cat("*PCA data not available. Run `tar_make()` to compute VST and PCA targets.*\n")
}
```

## Volcano Plot

The volcano plot shows statistical significance (-log10 adjusted p-value)
against biological effect size (log2 fold change). Genes in the upper corners
are both statistically significant and biologically meaningful.

```{r volcano, fig.cap = "Volcano plot of DESeq2 results. X-axis: log2 fold change (apeglm-shrunken if available). Y-axis: -log10 adjusted p-value. Dashed lines: |LFC| = 1 and padj = 0.05 thresholds. Top 10 genes labeled by adjusted p-value. Red = upregulated, blue = downregulated."}
volcano_data <- safe_tar_read("volcano_plot_data")

if (!is.null(volcano_data) && nrow(volcano_data) > 0) {
  p <- plot_volcano(
    volcano_data,
    lfc_threshold = 1,
    padj_threshold = 0.05,
    n_label = 10,
    title = "DESeq2 Volcano Plot"
  )
  print(p)
} else {
  cat("*Volcano plot data not available. Run `tar_make()` to build DE targets.*\n")
}
```

## MA Plot

The MA plot (Bland-Altman) shows mean expression (x-axis) against fold change
(y-axis). It reveals whether DE signals are concentrated at particular
expression levels and whether fold-change estimates are biased.

```{r ma-plot, fig.cap = "MA plot of DESeq2 results. X-axis: log10(baseMean + 1). Y-axis: log2 fold change. Red points are significant at padj < 0.05 and |LFC| > 1. Dashed lines at LFC = +/-1. Genes with low mean expression tend to have higher variance."}
ma_data <- safe_tar_read("ma_plot_data")

if (!is.null(ma_data) && nrow(ma_data) > 0) {
  p <- plot_ma(
    ma_data,
    lfc_threshold = 1,
    padj_threshold = 0.05,
    title = "DESeq2 MA Plot"
  )
  print(p)
} else {
  cat("*MA plot data not available. Run `tar_make()` to build DE targets.*\n")
}
```

## Heatmap of Top DE Genes

The heatmap shows Z-score scaled VST expression for the most significant
genes. Rows (genes) and columns (samples) are hierarchically clustered.

```{r heatmap, fig.height = 10, fig.cap = "Heatmap of top 50 DE genes by adjusted p-value. Expression values are VST-transformed and Z-score scaled per gene. Blue = below mean, red = above mean. Hierarchical clustering applied to both rows and columns. See [data dictionary](data-dictionary.html) for gene ID mappings."}
heatmap_data <- safe_tar_read("top_de_heatmap_data")

if (!is.null(heatmap_data) &&
    !is.null(heatmap_data$expr_matrix) &&
    !is.null(heatmap_data$de_results)) {
  p <- plot_heatmap_de(
    heatmap_data$expr_matrix,
    heatmap_data$de_results,
    n_genes = 50,
    title = "Top 50 DE Genes"
  )
  print(p)
} else {
  cat("*Heatmap data not available. Run `tar_make()` to build VST and DE targets.*\n")
}
```

## Consensus Genes

Genes identified as significant by all three methods (DESeq2, edgeR, limma)
represent the highest-confidence DE candidates.

```{r consensus}
consensus <- safe_tar_read("consensus_de_genes")

if (!is.null(consensus) && !is.null(consensus$n_consensus)) {
  cat(paste0(
    "**", consensus$n_consensus, " consensus DE genes** found across all three methods.\n\n"
  ))

  if (consensus$n_consensus > 0 && !is.null(consensus$consensus_genes)) {
    n_show <- min(20, length(consensus$consensus_genes))
    cat(paste0(
      "Top ", n_show, " consensus genes (sorted by mean rank): ",
      paste(utils::head(consensus$consensus_genes, n_show), collapse = ", "),
      if (consensus$n_consensus > n_show) paste0(", ... (", consensus$n_consensus - n_show, " more)") else "",
      "\n\n"
    ))
    cat("These genes are used for [pathway analysis](pathway-analysis.html) via ORA.\n")
  }
} else {
  cat("*Consensus DE genes not available. Run `tar_make()` to build DE targets.*\n")
}
```

## Paired Longitudinal DE

For patients with samples at multiple timepoints (baseline and relapse),
a paired design (`~ patient_id + visit`) controls for inter-patient
variability and tests for within-patient expression changes over time.

```{r paired-de}
paired_results <- safe_tar_read("deseq2_paired_results")

if (!is.null(paired_results) && !is.null(paired_results$results_table) &&
    nrow(paired_results$results_table) > 0) {
  n_patients <- paired_results$n_patients %||% NA
  n_samples <- paired_results$n_samples %||% NA
  n_deg <- paired_results$n_deg %||% 0

  cat(paste0(
    "**Paired analysis**: ", n_patients, " patients, ", n_samples, " samples\n\n",
    "**", n_deg, " DE genes** at padj < 0.05\n\n"
  ))

  # Show volcano for paired results
  p <- plot_volcano(
    paired_results$results_table,
    lfc_threshold = 1,
    padj_threshold = 0.05,
    n_label = 10,
    title = "Paired DE: Baseline vs Relapse"
  )
  print(p)
} else {
  cat("*Paired DE results not available (requires patients with >=2 timepoints).*\n")
}
```

## Next Steps

- **Pathway analysis**: Consensus DE genes are tested for enrichment in
  MSigDB Hallmark, KEGG, and other collections. See the
  [pathway analysis vignette](pathway-analysis.html).
- **Survival analysis**: DE gene signatures can be correlated with overall
  survival. See the [data acquisition vignette](data-acquisition.html) for
  clinical data integration.
