---
title: "03. Exploratory Data Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{03. Exploratory Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
library(targets)
library(dplyr)
library(ggplot2)

# --- Target store discovery (same pattern as data-acquisition.Rmd) ---
key <- "TAR_PROJECT"
proj_dir <- Sys.getenv(key)
targets_available <- FALSE

if (proj_dir != "" && dir.exists(file.path(proj_dir, "_targets"))) {
  tar_config_set(store = file.path(proj_dir, "_targets"))
  targets_available <- TRUE
} else if (dir.exists("_targets")) {
  targets_available <- TRUE
}

safe_tar_read <- function(name) {
  if (!targets_available) return(NULL)
  tryCatch(tar_read(name), error = function(e) NULL)
}
```

## Overview

This vignette provides an exploratory analysis of the MMRF CoMMpass data
retrieved by the pipeline. It covers patient demographics, biospecimen
characteristics, RNA-seq quality metrics, and shows how to query the
underlying parquet files with DuckDB.

All summaries are **pre-computed** as targets (`eda_*`). If you have not
yet run the pipeline, placeholder messages appear instead of plots.
Run `targets::tar_make()` to populate the data.

## Clinical Demographics

```{r clinical-demographics}
clin <- safe_tar_read("eda_clinical_summary")

if (!is.null(clin)) {
  cat("- **Patients:**", format(clin$n_patients, big.mark = ","), "\n")
  cat("- **Variables:**", clin$n_variables, "\n")
  cat("- **Data completeness:**",
      sprintf("%.1f%%", 100 * clin$completeness), "\n")
} else {
  cat("*Clinical summary not available. Run `tar_make()` first.*\n")
}
```

### Age at Diagnosis

```{r age-histogram, fig.cap = "Age at diagnosis distribution"}
if (!is.null(clin) && !is.null(clin$age_years) && length(clin$age_years) > 0) {
  age_df <- data.frame(age = clin$age_years)
  ggplot(age_df, aes(x = age)) +
    geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
    labs(x = "Age (years)", y = "Number of Patients",
         title = "Age at Diagnosis") +
    theme_minimal()
} else {
  cat("*Age data not available.*\n")
}
```

### Gender Distribution

```{r gender-table}
if (!is.null(clin) && !is.null(clin$gender_table)) {
  gt <- clin$gender_table
  gt$Percentage <- sprintf("%.1f%%", 100 * gt$Freq / sum(gt$Freq))
  knitr::kable(gt, caption = "Gender Distribution", row.names = FALSE)
} else {
  cat("*Gender data not available.*\n")
}
```

### Race Distribution

```{r race-table}
if (!is.null(clin) && !is.null(clin$race_table)) {
  rt <- clin$race_table
  rt$Percentage <- sprintf("%.1f%%", 100 * rt$Freq / sum(rt$Freq))
  knitr::kable(rt, caption = "Race Distribution", row.names = FALSE)
} else {
  cat("*Race data not available.*\n")
}
```

### Vital Status

```{r vital-status}
if (!is.null(clin) && !is.null(clin$vital_table)) {
  vt <- clin$vital_table
  vt$Percentage <- sprintf("%.1f%%", 100 * vt$Freq / sum(vt$Freq))
  knitr::kable(vt, caption = "Vital Status", row.names = FALSE)
} else {
  cat("*Vital status data not available.*\n")
}
```

## Survival Endpoints

```{r survival-endpoints}
if (!is.null(clin)) {
  if (!is.null(clin$days_to_death)) {
    cat("### Days to Death\n\n")
    cat("```\n")
    print(clin$days_to_death)
    cat("```\n\n")
  }

  if (!is.null(clin$vital_table)) {
    total <- sum(clin$vital_table$Freq)
    alive_row <- clin$vital_table[
      tolower(clin$vital_table$Status) %in% c("alive", "Alive"), , drop = FALSE
    ]
    if (nrow(alive_row) > 0) {
      cat("- **Censoring rate:**",
          sprintf("%.1f%%", 100 * sum(alive_row$Freq) / total),
          "(patients still alive at last follow-up)\n")
    }
  }

  if (!is.null(clin$days_to_last_followup)) {
    cat("\n### Days to Last Follow-up\n\n")
    cat("```\n")
    print(clin$days_to_last_followup)
    cat("```\n")
  }
} else {
  cat("*Survival endpoint data not available. Run `tar_make()` first.*\n")
}
```

## Biospecimen Overview

```{r biospecimen-overview}
bio <- safe_tar_read("eda_biospecimen_summary")

if (!is.null(bio)) {
  cat("- **Total samples:**", format(bio$n_samples, big.mark = ","), "\n")
  cat("- **Variables:**", bio$n_variables, "\n")
} else {
  cat("*Biospecimen summary not available. Run `tar_make()` first.*\n")
}
```

### Sample Types

```{r sample-type-bar, fig.cap = "Sample type distribution"}
if (!is.null(bio) && !is.null(bio$sample_type_table)) {
  st <- bio$sample_type_table
  ggplot(st, aes(x = reorder(Sample_Type, Freq), y = Freq)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    labs(x = NULL, y = "Count", title = "Sample Types") +
    theme_minimal()
} else {
  cat("*Sample type data not available.*\n")
}
```

### Samples per Patient

```{r samples-per-patient, fig.cap = "Samples per patient distribution"}
if (!is.null(bio) && !is.null(bio$samples_per_patient)) {
  spp_df <- data.frame(n = bio$samples_per_patient)
  ggplot(spp_df, aes(x = n)) +
    geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") +
    labs(x = "Number of Samples", y = "Number of Patients",
         title = "Samples per Patient") +
    theme_minimal()
} else {
  cat("*Samples-per-patient data not available.*\n")
}
```

## RNA-seq Quality

```{r rnaseq-quality}
rna <- safe_tar_read("eda_rnaseq_summary")

if (!is.null(rna)) {
  cat("- **Genes:**", format(rna$n_genes, big.mark = ","), "\n")
  cat("- **Samples:**", rna$n_samples, "\n")
  cat("- **Total counts:**", format(rna$total_counts, big.mark = ","), "\n")
  cat("- **Median library size:**",
      format(rna$median_library_size, big.mark = ","), "\n")
  cat("- **Sparsity (% zero entries):**",
      sprintf("%.1f%%", 100 * rna$sparsity), "\n")
} else {
  cat("*RNA-seq summary not available. Run `tar_make()` first.*\n")
}
```

### Library Size Distribution

```{r library-size-boxplot, fig.cap = "Library sizes across samples"}
if (!is.null(rna) && !is.null(rna$library_sizes)) {
  ls_df <- data.frame(library_size = rna$library_sizes / 1e6)
  ggplot(ls_df, aes(y = library_size)) +
    geom_boxplot(fill = "steelblue", width = 0.4) +
    labs(y = "Library Size (millions of reads)",
         title = "Library Size Distribution") +
    theme_minimal() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
} else {
  cat("*Library size data not available.*\n")
}
```

### Genes Detected per Sample

```{r genes-detected-histogram, fig.cap = "Genes detected per sample"}
if (!is.null(rna) && !is.null(rna$genes_detected)) {
  gd_df <- data.frame(genes = rna$genes_detected)
  ggplot(gd_df, aes(x = genes)) +
    geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
    labs(x = "Genes Detected (count > 0)", y = "Number of Samples",
         title = "Genes Detected per Sample") +
    theme_minimal()
} else {
  cat("*Genes-detected data not available.*\n")
}
```

## DuckDB Query Examples

The package provides two functions for querying the parquet files:

- `query_commpass_parquet()` -- one-shot query that returns a data frame
- `get_commpass_tbl()` -- returns a lazy `dplyr::tbl()` for chained operations

### Example: Simple Query

```r
# One-shot query (returns data frame)
clinical <- query_commpass_parquet("clinical")
head(clinical)

# With filters
females <- query_commpass_parquet(
  "clinical",
  filters = list(gender = "female")
)
```

```{r duckdb-demo-simple}
demo <- safe_tar_read("eda_duckdb_demo")

if (!is.null(demo)) {
  cat("**SQL equivalent:**\n\n")
  cat("```sql\n", demo$demo_sql, "\n```\n\n")
  knitr::kable(demo$demo_result, caption = "Sample query result (first 10 rows)")
} else {
  cat("*DuckDB demo not available. Run `tar_make()` first.*\n")
}
```

### Example: Aggregation with DuckDB

```r
# Lazy tbl for dplyr-style queries
con <- DBI::dbConnect(duckdb::duckdb())
clinical_tbl <- get_commpass_tbl("clinical", con = con)

result <- clinical_tbl |>
  filter(!is.na(gender)) |>
  group_by(gender) |>
  summarise(
    n = n(),
    mean_age_years = mean(age_at_diagnosis / 365.25, na.rm = TRUE)
  ) |>
  collect()

DBI::dbDisconnect(con, shutdown = TRUE)
```

```{r duckdb-demo-agg}
if (!is.null(demo)) {
  cat("**SQL equivalent:**\n\n")
  cat("```sql\n", demo$agg_sql, "\n```\n\n")
  knitr::kable(demo$agg_result, caption = "Aggregation result")
} else {
  cat("*DuckDB demo not available.*\n")
}
```

## Cross-dataset Integration

```{r integration}
integ <- safe_tar_read("eda_integration_summary")

if (!is.null(integ)) {
  cat("Patient ID overlap between clinical and biospecimen datasets:\n\n")
  cat("- **Clinical patients:**",
      format(integ$n_clinical_patients, big.mark = ","), "\n")
  cat("- **Biospecimen patients:**",
      format(integ$n_biospecimen_patients, big.mark = ","), "\n")
  cat("- **Shared (in both):**",
      format(integ$n_shared, big.mark = ","), "\n")
  cat("- **Clinical only:**",
      format(integ$n_clinical_only, big.mark = ","), "\n")
  cat("- **Biospecimen only:**",
      format(integ$n_biospecimen_only, big.mark = ","), "\n")
} else {
  cat("*Integration summary not available. Run `tar_make()` first.*\n")
}
```

## Reproducibility

<details>
<summary>Git Commit Info (click to expand)</summary>

```{r git-info, echo=FALSE}
if (requireNamespace("gert", quietly = TRUE)) {
  tryCatch({
    repo_path <- "."
    if (Sys.getenv("TAR_PROJECT") != "") repo_path <- Sys.getenv("TAR_PROJECT")

    info <- gert::git_info(repo = repo_path)
    log1 <- gert::git_log(max = 1, repo = repo_path)
    knitr::kable(data.frame(
      Item = c("Commit Hash", "Author", "Time", "Branch"),
      Value = c(info$commit, log1$author,
                as.character(log1$time), info$shorthand)
    ))
  }, error = function(e) {
    cat("Git info not available.")
  })
} else {
  cat("gert package not installed.")
}
```

</details>

<details>
<summary>Session Info (click to expand)</summary>

```{r session-info}
sessionInfo()
```

</details>
