---
title: "03. Exploratory Data Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{03. Exploratory Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
library(targets)
library(dplyr)
library(ggplot2)

# --- Target store discovery (same pattern as data-acquisition.Rmd) ---
# Discover targets store and project root
proj_dir <- Sys.getenv("TAR_PROJECT")
tar_store <- NULL
project_root <- "."

if (proj_dir != "" && dir.exists(file.path(proj_dir, "_targets"))) {
  tar_store <- file.path(proj_dir, "_targets")
  project_root <- proj_dir
} else if (dir.exists("_targets")) {
  tar_store <- "_targets"
  project_root <- "."
} else if (dir.exists("../_targets")) {
  tar_store <- "../_targets"
  project_root <- ".."
}

safe_tar_read <- function(name) {
  if (is.null(tar_store)) return(NULL)
  tryCatch(tar_read_raw(name, store = tar_store), error = function(e) NULL)
}
```

## Overview

This vignette provides an exploratory analysis of the MMRF CoMMpass data
retrieved by the pipeline. It covers patient demographics, biospecimen
characteristics, RNA-seq quality metrics, and shows how to query the
underlying parquet files with DuckDB.

All summaries are **pre-computed** as targets (`eda_*`). If you have not
yet run the pipeline, placeholder messages appear instead of plots.
Run `targets::tar_make()` to populate the data.

## Clinical Demographics

```{r clinical-demographics}
clin <- safe_tar_read("eda_clinical_summary")

if (!is.null(clin)) {
  cat("- **Patients:**", format(clin$n_patients, big.mark = ","), "\n")
  cat("- **Variables:**", clin$n_variables, "\n")
  cat("- **Data completeness:**",
      sprintf("%.1f%%", 100 * clin$completeness), "\n")
} else {
  cat("*Clinical summary not available. Run `tar_make()` first.*\n")
}
```

### Age at Diagnosis

```{r age-histogram, fig.cap = "Age at diagnosis distribution for CoMMpass patients. GDC stores age in days; values are converted to years (age / 365.25). Data: GDC clinical endpoint."}
if (!is.null(clin) && !is.null(clin$age_years) && length(clin$age_years) > 0) {
  age_df <- data.frame(age = clin$age_years)
  med_age <- sprintf("%.1f", median(clin$age_years, na.rm = TRUE))
  ggplot(age_df, aes(x = age)) +
    geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
    geom_vline(xintercept = median(clin$age_years, na.rm = TRUE),
               linetype = "dashed", colour = "red") +
    labs(x = "Age (years)", y = "Number of Patients",
         title = "Age at Diagnosis",
         subtitle = paste0("n = ", length(clin$age_years),
                           " patients, median = ", med_age, " years"),
         caption = "Data: GDC clinical. Dashed red line = median.") +
    theme_minimal()
} else {
  cat("*Age data not available.*\n")
}
```

### Gender Distribution

```{r gender-table}
if (!is.null(clin) && !is.null(clin$gender_table)) {
  gt <- clin$gender_table
  gt$Percentage <- sprintf("%.1f%%", 100 * gt$Freq / sum(gt$Freq))
  knitr::kable(
    gt, row.names = FALSE,
    caption = paste0(
      "Gender distribution for ", clin$n_patients,
      " CoMMpass patients. Data: [GDC clinical]",
      "(https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS)."
    )
  )
} else {
  cat("*Gender data not available.*\n")
}
```

### Race Distribution

```{r race-table}
if (!is.null(clin) && !is.null(clin$race_table)) {
  rt <- clin$race_table
  rt$Percentage <- sprintf("%.1f%%", 100 * rt$Freq / sum(rt$Freq))
  knitr::kable(
    rt, row.names = FALSE,
    caption = paste0(
      "Race distribution for ", clin$n_patients,
      " CoMMpass patients. Categories from GDC clinical data."
    )
  )
} else {
  cat("*Race data not available.*\n")
}
```

### Vital Status

```{r vital-status}
if (!is.null(clin) && !is.null(clin$vital_table)) {
  vt <- clin$vital_table
  vt$Percentage <- sprintf("%.1f%%", 100 * vt$Freq / sum(vt$Freq))
  knitr::kable(
    vt, row.names = FALSE,
    caption = paste0(
      "Vital status for ", clin$n_patients,
      " CoMMpass patients. Alive = censored at last follow-up. ",
      "Data: GDC clinical."
    )
  )
} else {
  cat("*Vital status data not available.*\n")
}
```

## ISS Staging

The International Staging System (ISS) classifies multiple myeloma into three
stages based on serum beta-2 microglobulin and albumin levels. Higher stages
indicate more advanced disease and worse prognosis.

```{r iss-distribution}
iss <- safe_tar_read("eda_iss_summary")

if (!is.null(iss) && isTRUE(iss$available)) {
  cat("- **Patients with ISS data:**", format(iss$n_with_iss, big.mark = ","), "\n")
  cat("- **Missing ISS:**", format(iss$n_missing, big.mark = ","), "\n")
  cat("- **Source column:**", iss$iss_col, "\n")
} else if (!is.null(iss)) {
  cat("*", iss$note, "*\n")
} else {
  cat("*ISS summary not available. Run `tar_make()` first.*\n")
}
```

### ISS Stage Distribution

```{r iss-table}
if (!is.null(iss) && isTRUE(iss$available)) {
  knitr::kable(
    iss$iss_table, row.names = FALSE,
    caption = paste0(
      "ISS stage distribution for ", iss$n_with_iss,
      " patients with staging data (",
      iss$n_missing, " missing). ",
      "ISS = [International Staging System]",
      "(https://doi.org/10.1200/JCO.2005.04.242) ",
      "based on serum beta-2 microglobulin and albumin."
    )
  )
}
```

```{r iss-barplot, fig.cap = "ISS stage distribution across CoMMpass patients. Higher stages indicate more advanced disease. ISS = International Staging System."}
if (!is.null(iss) && isTRUE(iss$available) && nrow(iss$iss_table) > 0) {
  iss_plot <- iss$iss_table[!is.na(iss$iss_table$ISS_Stage), ]
  if (nrow(iss_plot) > 0) {
    ggplot(iss_plot, aes(x = ISS_Stage, y = Freq)) +
      geom_col(fill = "steelblue") +
      labs(x = "ISS Stage", y = "Number of Patients",
           title = "ISS Stage Distribution",
           subtitle = paste0("n = ", sum(iss_plot$Freq), " patients with staging data"),
           caption = "ISS: International Staging System (Greipp et al., 2005)") +
      theme_minimal()
  }
}
```

### Age by ISS Stage

```{r iss-age}
if (!is.null(iss) && isTRUE(iss$available) && !is.null(iss$age_by_iss)) {
  age_tbl <- do.call(rbind, lapply(names(iss$age_by_iss), function(stage) {
    x <- iss$age_by_iss[[stage]]
    data.frame(
      ISS_Stage = stage, N = x$n,
      Mean_Age = sprintf("%.1f", x$mean),
      Median_Age = sprintf("%.1f", x$median),
      SD = sprintf("%.1f", x$sd)
    )
  }))
  knitr::kable(
    age_tbl, row.names = FALSE,
    caption = paste0(
      "Age at diagnosis (years) by ISS stage. ",
      "GDC stores age in days; converted via age / 365.25. ",
      "SD = standard deviation."
    )
  )
} else if (!is.null(iss) && isTRUE(iss$available)) {
  cat("*Age by ISS stage not available (age column not found).*\n")
}
```

## Survival Endpoints

```{r survival-endpoints}
if (!is.null(clin)) {
  if (!is.null(clin$days_to_death)) {
    cat("### Days to Death\n\n")
    cat("```\n")
    print(clin$days_to_death)
    cat("```\n\n")
  }

  if (!is.null(clin$vital_table)) {
    total <- sum(clin$vital_table$Freq)
    alive_row <- clin$vital_table[
      tolower(clin$vital_table$Status) %in% c("alive", "Alive"), , drop = FALSE
    ]
    if (nrow(alive_row) > 0) {
      cat("- **Censoring rate:**",
          sprintf("%.1f%%", 100 * sum(alive_row$Freq) / total),
          "(patients still alive at last follow-up)\n")
    }
  }

  if (!is.null(clin$days_to_last_followup)) {
    cat("\n### Days to Last Follow-up\n\n")
    cat("```\n")
    print(clin$days_to_last_followup)
    cat("```\n")
  }
} else {
  cat("*Survival endpoint data not available. Run `tar_make()` first.*\n")
}
```

## Biospecimen Overview

```{r biospecimen-overview}
bio <- safe_tar_read("eda_biospecimen_summary")

if (!is.null(bio)) {
  cat("- **Total samples:**", format(bio$n_samples, big.mark = ","), "\n")
  cat("- **Variables:**", bio$n_variables, "\n")
} else {
  cat("*Biospecimen summary not available. Run `tar_make()` first.*\n")
}
```

### Sample Types

```{r sample-type-bar, fig.cap = "Sample type distribution across all CoMMpass biospecimens. Data: GDC biospecimen endpoint."}
if (!is.null(bio) && !is.null(bio$sample_type_table)) {
  st <- bio$sample_type_table
  ggplot(st, aes(x = reorder(Sample_Type, Freq), y = Freq)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    labs(x = NULL, y = "Count",
         title = "Sample Types",
         subtitle = paste0("n = ", sum(st$Freq), " total biospecimens"),
         caption = "Data: GDC biospecimen") +
    theme_minimal()
} else {
  cat("*Sample type data not available.*\n")
}
```

### Samples per Patient

```{r samples-per-patient, fig.cap = "Number of biospecimen samples per patient. Most patients have multiple samples (longitudinal collection). Data: GDC biospecimen."}
if (!is.null(bio) && !is.null(bio$samples_per_patient)) {
  spp_df <- data.frame(n = bio$samples_per_patient)
  ggplot(spp_df, aes(x = n)) +
    geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") +
    labs(x = "Number of Samples", y = "Number of Patients",
         title = "Samples per Patient",
         subtitle = paste0("n = ", length(bio$samples_per_patient),
                           " patients, median = ",
                           median(bio$samples_per_patient), " samples"),
         caption = "Data: GDC biospecimen") +
    theme_minimal()
} else {
  cat("*Samples-per-patient data not available.*\n")
}
```

## RNA-seq Quality

```{r rnaseq-quality}
rna <- safe_tar_read("eda_rnaseq_summary")

if (!is.null(rna)) {
  cat("- **Genes:**", format(rna$n_genes, big.mark = ","), "\n")
  cat("- **Samples:**", rna$n_samples, "\n")
  cat("- **Total counts:**", format(rna$total_counts, big.mark = ","), "\n")
  cat("- **Median library size:**",
      format(rna$median_library_size, big.mark = ","), "\n")
  cat("- **Sparsity (% zero entries):**",
      sprintf("%.1f%%", 100 * rna$sparsity), "\n")
} else {
  cat("*RNA-seq summary not available. Run `tar_make()` first.*\n")
}
```

### Library Size Distribution

```{r library-size-boxplot, fig.cap = "Library sizes (millions of reads) across RNA-seq samples. Library size = total mapped read counts per sample. Data: GDC STAR-Counts."}
if (!is.null(rna) && !is.null(rna$library_sizes)) {
  ls_df <- data.frame(library_size = rna$library_sizes / 1e6)
  ggplot(ls_df, aes(y = library_size)) +
    geom_boxplot(fill = "steelblue", width = 0.4) +
    labs(y = "Library Size (millions of reads)",
         title = "Library Size Distribution",
         subtitle = paste0("n = ", rna$n_samples, " samples, median = ",
                           sprintf("%.1f", median(rna$library_sizes) / 1e6), "M"),
         caption = "Data: GDC STAR-Counts") +
    theme_minimal() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
} else {
  cat("*Library size data not available.*\n")
}
```

### Genes Detected per Sample

```{r genes-detected-histogram, fig.cap = "Number of genes detected (count > 0) per RNA-seq sample. Samples with unusually low detection may be outliers. Data: GDC STAR-Counts."}
if (!is.null(rna) && !is.null(rna$genes_detected)) {
  gd_df <- data.frame(genes = rna$genes_detected)
  ggplot(gd_df, aes(x = genes)) +
    geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
    labs(x = "Genes Detected (count > 0)", y = "Number of Samples",
         title = "Genes Detected per Sample",
         subtitle = paste0("n = ", rna$n_samples, " samples, median = ",
                           format(median(rna$genes_detected), big.mark = ","),
                           " genes"),
         caption = "Data: GDC STAR-Counts") +
    theme_minimal()
} else {
  cat("*Genes-detected data not available.*\n")
}
```

## DuckDB Query Examples

The package provides two functions for querying the parquet files:

- `query_commpass_parquet()` -- one-shot query that returns a data frame
- `get_commpass_tbl()` -- returns a lazy `dplyr::tbl()` for chained operations

### Example: Simple Query

```{r eda-simple-query-code, results='asis'}
code <- safe_tar_read("code_eda_simple_query")
if (!is.null(code)) {
  cat("<details>\n<summary>Show code</summary>\n\n```r\n")
  cat(paste(code, collapse = "\n"))
  cat("\n```\n</details>\n")
} else {
  cat("*Code example not available. Run `tar_make()` first.*\n")
}
```

```{r duckdb-demo-simple}
demo <- safe_tar_read("eda_duckdb_demo")

if (!is.null(demo)) {
  cat("**SQL equivalent:**\n\n")
  cat("```sql\n", demo$demo_sql, "\n```\n\n")
  knitr::kable(
    demo$demo_result, row.names = FALSE,
    caption = "First 10 patients from clinical data via DuckDB query. age_at_diagnosis is in days."
  )
} else {
  cat("*DuckDB demo not available. Run `tar_make()` first.*\n")
}
```

### Example: Aggregation with DuckDB

```{r eda-aggregation-code, results='asis'}
code <- safe_tar_read("code_eda_aggregation")
if (!is.null(code)) {
  cat("<details>\n<summary>Show code</summary>\n\n```r\n")
  cat(paste(code, collapse = "\n"))
  cat("\n```\n</details>\n")
} else {
  cat("*Code example not available. Run `tar_make()` first.*\n")
}
```

```{r duckdb-demo-agg}
if (!is.null(demo)) {
  cat("**SQL equivalent:**\n\n")
  cat("```sql\n", demo$agg_sql, "\n```\n\n")
  knitr::kable(
    demo$agg_result, row.names = FALSE,
    caption = "Patient count and mean age (years) by gender via DuckDB aggregation."
  )
} else {
  cat("*DuckDB demo not available.*\n")
}
```

## Cross-dataset Integration

```{r integration}
integ <- safe_tar_read("eda_integration_summary")

if (!is.null(integ)) {
  cat("Patient ID overlap between clinical and biospecimen datasets:\n\n")
  cat("- **Clinical patients:**",
      format(integ$n_clinical_patients, big.mark = ","), "\n")
  cat("- **Biospecimen patients:**",
      format(integ$n_biospecimen_patients, big.mark = ","), "\n")
  cat("- **Shared (in both):**",
      format(integ$n_shared, big.mark = ","), "\n")
  cat("- **Clinical only:**",
      format(integ$n_clinical_only, big.mark = ","), "\n")
  cat("- **Biospecimen only:**",
      format(integ$n_biospecimen_only, big.mark = ","), "\n")
} else {
  cat("*Integration summary not available. Run `tar_make()` first.*\n")
}
```

## Reproducibility

<details>
<summary>Git Commit Info (click to expand)</summary>

```{r git-info, echo=FALSE}
if (requireNamespace("gert", quietly = TRUE)) {
  tryCatch({
    repo_path <- "."
    if (Sys.getenv("TAR_PROJECT") != "") repo_path <- Sys.getenv("TAR_PROJECT")

    info <- gert::git_info(repo = repo_path)
    log1 <- gert::git_log(max = 1, repo = repo_path)
    knitr::kable(data.frame(
      Item = c("Commit Hash", "Author", "Time", "Branch"),
      Value = c(info$commit, log1$author,
                as.character(log1$time), info$shorthand)
    ), row.names = FALSE)
  }, error = function(e) {
    cat("Git info not available.")
  })
} else {
  cat("gert package not installed.")
}
```

</details>

<details>
<summary>Session Info (click to expand)</summary>

```{r session-info}
sessionInfo()
```

</details>
