---
title: "03. Exploratory Data Analysis"
format: html
vignette: >
  %\VignetteIndexEntry{03. Exploratory Data Analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
in_pkgdown <- nzchar(Sys.getenv("IN_PKGDOWN"))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  eval = !in_pkgdown,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
if (!in_pkgdown) library(targets)

# Discover targets store
proj_dir <- Sys.getenv("TAR_PROJECT")
tar_store <- NULL
if (proj_dir != "" && dir.exists(file.path(proj_dir, "_targets"))) {
  tar_store <- file.path(proj_dir, "_targets")
} else if (dir.exists("_targets")) {
  tar_store <- "_targets"
} else if (dir.exists("../_targets")) {
  tar_store <- "../_targets"
}

safe_tar_read <- function(name) {
  if (!is.null(tar_store)) {
    val <- tryCatch(tar_read_raw(name, store = tar_store), error = function(e) NULL)
    if (!is.null(val) && !(is.character(val) && length(val) == 1 && grepl("not available", val, ignore.case = TRUE))) {
      return(val)
    }
  }
  rds_candidates <- c(
    system.file(paste0("extdata/vignettes/", name, ".rds"), package = "coMMpass"),
    file.path("inst/extdata/vignettes", paste0(name, ".rds")),
    file.path("../inst/extdata/vignettes", paste0(name, ".rds"))
  )
  for (rds in rds_candidates) {
    if (nzchar(rds) && file.exists(rds)) return(readRDS(rds))
  }
  NULL
}
```

```{r pkgdown-banner, results='asis', eval=in_pkgdown, echo=FALSE}
cat("::: {.callout-note}\n## Online documentation\nThis vignette shows pre-computed results from the targets pipeline. Some outputs may appear as placeholders in the online version.\n:::\n")
```

## Overview

> See the [Glossary](glossary.html) for term definitions used throughout this project.

- Patient demographics, biospecimen characteristics, and RNA-seq quality metrics
- DuckDB queries against underlying parquet files
- Cytogenetic alteration frequencies and co-occurrence patterns

All summaries are **pre-computed** as targets (`vig_*`). Placeholder messages
appear in place of plots when pre-computed outputs are not yet available.

## Clinical Demographics

```{r clinical-demographics, results='asis'}
safe_tar_read("vig_clinical_demographics_text")
```

### Age at Diagnosis

```{r age-histogram, fig.cap = "Age at diagnosis distribution for CoMMpass patients. GDC stores age in days; values are converted to years (age / 365.25). Data: GDC clinical endpoint."}
safe_tar_read("vig_age_histogram")
```

### Gender Distribution

```{r gender-table}
safe_tar_read("vig_gender_table")
```

### Race Distribution

```{r race-table}
safe_tar_read("vig_race_table")
```

### Vital Status

```{r vital-status}
safe_tar_read("vig_vital_table")
```

## ISS Staging

The International Staging System (ISS) classifies multiple myeloma into three
stages based on serum beta-2 microglobulin and albumin levels. Higher stages
indicate more advanced disease and worse prognosis.

```{r iss-distribution, results='asis'}
safe_tar_read("vig_iss_distribution_text")
```

### ISS Stage Distribution

```{r iss-table}
safe_tar_read("vig_iss_table")
```

```{r iss-barplot, fig.cap = "ISS stage distribution across CoMMpass patients. Higher stages indicate more advanced disease. ISS = International Staging System."}
safe_tar_read("vig_iss_barplot")
```

### Age by ISS Stage

```{r iss-age}
safe_tar_read("vig_iss_age_table")
```

## Survival Endpoints

```{r survival-endpoints, results='asis'}
safe_tar_read("vig_survival_endpoints")
```

## Biospecimen Overview

```{r biospecimen-overview, results='asis'}
safe_tar_read("vig_biospecimen_overview_text")
```

### Sample Types

```{r sample-type-bar, fig.cap = "Sample type distribution across all CoMMpass biospecimens. Data: GDC biospecimen endpoint."}
safe_tar_read("vig_sample_type_barplot")
```

### Samples per Patient

```{r samples-per-patient, fig.cap = "Number of biospecimen samples per patient. Most patients have multiple samples (longitudinal collection). Data: GDC biospecimen."}
safe_tar_read("vig_samples_per_patient_hist")
```

## RNA-seq Quality

```{r rnaseq-quality, results='asis'}
safe_tar_read("vig_rnaseq_quality_text")
```

### Library Size Distribution

```{r library-size-boxplot, fig.cap = "Library sizes (millions of reads) across RNA-seq samples. Library size = total mapped read counts per sample. Data: GDC STAR-Counts."}
safe_tar_read("vig_library_size_boxplot")
```

### Genes Detected per Sample

```{r genes-detected-histogram, fig.cap = "Number of genes detected (count > 0) per RNA-seq sample. Samples with unusually low detection may be outliers. Data: GDC STAR-Counts."}
safe_tar_read("vig_genes_detected_hist")
```

## Sample Clustering (PCA)

Principal component analysis on the top 500 most variable genes (VST-transformed)
reveals sample-level structure. Points are colored by clinical covariates to check
for batch effects or biological groupings.

```{r pca-biplot, fig.cap = "PCA biplot of RNA-seq samples using top 500 most variable genes (VST). PC1 and PC2 with variance explained. Colored by gender. Clustering by sex indicates strong expression differences."}
safe_tar_read("vig_pca_biplot")
```

```{r pca-biplot-vital, fig.cap = "PCA biplot colored by vital status. Overlap between alive/dead suggests survival is not the dominant source of transcriptomic variation."}
safe_tar_read("vig_pca_biplot_vital")
```

## Cytogenetic Landscape

Cytogenetic markers from FISH testing are key to myeloma risk stratification
([IMWG 2014 criteria](https://doi.org/10.1200/JCO.2014.55.1519)).
High-risk alterations include t(4;14), t(14;16), del(17p), and gain(1q).

### Alteration Frequencies

```{r cyto-frequency}
safe_tar_read("vig_cyto_frequency_table")
```

### Oncoprint

```{r cyto-oncoprint, fig.height = 5, fig.cap = "Cytogenetic oncoprint showing alteration landscape across patients. Rows: FISH markers. Columns: patients sorted by alteration frequency. Blue = translocation, red = deletion, green = gain. Percentages show alteration frequency among tested patients."}
safe_tar_read("vig_cyto_oncoprint")
```

### Co-occurrence Analysis

```{r cyto-cooccurrence}
safe_tar_read("vig_cyto_cooccurrence_table")
```

```{r cyto-cooccurrence-heatmap, fig.cap = "Pairwise co-occurrence heatmap. Red: co-occurrence (positive association). Blue: mutual exclusivity (negative association). Color intensity: -log10(adjusted p-value). * = significant at padj < 0.05."}
safe_tar_read("vig_cyto_cooccurrence_heatmap")
```

Cytogenetic markers are used for [survival stratification](survival-analysis.html) and inform [differential expression](differential-expression.html) analysis.

## DuckDB Query Examples

The package provides two functions for querying the parquet files:

- `query_commpass_parquet()` -- one-shot query that returns a data frame
- `get_commpass_tbl()` -- returns a lazy `dplyr::tbl()` for chained operations

### Example: Simple Query

```{r eda-simple-query-code, results='asis'}
safe_tar_read("code_eda_simple_query")
```

```{r duckdb-demo-simple}
safe_tar_read("vig_duckdb_demo_simple")
```

### Example: Aggregation with DuckDB

```{r eda-aggregation-code, results='asis'}
safe_tar_read("code_eda_aggregation")
```

```{r duckdb-demo-agg}
safe_tar_read("vig_duckdb_demo_agg")
```

## Cross-dataset Integration

```{r integration, results='asis'}
safe_tar_read("vig_integration_text")
```

## Data Sources

Results in this vignette are derived from the
[MMRF CoMMpass study](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS)
(MMRF-COMMPASS, ~1,143 patients), downloaded via
[TCGAbiolinks](https://bioconductor.org/packages/TCGAbiolinks/).
The pipeline runs with a configurable `sample_limit` (default 100; CI uses 10).

For full citations, data access tiers, and the distinction between
pipeline data and synthetic test data, see the
[Data Sources](data-sources.html) vignette.

## Recent Changes

Recent project commits with lines added, files changed, and change categories.

```{r changelog}
safe_tar_read("vig_git_changelog")
```

## Reproducibility

<details>
<summary>Git Commit Info (click to expand)</summary>

```{r git-info}
git_info <- safe_tar_read("vig_git_info")
if (!is.null(git_info)) git_info else cat("*Git info available after pipeline run.*\n")
```

</details>

<details>
<summary>Session Info (click to expand)</summary>

```{r session-info, eval=TRUE}
sessionInfo()
```

</details>
