---
title: "Gene Report"
format:
  html:
    code-fold: true
    code-summary: "Show code"
params:
  gene: "CD70"
vignette: >
  %\VignetteIndexEntry{Gene Report}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
in_pkgdown <- nzchar(Sys.getenv("IN_PKGDOWN"))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !in_pkgdown,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
if (!in_pkgdown) library(targets)

# Discover targets store
proj_dir <- Sys.getenv("TAR_PROJECT")
tar_store <- NULL
if (proj_dir != "" && dir.exists(file.path(proj_dir, "_targets"))) {
  tar_store <- file.path(proj_dir, "_targets")
} else if (dir.exists("_targets")) {
  tar_store <- "_targets"
} else if (dir.exists("../_targets")) {
  tar_store <- "../_targets"
}

safe_tar_read <- function(name) {
  if (!is.null(tar_store)) {
    val <- tryCatch(tar_read_raw(name, store = tar_store), error = function(e) NULL)
    if (!is.null(val) && !(is.character(val) && length(val) == 1 && grepl("not available", val, ignore.case = TRUE))) {
      return(val)
    }
  }
  rds_candidates <- c(
    system.file(paste0("extdata/vignettes/", name, ".rds"), package = "coMMpass"),
    file.path("inst/extdata/vignettes", paste0(name, ".rds")),
    file.path("../inst/extdata/vignettes", paste0(name, ".rds"))
  )
  for (rds in rds_candidates) {
    if (nzchar(rds) && file.exists(rds)) {
      obj <- readRDS(rds)
      if (is.data.frame(obj) && requireNamespace("DT", quietly = TRUE))
        return(DT::datatable(obj, rownames = FALSE))
      if (!is.null(obj)) return(obj)
    }
  }
  invisible(NULL)
}

gene <- params$gene
```

```{r pkgdown-banner}
#| results: asis
#| eval: !expr in_pkgdown
#| echo: false
cat("::: {.callout-note}\n## Online documentation\nThis vignette displays pre-computed results. Run the targets pipeline locally for interactive analysis.\n:::\n")
```

## `r gene` — Single Gene Characterization

This report provides a multi-faceted characterization of **`r gene`** across
the CoMMpass cohort: expression distribution, survival stratification,
cytogenetic context, and co-expression correlations.

> Generated with `gene_report("`r gene`")`. See `?gene_report` for usage.

## Expression Distribution

VST-normalized expression values across all samples, showing the distribution shape and central tendency.

```{r expression-dist, fig.cap = paste0(gene, " expression distribution across samples (VST-transformed).")}
vst_counts <- safe_tar_read("vst_counts")
if (!is.null(vst_counts) && gene %in% rownames(SummarizedExperiment::assay(vst_counts, "vst"))) {
  vst_mat <- SummarizedExperiment::assay(vst_counts, "vst")
  expr_vals <- as.numeric(vst_mat[gene, ])
  if (requireNamespace("ggplot2", quietly = TRUE)) {
    df <- data.frame(expression = expr_vals)
    ggplot2::ggplot(df, ggplot2::aes(x = expression)) +
      ggplot2::geom_histogram(bins = 30, fill = "#2166AC", alpha = 0.7) +
      ggplot2::geom_vline(xintercept = median(expr_vals), linetype = "dashed",
                           color = "#B2182B") +
      ggplot2::labs(
        title = paste0(gene, " Expression Distribution"),
        subtitle = paste0("n = ", length(expr_vals), " samples | ",
                           "Median = ", round(median(expr_vals), 2),
                           " | SD = ", round(sd(expr_vals), 2)),
        x = "VST Expression", y = "Count"
      ) +
      ggplot2::theme_minimal(base_size = 12)
  }
} else {
  cat(paste0("Gene ", gene, " not found in VST matrix or targets store not available."))
}
```

## Survival by Expression

Patients split at median `r gene` expression into High/Low groups.

```{r km-expression, fig.cap = paste0("Kaplan-Meier curves stratified by median ", gene, " expression.")}
survival_data <- safe_tar_read("survival_data")
if (!is.null(vst_counts) && !is.null(survival_data) &&
    gene %in% rownames(SummarizedExperiment::assay(vst_counts, "vst"))) {
  vst_mat <- SummarizedExperiment::assay(vst_counts, "vst")
  km_res <- coMMpass::run_km_by_expression(
    survival_data, vst_mat, gene = gene, split = "median"
  )
  if (!is.null(km_res$fit)) {
    coMMpass::plot_km(km_res,
      title = paste0("Survival by ", gene, " Expression (Median Split)"))
  } else {
    cat(paste0("KM analysis not possible: ", km_res$note))
  }
} else {
  cat("Survival or expression data not available in targets store.")
}
```

## Expression by Cytogenetic Subtype

Boxplots comparing expression levels across FISH-defined cytogenetic marker groups (t(4;14), t(14;16), del(17p)).

```{r expr-subtype, fig.cap = paste0(gene, " expression by cytogenetic marker status.")}
cyto_data <- safe_tar_read("cytogenetic_data")
if (!is.null(vst_counts) && !is.null(cyto_data) &&
    gene %in% rownames(SummarizedExperiment::assay(vst_counts, "vst"))) {
  vst_mat <- SummarizedExperiment::assay(vst_counts, "vst")
  cyto_df <- if (is.character(cyto_data) && file.exists(cyto_data)) {
    arrow::read_parquet(cyto_data)
  } else if (is.data.frame(cyto_data)) {
    cyto_data
  } else {
    NULL
  }
  if (!is.null(cyto_df)) {
    coMMpass::plot_expression_by_subtype(vst_mat, cyto_df, gene = gene)
  } else {
    cat("Cytogenetic data not available.")
  }
} else {
  cat("Expression or cytogenetic data not available in targets store.")
}
```

## Gene-Gene Correlations

Top correlated genes with `r gene` (Pearson, VST expression).

```{r correlations}
if (!is.null(vst_counts) &&
    gene %in% rownames(SummarizedExperiment::assay(vst_counts, "vst"))) {
  vst_mat <- SummarizedExperiment::assay(vst_counts, "vst")

  # Top 500 most variable genes as candidates
  gene_vars <- apply(vst_mat, 1, var)
  top_candidates <- names(sort(gene_vars, decreasing = TRUE))[1:500]
  top_candidates <- setdiff(top_candidates, gene)

  cor_batch <- coMMpass::correlate_genes_batch(
    vst_mat, target_gene = gene,
    candidate_genes = top_candidates, method = "pearson"
  )
  if (nrow(cor_batch) > 0) {
    top10 <- head(cor_batch, 10)
    top10_display <- top10[, c("gene", "estimate", "p_value", "padj")]
    if (requireNamespace("DT", quietly = TRUE)) {
      DT::datatable(
        top10_display, filter = "top", rownames = FALSE,
        colnames = c("Gene", "Pearson R", "p-value", "Adjusted p"),
        options = list(pageLength = 10, scrollX = TRUE),
        caption = htmltools::tags$caption(
          style = "caption-side: top; text-align: left;",
          paste0("Top 10 genes correlated with ", gene,
                 " (n = ", top10$n_samples[1], " samples)")
        )
      ) |> DT::formatRound(columns = c("estimate", "p_value", "padj"),
                            digits = c(3, 4, 4))
    } else {
      knitr::kable(
        top10_display,
        digits = c(0, 3, 4, 4),
        col.names = c("Gene", "Pearson R", "p-value", "Adjusted p"),
        caption = paste0("Top 10 genes correlated with ", gene,
                          " (n = ", top10$n_samples[1], " samples)")
      )
    }
  } else {
    cat("No correlation results available.")
  }
} else {
  cat("VST data not available in targets store.")
}
```

```{r top-correlation-plot, fig.cap = paste0("Scatter plot of ", gene, " vs top correlated gene.")}
if (exists("cor_batch") && nrow(cor_batch) > 0) {
  top_gene <- cor_batch$gene[1]
  cor_res <- coMMpass::correlate_genes(vst_mat, gene, top_gene)
  coMMpass::plot_gene_correlation(cor_res)
}
```

## Data Sources

Results in this vignette are derived from the
[MMRF CoMMpass study](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS)
(MMRF-COMMPASS, ~1,143 patients), downloaded via
[TCGAbiolinks](https://bioconductor.org/packages/TCGAbiolinks/).
The pipeline runs with a configurable `sample_limit` (default 200; CI uses 20).

For full citations, data access tiers, and the distinction between
pipeline data and synthetic test data, see the
[Data Sources](data-sources.html) vignette.

## Recent Changes

Recent project commits with lines added, files changed, and change categories.

```{r changelog}
safe_tar_read("vig_git_changelog")
```

## Reproducibility

<details>
<summary>Session Info (click to expand)</summary>

```{r session-info, eval=TRUE}
sessionInfo()
```

</details>
