---
title: "Gene Report"
format:
  html:
    code-fold: true
    code-summary: "Show code"
params:
  gene: "CD70"
vignette: >
  %\VignetteIndexEntry{Gene Report}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
in_pkgdown <- nzchar(Sys.getenv("IN_PKGDOWN"))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
if (!in_pkgdown) library(targets)

# Discover targets store
proj_dir <- Sys.getenv("TAR_PROJECT")
tar_store <- NULL
if (proj_dir != "" && dir.exists(file.path(proj_dir, "_targets"))) {
  tar_store <- file.path(proj_dir, "_targets")
} else if (dir.exists("_targets")) {
  tar_store <- "_targets"
} else if (dir.exists("../_targets")) {
  tar_store <- "../_targets"
}

safe_tar_read <- function(name) {
  render_val <- function(val) {
    if (inherits(val, c("gtable", "grob", "gTree"))) {
      grid::grid.newpage(); grid::grid.draw(val); return(invisible(val))
    }
    val
  }
  if (!is.null(tar_store)) {
    val <- tryCatch(tar_read_raw(name, store = tar_store), error = function(e) NULL)
    if (!is.null(val) && !(is.character(val) && length(val) == 1 && grepl("not available", val, ignore.case = TRUE))) {
      return(render_val(val))
    }
  }
  rds_candidates <- c(
    system.file(paste0("extdata/vignettes/", name, ".rds"), package = "coMMpass"),
    file.path("inst/extdata/vignettes", paste0(name, ".rds")),
    file.path("../inst/extdata/vignettes", paste0(name, ".rds"))
  )
  for (rds in rds_candidates) {
    if (nzchar(rds) && file.exists(rds)) {
      obj <- readRDS(rds)
      if (is.data.frame(obj) && requireNamespace("DT", quietly = TRUE))
        return(DT::datatable(obj, rownames = FALSE))
      if (!is.null(obj)) return(render_val(obj))
    }
  }
  invisible(NULL)
}

gene <- params$gene
```

```{r pkgdown-banner}
#| results: asis
#| eval: !expr in_pkgdown
#| echo: false
cat("::: {.callout-note}\n## Online documentation\nThis vignette displays pre-computed results for the default gene (CD70). Run the targets pipeline locally for interactive gene-specific analysis.\n:::\n")
```

## `r gene` — Single Gene Characterization

This report provides a multi-faceted characterization of **`r gene`** across
the CoMMpass cohort: expression distribution, survival stratification,
cytogenetic context, and co-expression correlations.

> Generated with `gene_report("`r gene`")`. See `?gene_report` for usage.

## Expression Distribution

VST-normalized expression values across all samples, showing the distribution shape and central tendency.

```{r expression-dist, fig.cap = paste0(gene, " expression distribution across samples (VST-transformed).")}
safe_tar_read("vig_gene_expression_dist")
```

## Survival by Expression

Patients split at median `r gene` expression into High/Low groups.

```{r km-expression, fig.cap = paste0("Kaplan-Meier curves stratified by median ", gene, " expression.")}
safe_tar_read("vig_gene_km_expression")
```

## Expression by Cytogenetic Subtype

Boxplots comparing expression levels across FISH-defined cytogenetic marker groups (t(4;14), t(14;16), del(17p)).

```{r expr-subtype, fig.cap = paste0(gene, " expression by cytogenetic marker status.")}
safe_tar_read("vig_gene_expr_subtype")
```

## Gene-Gene Correlations

Top correlated genes with `r gene` (Pearson, VST expression).

```{r correlations}
safe_tar_read("vig_gene_correlations")
```

```{r top-correlation-plot, fig.cap = paste0("Scatter plot of ", gene, " vs top correlated gene.")}
safe_tar_read("vig_gene_correlation_plot")
```

## Data Sources

Results in this vignette are derived from the
[MMRF CoMMpass study](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS)
(MMRF-COMMPASS, ~1,143 patients), downloaded via
[TCGAbiolinks](https://bioconductor.org/packages/TCGAbiolinks/).
The pipeline runs with a configurable `sample_limit` (default 200; CI uses 20).

For full citations, data access tiers, and the distinction between
pipeline data and synthetic test data, see the
[Data Sources](data-sources.html) vignette.

## Recent Changes

Recent project commits with lines added, files changed, and change categories.

```{r changelog}
safe_tar_read("vig_git_changelog")
```

## Reproducibility

<details>
<summary>Session Info (click to expand)</summary>

```{r session-info, eval=TRUE}
sessionInfo()
```

</details>
