---
title: "06. Glossary"
format:
  html:
    toc: true
    toc-expand: 2
    toc-location: left
    code-fold: true
    code-summary: "Show code"
vignette: >
  %\VignetteIndexEntry{06. Glossary}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
in_pkgdown <- nzchar(Sys.getenv("IN_PKGDOWN"))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
if (!in_pkgdown) library(targets)

# Shared vignette utilities (safe_tar_read, show_target, render helpers)
utils_path <- system.file("vignette_utils.R", package = "coMMpass")
if (utils_path == "") {
  utils_path <- if (file.exists("../inst/vignette_utils.R")) "../inst/vignette_utils.R"
  else if (file.exists("inst/vignette_utils.R")) "inst/vignette_utils.R"
  else stop("Cannot find vignette_utils.R")
}
source(utils_path, local = TRUE)
```

```{r pkgdown-banner}
#| results: asis
#| eval: !expr in_pkgdown
#| echo: false
cat("::: {.callout-note}\n## Online documentation\nThis vignette displays pre-computed results. Run the targets pipeline locally for interactive analysis.\n:::\n")
```

## Definitions

Terms used throughout the CoMMpass analysis vignettes, grouped by category. See individual vignettes for detailed usage.

```{r glossary-table}
safe_tar_read("glossary_table")
```

## Units Reference {#units-reference}

**Age and time variables are stored in DAYS by GDC.** This is the single most common source of errors in GDC analyses. See the [data dictionary](data-dictionary.html) for all variable definitions.

```{r units-table}
safe_tar_read("vig_units_table")
```

## Recent Changes

Recent project commits with lines added, files changed, and change categories.

```{r changelog}
safe_tar_read("vig_git_changelog")
```

## Reproducibility

<details>
<summary>Session Info (click to expand)</summary>

```{r session-info, eval=TRUE}
sessionInfo()
```

</details>
