---
title: "Pipeline DAG"
format: html
vignette: >
  %\VignetteIndexEntry{Pipeline DAG}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
in_pkgdown <- nzchar(Sys.getenv("IN_PKGDOWN"))
knitr::opts_chunk$set(
  echo = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
if (!in_pkgdown) library(targets)

proj_dir <- Sys.getenv("TAR_PROJECT")
tar_store <- NULL
if (proj_dir != "" && dir.exists(file.path(proj_dir, "_targets"))) {
  tar_store <- file.path(proj_dir, "_targets")
} else if (dir.exists("_targets")) {
  tar_store <- "_targets"
} else if (dir.exists("../_targets")) {
  tar_store <- "../_targets"
}

safe_tar_read <- function(name) {
  if (!is.null(tar_store)) {
    val <- tryCatch(tar_read_raw(name, store = tar_store), error = function(e) NULL)
    if (!is.null(val) && !(is.character(val) && length(val) == 1 && grepl("not available", val, ignore.case = TRUE))) {
      return(val)
    }
  }
  # RDS fallback: try system.file, then relative paths from package root / vignettes/
  rds_candidates <- c(
    system.file(paste0("extdata/vignettes/", name, ".rds"), package = "coMMpass"),
    file.path("inst/extdata/vignettes", paste0(name, ".rds")),
    file.path("../inst/extdata/vignettes", paste0(name, ".rds"))
  )
  for (rds in rds_candidates) {
    if (nzchar(rds) && file.exists(rds)) return(readRDS(rds))
  }
  NULL
}
```

```{r pkgdown-banner, results='asis', eval=in_pkgdown, echo=FALSE}
cat("::: {.callout-note}\n## Online documentation\nThis vignette shows pre-computed results from the targets pipeline.\n:::\n")
```

## Pipeline Overview

The CoMMpass analysis pipeline is organized into isolated layers.
Each layer corresponds to a functional area of the analysis, with
explicit allowed dependencies between layers.

| Layer | Description |
|---|---|
| data-acquisition | Download RNA-seq and clinical data from GDC |
| data-cleaning | Standardize and integrate datasets |
| cytogenetics | Extract FISH markers and classify risk |
| quality-control | Sample QC, filtering, normalization |
| differential-expression | DESeq2, edgeR, limma-voom analysis |
| survival | Kaplan-Meier, Cox PH regression |
| pathway | GSEA, ORA, gene annotation |
| eda | Exploratory summaries |
| storage | DuckDB/parquet querying |
| api | Plumber API for programmatic access |
| documentation | Vignette pre-computed outputs |
| infrastructure | Nix sync, pkgctx, DAG validation |

## DAG Validation

Validation checks on the targets pipeline DAG: no orphaned targets, no circular dependencies, and all vignette targets are reachable from raw data inputs. A passing validation confirms the pipeline is structurally sound.

```{r dag-validation, results='asis'}
res <- safe_tar_read("dag_layer_validation")
if (!is.null(res)) res
```

## Target Network

Interactive DAG showing the targets pipeline network with layer-based coloring. Nodes are colored by functional layer. Hover for target details, use the dropdown to search by name.

```{r pipeline-visnetwork, results='asis'}
res <- safe_tar_read("vig_pipeline_dag")
if (!is.null(res)) res
```

## Interactive DAG (live)

The interactive DAG below is rendered directly from the pipeline metadata
via `tar_visnetwork()`. This requires a local targets store.

```{r visnetwork-live, fig.height=8}
#| eval: !expr '!in_pkgdown && !is.null(tar_store) && dir.exists(tar_store)'
tryCatch(
  targets::tar_visnetwork(targets_only = TRUE, label = "time"),
  error = function(e) cat("*Pipeline store unavailable for live DAG.*\n")
)
```

For the full interactive DAG explorer with filtering, coloring, and
selection, see `inst/shinylive/pipeline_dag.qmd`.

## Recent Changes

Recent project commits with lines added, files changed, and change categories.

```{r changelog, results='asis'}
res <- safe_tar_read("vig_git_changelog")
if (!is.null(res)) res
```

## Reproducibility

<details>
<summary>Session Info (click to expand)</summary>

```{r session-info, eval=TRUE}
sessionInfo()
```

</details>
