---
title: "Pipeline DAG"
format: html
vignette: >
  %\VignetteIndexEntry{Pipeline DAG}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
in_pkgdown <- nzchar(Sys.getenv("IN_PKGDOWN"))
knitr::opts_chunk$set(
  echo = FALSE,
  eval = !in_pkgdown,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
if (!in_pkgdown) {
  library(targets)
  store_candidates <- c(
    here::here("_targets"),
    file.path(Sys.getenv("TARGETS_STORE", unset = "_targets")),
    "_targets"
  )
  for (store in store_candidates) {
    if (dir.exists(store)) {
      Sys.setenv(TAR_STORE = store)
      break
    }
  }
}

safe_tar_read <- function(name, store = Sys.getenv("TAR_STORE", "_targets")) {
  val <- tryCatch(
    targets::tar_read_raw(name, store = store),
    error = function(e) NULL
  )
  if (is.character(val) && length(val) == 1 && grepl("not available", val, ignore.case = TRUE)) {
    return(NULL)
  }
  val
}
```

```{r pkgdown-banner, results='asis', eval=in_pkgdown, echo=FALSE}
cat("::: {.callout-note}\n## Online documentation\nPlots and tables in this vignette are generated from the `targets` pipeline. Run `targets::tar_make()` locally to see full output.\n:::\n")
```

## Pipeline Overview

The CoMMpass analysis pipeline is organized into isolated layers.
Each layer corresponds to a functional area of the analysis, with
explicit allowed dependencies between layers.

| Layer | Description |
|---|---|
| data-acquisition | Download RNA-seq and clinical data from GDC |
| data-cleaning | Standardize and integrate datasets |
| cytogenetics | Extract FISH markers and classify risk |
| quality-control | Sample QC, filtering, normalization |
| differential-expression | DESeq2, edgeR, limma-voom analysis |
| survival | Kaplan-Meier, Cox PH regression |
| pathway | GSEA, ORA, gene annotation |
| eda | Exploratory summaries |
| storage | DuckDB/parquet querying |
| api | Plumber API for programmatic access |
| documentation | Vignette pre-computed outputs |
| infrastructure | Nix sync, pkgctx, DAG validation |

## DAG Validation

```{r dag-validation}
safe_tar_read("dag_layer_validation")
```

## Target Network

```{r pipeline-visnetwork}
safe_tar_read("vig_pipeline_dag")
```

## Interactive DAG (live)

The interactive DAG below is rendered directly from the pipeline metadata
via `tar_visnetwork()`. This is the ONE exception to the zero-computation
rule: it produces an htmlwidget that cannot be serialized as a target.

```{r visnetwork-live, fig.height=8}
#| eval: !expr 'dir.exists(Sys.getenv("TAR_STORE", "_targets"))'
tryCatch(
  targets::tar_visnetwork(targets_only = TRUE, label = "time"),
  error = function(e) cat("*Pipeline store not available for live DAG.*\n")
)
```

For the full interactive DAG explorer with filtering, coloring, and
selection, see `inst/shinylive/pipeline_dag.qmd`.
