---
title: "CoMMpass"
subtitle: "Genomics of 1,143 multiple myeloma patients over 8 years"
author: "John Gavin"
format:
  revealjs:
    theme: [dark, custom.scss]
    slide-number: true
    width: 1280
    height: 720
    embed-resources: true
    fig-align: center
    transition: fade
    footer: "[johngavin.github.io/coMMpass-analysis](https://johngavin.github.io/coMMpass-analysis/)"
    title-slide-attributes:
      data-notes: "CoMMpass = Clinical Outcomes in Multiple Myeloma to Personal Assessment of Genetic Profile"
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(ggplot2)
library(targets)
tar_store <- if (dir.exists("../_targets")) "../_targets" else if (dir.exists("_targets")) "_targets" else NULL
safe_read <- function(name) {
  if (!is.null(tar_store)) {
    val <- tryCatch(targets::tar_read_raw(name, store = tar_store), error = function(e) NULL)
    if (!is.null(val)) return(val)
  }
  for (p in c(paste0("../inst/extdata/vignettes/", name, ".rds"),
              paste0("inst/extdata/vignettes/", name, ".rds"))) {
    if (file.exists(p)) return(readRDS(p))
  }
  NULL
}
show_grob <- function(val) {
  if (is.null(val) || !inherits(val, c("gtable", "grob", "gTree"))) return(invisible(NULL))
  p <- ggplot() + annotation_custom(val) + theme_void() +
    theme(plot.background = element_rect(fill = "black", color = NA))
  print(p)
}
```

```{=html}
<script>
// Move title attrs to data-tip to prevent duplicate native tooltips
document.addEventListener('DOMContentLoaded', function() {
  document.querySelectorAll('abbr[title]').forEach(function(el) {
    el.setAttribute('data-tip', el.getAttribute('title'));
    el.removeAttribute('title');
  });
});
</script>
```

## The patients

- <abbr title="CoMMpass = Clinical Outcomes in Multiple Myeloma to Personal Assessment of Genetic Profile.
A study by the MMRF (Multiple Myeloma Research Foundation) that tracked 1,143 patients from diagnosis.
Not a drug trial — an observational study. Patients received standard-of-care treatment.
Example: patient enrolled at diagnosis in 2012, followed with blood tests and bone marrow biopsies until 2019.">CoMMpass</abbr> enrolled **1,143 newly diagnosed <abbr title="Multiple myeloma: a blood cancer where plasma cells in the bone marrow grow uncontrollably.
Normal plasma cells make antibodies to fight infection.
Myeloma cells crowd out healthy cells, causing bone pain, kidney damage, and low blood counts.
Example: a patient's bone marrow biopsy shows 60% plasma cells instead of the normal 2–3%.">myeloma</abbr> patients** (2011–2019)
- Median age at diagnosis: **69 years** (range: 27–93)
- 8-year follow-up with <abbr title="RNA sequencing: a technology that counts how many copies of each gene's RNA are present in a sample.
More copies = gene is more active.
Example: gene TP53 has 500 RNA copies in one patient but 50 in another — it is 10× more active.">RNA-seq</abbr>, clinical records, and treatment data
- This is an **observational study**, not a drug trial — patients received whatever treatment their doctors chose

| Measurement | Example from the data |
|-------------|----------------------|
| Age | <abbr title="MMRF = Multiple Myeloma Research Foundation. MMRF_2240 is an anonymised patient identifier. 71 is the patient's age in years at diagnosis (converted from days — the database stores age in days).">MMRF_2240</abbr>: 71 years at diagnosis |
| <abbr title="International Staging System: classifies myeloma severity using two blood tests.
Stage I: albumin ≥ 3.5 g/dL AND beta-2-microglobulin < 3.5 mg/L (low tumour burden).
Stage III: beta-2-microglobulin ≥ 5.5 mg/L (high tumour burden).
Stage II: everything in between.
Example: a patient with albumin 2.8 and B2M 8.0 is Stage III.">ISS</abbr> stage | I: 348 patients, II: 353, III: 266 |
| Vital status | 803 alive, 191 dead at last follow-up |
| Top drugs | Dexamethasone (1,760 records), Lenalidomide (1,393), Bortezomib (1,076) |

## What 60,660 genes means

- The human genome has ~20,000 protein-coding genes
- The other ~40,000 are non-coding <abbr title="RNA = ribonucleic acid. Non-coding RNAs do not make proteins but regulate other genes. Types include lncRNA (long non-coding), miRNA (micro), snoRNA (small nucleolar). Example: lncRNA MALAT1 is often overexpressed in myeloma and may promote tumour growth.">RNA</abbr> (regulatory genes, pseudogenes)
- Every patient had the **same 60,660 genes measured** — what differs is how active each gene is
- Genes are spread across 23 chromosome pairs:
  - Chromosome 1 has ~2,000 genes; chromosome 21 has ~200
  - Myeloma-relevant: chromosome 14 (immunoglobulin genes), chromosome 17 (TP53 tumour suppressor)
- Known myeloma pathway genes: ~200–500 (from <abbr title="MSigDB = Molecular Signatures Database. A curated collection of gene groups. The 50 Hallmark sets define well-studied biological processes. Example: HALLMARK_MYC_TARGETS contains 200 genes regulated by the MYC oncogene, frequently overactive in myeloma.">MSigDB</abbr>)
- Each patient was sampled **once** (at diagnosis, before treatment)
  - Some patients had a second sample at relapse (~100 paired samples)
  - This study does not track gene changes continuously over time

## The data we have (part 1)

- **<abbr title="RNA sequencing counts: the number of RNA fragments detected for each gene.
Higher count = gene is more active.
Example: housekeeping gene ACTB has ~50,000 counts (always active).
Rare gene FOXP3 has ~12 counts (mostly silent).">RNA-seq</abbr> expression data** — 100 patient samples, 60,660 genes each
  - Measured once at diagnosis (some patients also at relapse)
  - Source: <abbr title="Genomic Data Commons: the US National Cancer Institute's open data portal. Free to access. Example: visit portal.gdc.cancer.gov, search MMRF-COMMPASS, download clinical or expression files.">GDC</abbr> <abbr title="Spliced Transcripts Alignment to a Reference: the software used to count how many RNA fragments map to each gene.">STAR</abbr> counts (60.5 <abbr title="Megabytes: 1 MB ≈ 500 pages of text.">MB</abbr>)
- **Clinical metadata** — 995 patients, 88 variables each
  - Demographics (age, sex, race), staging (<abbr title="International Staging System">ISS</abbr>), outcomes (alive/dead, days to death)
  - Source: <abbr title="Genomic Data Commons">GDC</abbr> clinical endpoint (1.2 <abbr title="Megabytes">MB</abbr>)

## The data we have (part 2)

- **Treatment records** — 7,184 drug administration records for 994 patients
  - Which drug, which line of therapy (1st, 2nd, 3rd...), start and end dates
  - Most patients received combination therapy (e.g. VRd = Bortezomib + Lenalidomide + Dexamethasone)
  - Patients who relapsed switched to later lines of therapy and stayed in the study
  - Source: extracted from <abbr title="Genomic Data Commons">GDC</abbr> <abbr title="Application Programming Interface: a way for software to request data from a server. Example: our pipeline sends a request to api.gdc.cancer.gov asking for treatment records, and gets back structured data.">API</abbr> (34 <abbr title="Kilobytes">KB</abbr>)
- **<abbr title="Molecular Signatures Database: 708 predefined groups of genes with known biological functions. Bundled in the R package — not downloaded from GDC.">MSigDB</abbr> gene sets** — 708 predefined gene groups (bundled in the R package)

## The data we don't have

| Missing data | Why it matters | Why it's missing |
|-------------|----------------|-----------------|
| <abbr title="FISH = Fluorescence In Situ Hybridisation. A lab test using fluorescent probes to detect chromosome changes. Example: t(4;14) means chromosomes 4 and 14 swapped pieces — this happens in ~15% of myeloma patients and predicts worse outcomes.">FISH</abbr> markers | Classify patients as high-risk vs standard-risk | Not in <abbr title="Genomic Data Commons">GDC</abbr> open data; stored in <abbr title="MMRF = Multiple Myeloma Research Foundation. The organisation that runs the CoMMpass study.">MMRF</abbr> Gateway |
| <abbr title="Progression-Free Survival: time from treatment start until disease worsens or patient dies. Example: PFS of 24 months means treatment controlled cancer for 2 years before it returned.">PFS</abbr> endpoints | Measure how long treatment works | Not in <abbr title="Genomic Data Commons">GDC</abbr>; requires <abbr title="MMRF = Multiple Myeloma Research Foundation">MMRF</abbr> Gateway |
| Treatment response | Did the drug work? (complete response, partial, stable, progression) | <abbr title="Genomic Data Commons">GDC</abbr> field exists but is empty for all patients |
| <abbr title="Biomarkers: blood test results like creatinine (kidney function), LDH (cell damage), albumin (nutrition). Example: creatinine > 2 mg/dL suggests kidney damage from myeloma.">Biomarker</abbr> lab values | Track disease progression via blood tests | Stored in nested <abbr title="Genomic Data Commons">GDC</abbr> data structures |

- Nested <abbr title="Genomic Data Commons">GDC</abbr> fields: the biomarker values exist but are inside `molecular_tests` sub-records, not flat columns
  - Each patient has multiple test records at different dates
  - Parsing requires matching tests by date and type — not a simple column extraction
- 15 of 250 pipeline targets return empty because of these gaps
- Registration submitted to <abbr title="MMRF = Multiple Myeloma Research Foundation">MMRF</abbr> Researcher Gateway

## Who survives longer? Staging matters

```{r km-iss}
#| fig-height: 4.5
show_grob(safe_read("vig_km_iss"))
```

- <abbr title="ISS = International Staging System">ISS</abbr> Stage III patients die at **5× the rate** of Stage I (<abbr title="Hazard Ratio: the relative risk of death compared to a reference group. HR = 1 means same risk. HR = 2 means twice the risk. HR = 0.5 means half the risk. Example: HR = 4.9 for ISS III vs I means Stage III patients die at nearly 5× the rate.">HR</abbr> = 4.9, p < 0.001)
- 994 patients total: 803 alive, 191 dead at last follow-up
- The three curves separate early and stay separated — staging predicts long-term outcome

## What predicts survival?

- A <abbr title="Cox Proportional Hazards: a statistical model that estimates how much each factor (age, sex, staging) changes the risk of death, while accounting for different follow-up times. Example: if ISS stage has HR = 4.9, that means ISS III patients have 4.9× the death rate of ISS I patients, after adjusting for age and sex.">Cox regression</abbr> adjusting for age, sex, and <abbr title="International Staging System">ISS</abbr> stage:

| Factor | <abbr title="Hazard Ratio: relative risk of death. HR = 1: no effect. HR > 1: increased risk. HR < 1: decreased risk. Example: HR = 1.5 for male sex means men have 50% higher death rate than women in this data.">HR</abbr> | What it means |
|--------|---:|---------|
| <abbr title="International Staging System">ISS</abbr> II vs I | 2.4 | 2.4× higher death rate than Stage I |
| <abbr title="International Staging System">ISS</abbr> III vs I | 4.9 | Nearly 5× higher death rate |
| Male vs female | 1.5 | 50% higher death rate for men |
| Age (per year) | Not significant | After adjusting for staging, age adds little |

- <abbr title="Concordance index: how well the model predicts who dies first. C = 0.5: coin flip (useless). C = 0.7: correctly ranks 70% of patient pairs. C = 1.0: perfect. Example: C = 0.687 means if you pick two random patients, the model correctly predicts which dies first 69% of the time.">C-index</abbr> = 0.687 — moderate discrimination (better than random, not perfect)

## Which genes differ between alive and dead patients?

```{r volcano}
#| fig-height: 4.5
show_grob(safe_read("vig_volcano_plot"))
```

- **1,615 genes** significantly differ (out of 60,660 tested)
- Each dot is one gene; x-axis = size of change, y-axis = confidence
- Grey dots: below significance thresholds; coloured dots: significant

## What those gene changes mean (part 1)

- <abbr title="log₂ Fold Change: how much a gene's activity differs between two groups. log₂FC = 1 means 2× change. log₂FC = 2 means 4× change. We use log₂ (not plain ratio) because it makes increases and decreases symmetric: +1 = doubled, −1 = halved. Example: gene with 1,000 counts in alive patients and 2,000 in dead has log₂FC = 1.">log₂FC</abbr> measures how much each gene changed between alive and dead patients
- **Typical significant gene**: log₂FC ≈ 0.8
  - A 1.7× expression difference — subtle but real
  - Example: gene X has 1,000 <abbr title="RNA sequencing">RNA</abbr> counts in alive patients, 1,700 in dead
  - This kind of modest change, multiplied across hundreds of genes, shifts biological pathways
- **Extreme gene**: log₂FC = 8.0
  - A 256× difference — almost certainly a measurement artefact
  - Usually from a gene with very few counts in one group (e.g. 1 vs 256)
  - These are filtered out or flagged for manual review

## What those gene changes mean (part 2)

- **Thresholds** (how we defined "significant"):
  - |log₂FC| > 1: at least a 2× change in either direction
  - Adjusted p-value < 0.05: controls false discoveries across 60,660 tests
  - At these thresholds: expect ~80 false positives (5% of 1,615)
- The <abbr title="Volcano plot: named because the shape of significant points rising on both sides resembles a volcanic eruption. It is NOT a contour plot — there are no contour lines. Each dot is one gene. The grey region in the middle contains genes that changed too little (below fold-change threshold) or with too little confidence (below significance threshold) to be called significant.">volcano plot</abbr> shows all 60,660 genes at once
  - Most cluster in the grey centre: small changes or low confidence
  - The 1,615 coloured dots are in the upper corners: big change AND high confidence

## Which biological pathways are disrupted?

- Instead of looking at one gene at a time, <abbr title="GSEA = Gene Set Enrichment Analysis. Tests whether genes in a known biological group (e.g. 'cell death genes') are collectively shifted in our data. Example: 100 of 161 apoptosis genes are slightly upregulated — no single gene is dramatic, but the GROUP shift is statistically significant. This reveals biology that single-gene analysis misses.">GSEA</abbr> tests groups of related genes together
- We tested 50 <abbr title="Hallmark gene sets: 50 well-defined biological processes curated by MSigDB. Each represents a specific cellular function. Example: HALLMARK_APOPTOSIS (161 genes for programmed cell death), HALLMARK_HYPOXIA (200 genes activated when oxygen is low). Unlike broader gene set collections (thousands of overlapping sets), Hallmark sets are non-redundant and biologically interpretable.">Hallmark</abbr> pathways

```{r gsea}
#| fig-height: 3.5
show_grob(safe_read("vig_gsea_dotplot"))
```

- Each dot represents one pathway; position shows how strongly the pathway genes are shifted in dead vs alive patients

## Can we prove causation?

```{r dag}
#| fig-height: 4
show_grob(safe_read("vig_causal_dag_plot"))
```

- This <abbr title="DAG = Directed Acyclic Graph. A diagram mapping assumed cause-and-effect relationships. Arrows mean 'A causes B'. If there is a hidden common cause (confounder) of both A and B, we cannot prove A causes B — even if they are correlated. Example: ice cream sales and drowning deaths correlate, but the common cause is hot weather.">DAG</abbr> maps our causal assumptions
- <abbr title="Cytogenetics: the study of chromosome structure and abnormalities. In myeloma, specific chromosome changes (translocations, deletions) predict outcomes. Example: deletion of part of chromosome 17 (del17p) removes the TP53 tumour suppressor gene, making the cancer more aggressive.">Chromosome changes</abbr> → survival: **cannot be proven** because unmeasured tumour biology affects both
- <abbr title="International Staging System">ISS</abbr> → survival: **can be estimated** if we adjust for age, chromosome changes, and <abbr title="Gene expression: how active each gene is, measured by counting RNA molecules. Genes can be highly expressed (thousands of copies) or barely expressed (a few copies). Example: the MYC oncogene is highly expressed in aggressive myeloma.">gene expression</abbr>

## Missing data tells a story

| Variable | What "missing" means | <abbr title="Hazard Ratio: relative risk of death.">HR</abbr> | p-value | What it means |
|----------|---------------------|---:|--------:|------|
| Cause of death | Patient is alive | ≈ 0 | 2 × 10⁻²⁶ | Missing = alive |

- Patients without a cause of death recorded are alive — the gap in the data **is** the answer
- The p-value of 2 × 10⁻²⁶ means:
  - The probability of seeing this association by chance is 0.00000000000000000000000002
  - With 994 patients (803 alive, 191 dead), the association between "missing cause of death" and "being alive" is essentially perfect
  - This is expected — it would be remarkable if it were NOT significant
- This is <abbr title="MNAR = Missing Not At Random. The reason data is missing depends on the value itself. Example: cause of death is missing BECAUSE the patient is alive — the missingness IS the information. Contrast with MCAR (Missing Completely At Random) where data is missing for unrelated reasons like a dropped test tube.">MNAR</abbr>: the missingness is informative, not noise

## Summary

| What we measured | Result |
|------------------|--------|
| Patients analysed | 994 of 1,143 enrolled |
| Genes tested | 60,660 per patient |
| Significant genes (alive vs dead) | 1,615 |
| <abbr title="International Staging System">ISS</abbr> III vs I death rate | 4.9× higher |
| Model discrimination | <abbr title="Concordance index">C-index</abbr> = 0.687 (69%) |

- **Website**: [johngavin.github.io/coMMpass-analysis](https://johngavin.github.io/coMMpass-analysis/)
- **Code**: [github.com/JohnGavin/coMMpass-analysis](https://github.com/JohnGavin/coMMpass-analysis)
- **Data**: [GDC portal](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS) (open access)
- **Study**: [NCT01454297](https://clinicaltrials.gov/study/NCT01454297)
