---
title: "CoMMpass: Multiple Myeloma Data Analysis"
subtitle: "Inputs, Outputs, and What the Data Tells Us"
author: "John Gavin"
format:
  revealjs:
    theme: [dark, custom.scss]
    slide-number: true
    width: 1280
    height: 720
    embed-resources: true
    code-overflow: wrap
    fig-align: center
    transition: fade
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(targets)
tar_store <- if (dir.exists("../_targets")) "../_targets" else if (dir.exists("_targets")) "_targets" else NULL
safe_read <- function(name) {
  if (!is.null(tar_store)) {
    val <- tryCatch(targets::tar_read_raw(name, store = tar_store), error = function(e) NULL)
    if (!is.null(val)) return(val)
  }
  for (p in c(paste0("../inst/extdata/vignettes/", name, ".rds"),
              paste0("inst/extdata/vignettes/", name, ".rds"))) {
    if (file.exists(p)) return(readRDS(p))
  }
  NULL
}
```

# What is Multiple Myeloma? {.inverse}

## The Disease

:::: {.columns}
::: {.column width="60%"}
- Cancer of **plasma cells** in bone marrow
- ~35,000 new US cases/year
- Median age at diagnosis: **69 years**
- 5-year survival: **~58%** (improving)

. . .

**Key question**: Which patients do well, and why?
:::
::: {.column width="40%"}
```{mermaid}
%%| fig-width: 5
flowchart TD
    BM[Bone Marrow] --> PC[Plasma Cells]
    PC --> |normal| AB[Antibodies<br/>immune defence]
    PC --> |cancer| MM[Myeloma Cells]
    MM --> BD[Bone Damage]
    MM --> KD[Kidney Damage]
    MM --> IS[Immune<br/>Suppression]
    style MM fill:#e74c3c,color:white
    style AB fill:#2ecc71,color:white
```
:::
::::

## The CoMMpass Study

| | |
|---|---|
| **Patients** | 1,143 newly diagnosed |
| **Follow-up** | 8 years (2011–2019) |
| **RNA-seq samples** | 859 (STAR counts) |
| **Clinical variables** | 88 per patient |
| **Treatment records** | 7,184 (from GDC API) |
| **GDC files** | 34,109 total |

. . .

> The largest longitudinal genomics study of multiple myeloma ever conducted.

# What Data Did We Get? {.inverse}

## Data Access: 4 Tiers

| Tier | What | We Use It? |
|------|------|-----------|
| **GDC Open** (API) | RNA-seq, clinical, treatment | ✅ Yes |
| **AWS S3** (open) | RNA-seq files | ✅ Yes |
| **MMRF Gateway** | FISH, PFS, response | ❌ Pending |
| **dbGaP** | Raw BAM/FASTQ | ❌ No IRB |

. . .

**12 of 250 pipeline targets are NULL** because MMRF Gateway data is unavailable.

## What We Downloaded

```{r data-flow}
#| fig-height: 4
# Simple data flow diagram as a plot
df <- data.frame(
  source = c("GDC API", "GDC API", "GDC API", "Pre-bundled"),
  dataset = c("RNA-seq\n(STAR counts)", "Clinical\nmetadata", "Treatment\nrecords", "MSigDB\ngene sets"),
  records = c(100, 995, 7184, 708),
  size_mb = c(60.5, 1.2, 0.03, 0.08),
  stringsAsFactors = FALSE
)
p <- ggplot2::ggplot(df, ggplot2::aes(x = reorder(dataset, -records), y = records, fill = source)) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::geom_text(ggplot2::aes(label = paste0(format(records, big.mark = ","), "\nrecords")),
                     vjust = -0.3, color = "white", size = 4) +
  ggplot2::scale_fill_manual(values = c("GDC API" = "#3498db", "Pre-bundled" = "#9b59b6")) +
  ggplot2::labs(title = "Datasets in the Pipeline",
                x = NULL, y = "Records", fill = "Source") +
  ggplot2::theme_minimal(base_size = 16) +
  ggplot2::theme(
    plot.background = ggplot2::element_rect(fill = "#1a1a2e", color = NA),
    panel.background = ggplot2::element_rect(fill = "#1a1a2e", color = NA),
    panel.grid = ggplot2::element_line(color = "#2d2d44"),
    text = ggplot2::element_text(color = "white"),
    axis.text = ggplot2::element_text(color = "#b0b0b0", size = 12),
    legend.background = ggplot2::element_rect(fill = "#1a1a2e", color = NA),
    legend.text = ggplot2::element_text(color = "white")
  )
print(p)
```

# The Pipeline {.inverse}

## 250 Targets, 7 Minutes

```{r pipeline-stats}
#| fig-height: 4.5
ps <- safe_read("readme_pipeline_summary")
if (!is.null(ps)) {
  df <- data.frame(
    Category = c("Vignette display", "Code provenance", "Shiny precompute",
                 "Data pipeline", "Survival", "DE + pathway", "Other"),
    Count = c(107, 36, 16, 8, 8, 16, 59),
    stringsAsFactors = FALSE
  )
  p <- ggplot2::ggplot(df, ggplot2::aes(x = reorder(Category, Count), y = Count)) +
    ggplot2::geom_col(fill = "#3498db", width = 0.7) +
    ggplot2::geom_text(ggplot2::aes(label = Count), hjust = -0.3, color = "white", size = 5) +
    ggplot2::coord_flip() +
    ggplot2::labs(title = paste0(ps$n_targets, " Pipeline Targets (build time: ", ps$total_time, ")"),
                  x = NULL, y = "Targets") +
    ggplot2::theme_minimal(base_size = 16) +
    ggplot2::theme(
      plot.background = ggplot2::element_rect(fill = "#1a1a2e", color = NA),
      panel.background = ggplot2::element_rect(fill = "#1a1a2e", color = NA),
      panel.grid = ggplot2::element_line(color = "#2d2d44"),
      text = ggplot2::element_text(color = "white"),
      axis.text = ggplot2::element_text(color = "#b0b0b0", size = 14)
    )
  print(p)
}
```

## Input → Output: The Chain

```{mermaid}
%%| fig-width: 10
flowchart LR
    subgraph Input
      GDC[GDC API<br/>995 patients]
    end
    subgraph Clean
      SE[Summarized<br/>Experiment]
      CD[Clinical<br/>DataFrame]
    end
    subgraph Analyse
      DE[1615 DE genes]
      KM[KM curves]
      COX[Cox PH<br/>C=0.687]
    end
    subgraph Output
      VIG[14 vignettes]
      RDS[156 RDS files]
      WEB[pkgdown site]
    end
    GDC --> SE --> DE --> VIG
    GDC --> CD --> KM --> VIG
    CD --> COX --> VIG
    VIG --> RDS --> WEB
    style Input fill:#2ecc71,color:white
    style Clean fill:#3498db,color:white
    style Analyse fill:#e74c3c,color:white
    style Output fill:#9b59b6,color:white
```

# Survival Analysis {.inverse}

## Kaplan-Meier: Overall Survival

```{r km-overall}
#| fig-height: 5
val <- safe_read("vig_km_overall")
if (inherits(val, c("gtable", "grob"))) {
  p <- ggplot2::ggplot() + ggplot2::annotation_custom(val) +
    ggplot2::theme_void() +
    ggplot2::theme(plot.background = ggplot2::element_rect(fill = "#1a1a2e", color = NA))
  print(p)
}
```

**994 patients. Median OS not reached** — more than half alive at last follow-up.

## What "Median Not Reached" Means

:::: {.columns}
::: {.column width="50%"}
### Typical (our data)
- 994 patients
- 191 deaths (19%)
- **Median: NA** (not reached)
- *Good news: most patients alive*
:::
::: {.column width="50%"}
### Extreme example
- If all 994 died by day 30:
- Median = **30 days**
- Survival curve drops to 0 immediately
- *This is what aggressive cancers look like*
:::
::::

. . .

> The fact that median is NOT reached means modern myeloma therapies are working.

## ISS Staging: Who Does Worse?

```{r km-iss}
#| fig-height: 5
val <- safe_read("vig_km_iss")
if (inherits(val, c("gtable", "grob"))) {
  p <- ggplot2::ggplot() + ggplot2::annotation_custom(val) +
    ggplot2::theme_void() +
    ggplot2::theme(plot.background = ggplot2::element_rect(fill = "#1a1a2e", color = NA))
  print(p)
}
```

ISS III patients have **~5x the hazard** of ISS I (HR=4.9, p<0.001).

## Cox Regression: What Predicts Survival?

| Covariate | HR | 95% CI | p |
|-----------|---:|--------|---|
| **ISS II** | 2.35 | [1.51, 3.67] | <0.001 |
| **ISS III** | 4.90 | [3.21, 7.48] | <0.001 |
| **Male** | 1.45 | [1.07, 1.97] | 0.018 |
| Age (per year) | NA | — | — |

. . .

**C-index = 0.687** — the model discriminates better than a coin flip (0.5) but not perfectly (1.0).

# Differential Expression {.inverse}

## 1,615 DE Genes

```{r volcano}
#| fig-height: 5
val <- safe_read("vig_volcano_plot")
if (inherits(val, c("gtable", "grob"))) {
  p <- ggplot2::ggplot() + ggplot2::annotation_custom(val) +
    ggplot2::theme_void() +
    ggplot2::theme(plot.background = ggplot2::element_rect(fill = "#1a1a2e", color = NA))
  print(p)
}
```

**Alive vs Dead** comparison: 1,615 genes at |log2FC|>1, padj<0.05.

## Typical vs Extreme: Log2 Fold Change

:::: {.columns}
::: {.column width="50%"}
### Typical DE gene
- log2FC = **0.8** (1.7x expression change)
- padj = 0.01
- *Modest but real biological signal*
- Most DE genes look like this
:::
::: {.column width="50%"}
### Extreme DE gene
- log2FC = **8.0** (256x change!)
- padj = 1e-50
- *Suspiciously large — check for:*
  - Low-count artefact
  - Single outlier sample
  - Batch effect
:::
::::

. . .

> Rule of thumb: log2FC > 4 in bulk RNA-seq warrants manual inspection.

## Pathway Enrichment: What Pathways?

```{r gsea}
#| fig-height: 5
val <- safe_read("vig_gsea_dotplot")
if (inherits(val, c("gtable", "grob"))) {
  p <- ggplot2::ggplot() + ggplot2::annotation_custom(val) +
    ggplot2::theme_void() +
    ggplot2::theme(plot.background = ggplot2::element_rect(fill = "#1a1a2e", color = NA))
  print(p)
}
```

# Causal Thinking {.inverse}

## The Causal DAG

```{r dag}
#| fig-height: 5
val <- safe_read("vig_causal_dag_plot")
if (inherits(val, c("gtable", "grob"))) {
  p <- ggplot2::ggplot() + ggplot2::annotation_custom(val) +
    ggplot2::theme_void() +
    ggplot2::theme(plot.background = ggplot2::element_rect(fill = "#1a1a2e", color = NA))
  print(p)
}
```

**Key insight**: cytogenetic risk → survival is **not identifiable** due to unmeasured tumor biology.

## Adjustment Sets

| Causal Question | Identifiable? | Adjust For |
|-----------------|:------------:|------------|
| Cytogenetic risk → OS | ❌ No | Latent confounders |
| ISS stage → OS | ✅ Yes | {age, cytogenetics, expression} |
| Gene expression → OS | ✅ Yes | {age, cytogenetics, ISS} |
| Treatment → OS | ❌ No | Latent confounders |

. . .

> You cannot estimate the causal effect of cytogenetic risk on survival without measuring tumor biology directly.

# Missingness as Signal {.inverse}

## Missing Data Tells a Story

```{r missingness}
#| fig-height: 4.5
val <- safe_read("vig_missingness_heatmap")
if (inherits(val, c("gtable", "grob"))) {
  p <- ggplot2::ggplot() + ggplot2::annotation_custom(val) +
    ggplot2::theme_void() +
    ggplot2::theme(plot.background = ggplot2::element_rect(fill = "#1a1a2e", color = NA))
  print(p)
}
```

## The cause_of_death Finding

| Variable | HR | p-value | Interpretation |
|----------|---:|---------|---------------|
| cause_of_death | ≈0 | 2e-26 | Missing → **alive** |

. . .

**This is obvious in hindsight**: patients without a cause of death are alive. But the statistical test confirms it — this is **MNAR** (missing not at random). The missingness IS the signal.

. . .

> Typical: missingness is noise. Extreme: missingness IS the variable.

# Bayesian vs Frequentist {.inverse}

## Same Data, Two Frameworks

| Covariate | Frequentist HR | Bayesian HR |
|-----------|---------------:|------------:|
| **Male** | 1.466 [1.08, 1.99] | 1.458 [1.07, 2.00] |
| Age (scaled) | NA | 1.010 [0.14, 7.71] |

. . .

**They agree.** The Bayesian model adds:

- Posterior distributions (not just point estimates)
- Credible intervals (direct probability, not repeated sampling)
- Natural hierarchical structure (ISS random intercepts)

# Quality & Reproducibility {.inverse}

## Every Output is a Target

```{mermaid}
%%| fig-width: 10
flowchart LR
    SRC[R/viz/*.R<br/>source code] --> CODE[code_vig_*<br/>deparse body]
    SRC --> VIG[vig_*<br/>plot/table]
    CODE --> |"&lt;details&gt;"| HTML[Vignette HTML]
    VIG --> |"&lt;img&gt;"| HTML
    VIG --> RDS[RDS file<br/>for CI]
    RDS --> CI[pkgdown build<br/>Ubuntu]
    CI --> WEB[Live site]
    style SRC fill:#2ecc71,color:white
    style WEB fill:#9b59b6,color:white
```

## Validation Pipeline

| Check | Count | Status |
|-------|------:|--------|
| Pipeline targets | 250 | ✅ 0 errored |
| Unit tests | 892 | ✅ 0 failed |
| R CMD check | — | ✅ 0/0/0 |
| Chunk-target mapping | 165 chunks | ✅ 0 violations |
| Nix path check | 156 RDS | ✅ 0 violations |
| CI deploys | last 5 | ✅ all green |

# What's Missing {.inverse}

## 15 NULL Targets

| Missing Data | Source | Targets Affected |
|-------------|--------|:----------------:|
| FISH markers (all NA) | MMRF Gateway | 9 |
| PFS endpoints | MMRF Gateway | 7 |
| Treatment response | MMRF Gateway | 2 |
| Biomarker columns | GDC (in molecular_tests) | 1 |

. . .

**All require MMRF Researcher Gateway access** — registration submitted, pending approval.

## The Architecture Handles It

```{r null-example}
#| results: asis
cat('<div style="padding:1em;border-left:4px solid #f39c12;background:#2d2d2d;color:#ccc;border-radius:4px;font-size:0.9em;">')
cat('<strong>Data not available:</strong> Target <code>vig_km_risk</code> returned NULL.')
cat(' See <a href="https://johngavin.github.io/coMMpass-analysis/articles/data-sources.html">Data Sources</a>.')
cat('</div>')
```

. . .

Targets return `NULL` → vignettes show informative callout → no broken plots.

# Summary {.inverse}

## Key Numbers

| Metric | Value |
|--------|-------|
| Patients | 994 (of 1,143 in study) |
| Targets | 250 (0 errored) |
| Tests | 892 (0 failed) |
| Vignettes | 14 articles |
| DE genes | 1,615 (padj<0.05) |
| Cox C-index | 0.687 |
| Build time | ~7 minutes |
| RDS files | 156 (3.4 MB) |

## Links

| | |
|---|---|
| **Website** | [johngavin.github.io/coMMpass-analysis](https://johngavin.github.io/coMMpass-analysis/) |
| **GitHub** | [github.com/JohnGavin/coMMpass-analysis](https://github.com/JohnGavin/coMMpass-analysis) |
| **GDC Portal** | [portal.gdc.cancer.gov/projects/MMRF-COMMPASS](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS) |
| **CoMMpass Study** | [NCT01454297](https://clinicaltrials.gov/study/NCT01454297) |
| **Cachix** | [johngavin.cachix.org](https://johngavin.cachix.org) |

## References

- Keats JJ et al. *Blood* 2013 — CoMMpass interim analysis
- Hussain Z et al. *npj Digital Medicine* 2024 — SCOPE clinical normalisation
- Textor J et al. *IJE* 2016 — dagitty causal DAGs
- Hernán MA, Robins JM. *Causal Inference: What If* 2020
