---
title: "CoMMpass"
subtitle: "Inputs, outputs, and what the data tells us"
author: "John Gavin"
format:
  revealjs:
    theme: [dark, custom.scss]
    slide-number: true
    width: 1280
    height: 720
    embed-resources: true
    code-overflow: wrap
    fig-align: center
    transition: fade
    footer: "[johngavin.github.io/coMMpass-analysis](https://johngavin.github.io/coMMpass-analysis/)"
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(targets)
library(ggplot2)
tar_store <- if (dir.exists("../_targets")) "../_targets" else if (dir.exists("_targets")) "_targets" else NULL
safe_read <- function(name) {
  if (!is.null(tar_store)) {
    val <- tryCatch(targets::tar_read_raw(name, store = tar_store), error = function(e) NULL)
    if (!is.null(val)) return(val)
  }
  for (p in c(paste0("../inst/extdata/vignettes/", name, ".rds"),
              paste0("inst/extdata/vignettes/", name, ".rds"))) {
    if (file.exists(p)) return(readRDS(p))
  }
  NULL
}
# Shared dark theme for all plots
theme_preso <- function(base_size = 18) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.background = element_rect(fill = "#1a1a2e", color = NA),
      panel.background = element_rect(fill = "#1a1a2e", color = NA),
      panel.grid = element_line(color = "#2d2d44"),
      text = element_text(color = "white"),
      axis.text = element_text(color = "#b0b0b0", size = rel(0.9)),
      axis.title = element_text(color = "#d0d0d0"),
      plot.title = element_text(size = rel(1.1), face = "bold"),
      legend.background = element_rect(fill = "#1a1a2e", color = NA),
      legend.text = element_text(color = "white", size = rel(0.85)),
      legend.position = "bottom"
    )
}
# Render grob as ggplot
show_grob <- function(val) {
  if (is.null(val) || !inherits(val, c("gtable", "grob", "gTree"))) return(invisible(NULL))
  p <- ggplot() + annotation_custom(val) + theme_void() +
    theme(plot.background = element_rect(fill = "#1a1a2e", color = NA))
  print(p)
}
```

## The <abbr title="Clinical Outcomes in Multiple Myeloma to Personal Assessment of Genetic Profile">CoMMpass</abbr> study

- Cancer of plasma cells in bone marrow
- ~35,000 new <abbr title="United States">US</abbr> cases per year; median age 69
- <abbr title="CoMMpass: Clinical Outcomes in Multiple Myeloma to Personal Assessment of Genetic Profile">CoMMpass</abbr> enrolled **1,143 patients** (2011–2019) with 8-year follow-up
- Largest longitudinal genomics study of multiple myeloma

| | |
|---|---|
| Patients | 1,143 newly diagnosed |
| <abbr title="RNA sequencing">RNA-seq</abbr> samples | 859 (<abbr title="Spliced Transcripts Alignment to a Reference">STAR</abbr> counts) |
| Clinical variables | 88 per patient |
| Treatment records | 7,184 (from <abbr title="Genomic Data Commons">GDC</abbr> <abbr title="Application Programming Interface">API</abbr>) |

## Data we downloaded

| Dataset | Records | Size | Source |
|---------|--------:|-----:|--------|
| <abbr title="RNA sequencing">RNA-seq</abbr> (<abbr title="Spliced Transcripts Alignment to a Reference">STAR</abbr> counts) | 100 samples | 60.5 <abbr title="Megabytes">MB</abbr> | <abbr title="Genomic Data Commons">GDC</abbr> open access |
| Clinical metadata | 995 patients | 1.2 <abbr title="Megabytes">MB</abbr> | <abbr title="Genomic Data Commons">GDC</abbr> open access |
| Treatment records | 7,184 lines | 34 <abbr title="Kilobytes">KB</abbr> | <abbr title="Genomic Data Commons">GDC</abbr> <abbr title="Application Programming Interface">API</abbr> |
| <abbr title="Molecular Signatures Database">MSigDB</abbr> gene sets | 708 sets | 85 <abbr title="Kilobytes">KB</abbr> | Bundled in package |

- <abbr title="Fluorescence In Situ Hybridisation">FISH</abbr> markers, <abbr title="Progression-Free Survival">PFS</abbr> endpoints, and treatment response are **not available** from <abbr title="Genomic Data Commons">GDC</abbr>
- These require <abbr title="Multiple Myeloma Research Foundation">MMRF</abbr> Researcher Gateway access (pending)

## Pipeline overview

| | |
|---|---|
| Total targets | 250 |
| Build time | ~7 minutes |
| Errored | 0 |
| <abbr title="Target returned NULL due to missing data">NULL</abbr> (data gaps) | 15 |
| <abbr title="Serialised R object files">RDS</abbr> exports | 156 files, 3.4 <abbr title="Megabytes">MB</abbr> |
| Vignettes | 14 articles |
| Unit tests | 892 (0 failed) |

```{mermaid}
%%| fig-width: 10
flowchart LR
    GDC["GDC API"] --> SE["RNA-seq<br>+ clinical"]
    SE --> DE["1,615 DE genes"]
    SE --> KM["Survival curves"]
    SE --> DAG["Causal DAG"]
    DE --> GSEA["Pathway<br>enrichment"]
    KM --> COX["Cox model<br>C = 0.687"]
    style GDC fill:#2ecc71,stroke:#2ecc71,color:white
    style DE fill:#e74c3c,stroke:#e74c3c,color:white
    style KM fill:#3498db,stroke:#3498db,color:white
    style COX fill:#3498db,stroke:#3498db,color:white
    style GSEA fill:#e74c3c,stroke:#e74c3c,color:white
    style DAG fill:#9b59b6,stroke:#9b59b6,color:white
    style SE fill:#2ecc71,stroke:#2ecc71,color:white
```

## Survival by <abbr title="International Staging System: I = low tumour burden, II = intermediate, III = high">ISS</abbr> stage

```{r km-iss}
#| fig-height: 5
show_grob(safe_read("vig_km_iss"))
```

- <abbr title="International Staging System">ISS</abbr> III patients have ~5× the hazard of <abbr title="International Staging System">ISS</abbr> I (<abbr title="Hazard Ratio: risk of death relative to reference group">HR</abbr> = 4.9, p < 0.001)
- 994 patients from <abbr title="Genomic Data Commons">GDC</abbr> clinical data; condition derived from vital status (alive vs dead)

## <abbr title="Cox Proportional Hazards: a regression model for time-to-event data">Cox</abbr> regression: what predicts survival?

| Covariate | <abbr title="Hazard Ratio: >1 means higher risk of death">HR</abbr> | 95% <abbr title="Confidence Interval">CI</abbr> | p-value |
|-----------|---:|--------|------:|
| <abbr title="International Staging System">ISS</abbr> stage II | 2.35 | [1.51, 3.67] | <0.001 |
| <abbr title="International Staging System">ISS</abbr> stage III | 4.90 | [3.21, 7.48] | <0.001 |
| Male sex | 1.45 | [1.07, 1.97] | 0.018 |

- <abbr title="Concordance index: 0.5 = coin flip, 1.0 = perfect discrimination">C-index</abbr> = 0.687 — moderate discrimination
- Age not significant after adjusting for <abbr title="International Staging System">ISS</abbr>
- Model uses clinical data only (no genomics)

## Differential expression: alive vs dead

```{r volcano}
#| fig-height: 5
show_grob(safe_read("vig_volcano_plot"))
```

- **1,615 genes** significantly changed between alive and dead patients
- Thresholds: |log₂ fold change| > 1 and adjusted p-value < 0.05
- <abbr title="Differential Expression analysis using DESeq2, a Bioconductor package">DESeq2</abbr> with <abbr title="Variance Stabilising Transformation: normalises RNA-seq counts for visualisation">VST</abbr>-normalised counts

## What "1,615 genes" means in practice

- A **typical** differentially expressed gene has log₂FC ≈ 0.8
  - That is a 1.7× expression difference — subtle but real
  - Example: gene X has 1,000 counts in alive patients, 1,700 in dead
- An **extreme** gene might show log₂FC = 8
  - That is a 256× difference — almost certainly an artefact
  - Likely caused by a single outlier sample or a low-count gene
- The adjusted p-value (0.05 threshold) controls false discoveries
  - At 0.05: expect ~80 false positives among 1,615 (5%)
  - The volcano plot's y-axis shows how confident each call is

## Pathway enrichment

```{r gsea}
#| fig-height: 5
show_grob(safe_read("vig_gsea_dotplot"))
```

- <abbr title="Gene Set Enrichment Analysis: tests whether predefined gene sets are enriched at the top/bottom of a ranked gene list">GSEA</abbr> of 50 <abbr title="Molecular Signatures Database: curated gene set collections from the Broad Institute">MSigDB</abbr> Hallmark pathways
- Enriched pathways indicate biological processes disrupted in patients who died

## Causal assumptions (<abbr title="Directed Acyclic Graph: a diagram encoding causal assumptions between variables">DAG</abbr>)

```{r dag}
#| fig-height: 5
show_grob(safe_read("vig_causal_dag_plot"))
```

- Cytogenetic risk → survival is **not identifiable** (unmeasured tumour biology confounds it)
- <abbr title="International Staging System">ISS</abbr> → survival IS identifiable if we adjust for age, cytogenetics, and expression

## Missingness predicts outcomes

| Variable | What "missing" means | <abbr title="Hazard Ratio">HR</abbr> | p-value |
|----------|---------------------|---:|--------:|
| Cause of death | Patient is alive | ≈ 0 | 2 × 10⁻²⁶ |

- Patients without a recorded cause of death are alive — the missingness IS the outcome
- This is <abbr title="Missing Not At Random: the probability of being missing depends on the unobserved value itself">MNAR</abbr> — the missing value is informative, not noise
- Obvious in hindsight, but the statistical test confirms the mechanism

## Bayesian vs frequentist: same answer

| Covariate | Frequentist <abbr title="Hazard Ratio">HR</abbr> | Bayesian <abbr title="Hazard Ratio">HR</abbr> |
|-----------|---:|---:|
| Male sex | 1.47 [1.08, 1.99] | 1.46 [1.07, 2.00] |

- Both frameworks agree on the effect of sex on survival
- Bayesian model fitted with <abbr title="brms: Bayesian Regression Models using Stan">brms</abbr> (weakly informative priors)
- The Bayesian approach adds posterior distributions and natural hierarchical structure

## Summary

| What we measured | Result |
|------------------|--------|
| Patients analysed | 994 (of 1,143 enrolled) |
| <abbr title="Differential Expression">DE</abbr> genes | 1,615 (adjusted p < 0.05) |
| <abbr title="International Staging System">ISS</abbr> III hazard ratio | 4.9× vs <abbr title="International Staging System">ISS</abbr> I |
| <abbr title="Cox Proportional Hazards model">Cox</abbr> C-index | 0.687 |
| Pipeline targets | 250 (0 errored) |
| Build time | ~7 minutes |

- **Website**: [johngavin.github.io/coMMpass-analysis](https://johngavin.github.io/coMMpass-analysis/)
- **Code**: [github.com/JohnGavin/coMMpass-analysis](https://github.com/JohnGavin/coMMpass-analysis)
- **Data**: [GDC portal](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS) (open access)
- **Study**: [NCT01454297](https://clinicaltrials.gov/study/NCT01454297)
