---
title: "CoMMpass"
subtitle: "Inputs, outputs, and what the data tells us"
author: "John Gavin"
format:
  revealjs:
    theme: [dark, custom.scss]
    slide-number: true
    width: 1280
    height: 720
    embed-resources: true
    fig-align: center
    transition: fade
    footer: "[johngavin.github.io/coMMpass-analysis](https://johngavin.github.io/coMMpass-analysis/)"
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(ggplot2)
library(targets)
tar_store <- if (dir.exists("../_targets")) "../_targets" else if (dir.exists("_targets")) "_targets" else NULL
safe_read <- function(name) {
  if (!is.null(tar_store)) {
    val <- tryCatch(targets::tar_read_raw(name, store = tar_store), error = function(e) NULL)
    if (!is.null(val)) return(val)
  }
  for (p in c(paste0("../inst/extdata/vignettes/", name, ".rds"),
              paste0("inst/extdata/vignettes/", name, ".rds"))) {
    if (file.exists(p)) return(readRDS(p))
  }
  NULL
}
theme_preso <- function(base_size = 22) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.background = element_rect(fill = "black", color = NA),
      panel.background = element_rect(fill = "black", color = NA),
      panel.grid.major = element_line(color = "#333333"),
      panel.grid.minor = element_blank(),
      text = element_text(color = "white"),
      axis.text = element_text(color = "#cccccc", size = rel(1)),
      axis.title = element_text(color = "white", size = rel(1.1)),
      axis.line = element_line(color = "white"),
      plot.title = element_blank(),
      plot.caption = element_blank(),
      legend.background = element_rect(fill = "black", color = NA),
      legend.text = element_text(color = "white", size = rel(0.9)),
      legend.title = element_text(color = "white"),
      legend.position = "bottom",
      strip.text = element_text(color = "white", size = rel(1), face = "bold"),
      strip.background = element_rect(fill = "#222222", color = NA)
    )
}
show_grob <- function(val) {
  if (is.null(val) || !inherits(val, c("gtable", "grob", "gTree"))) return(invisible(NULL))
  p <- ggplot() + annotation_custom(val) + theme_void() +
    theme(plot.background = element_rect(fill = "black", color = NA))
  print(p)
}
```

## The patients

- <abbr title="CoMMpass: Clinical Outcomes in Multiple Myeloma to Personal Assessment of Genetic Profile. A longitudinal study tracking 1,143 myeloma patients over 8 years. Example: patient MMRF_1618 enrolled in 2012, followed until 2019.">CoMMpass</abbr> enrolled **1,143 newly diagnosed <abbr title="Multiple myeloma: a cancer of plasma cells in the bone marrow. Plasma cells normally make antibodies. In myeloma, they grow uncontrollably. Example: a patient's bone marrow biopsy shows 60% plasma cells instead of the normal 2-3%.">multiple myeloma</abbr> patients** (2011–2019)
- Median age at diagnosis: **69 years** (range: 27–93)
- 8-year follow-up with <abbr title="RNA sequencing: measures which genes are active and how active they are, by counting RNA molecules. Example: gene TP53 has 500 RNA copies in sample A but only 50 in sample B — it is 10× more active in A.">RNA-seq</abbr>, clinical data, and treatment records

| What was measured | Example from the data |
|-------------------|----------------------|
| Age at diagnosis | Patient MMRF_2240: 71 years |
| <abbr title="International Staging System: classifies myeloma severity using two blood tests (albumin and beta-2 microglobulin). Stage I = low tumour burden, Stage III = high tumour burden. Example: albumin 3.0 g/dL + B2M 8.0 mg/L = Stage III.">ISS</abbr> stage | 348 patients Stage I, 353 Stage II, 266 Stage III |
| Vital status | 803 alive, 191 dead at last follow-up |
| Genes measured | 60,660 genes per patient |
| Treatment drugs | Dexamethasone (1,760 records), Lenalidomide (1,393), Bortezomib (1,076) |

## The data we have

- **<abbr title="RNA sequencing counts: the number of RNA fragments detected for each gene in each sample. Higher count = gene is more active. Example: gene ACTB has 50,000 counts (very active housekeeping gene) while gene FOXP3 has 12 counts (rarely active).">RNA-seq</abbr>** — 100 patient samples, 60,660 genes each
  - Each sample is a snapshot of which genes are turned on or off
  - Size: 60.5 <abbr title="Megabytes: a unit of file size. 1 MB ≈ 500 pages of text. Example: the RNA-seq data is 60 MB, roughly equivalent to 30,000 pages.">MB</abbr> from <abbr title="Genomic Data Commons: the US National Cancer Institute's data portal for cancer genomics. Open access for most clinical and expression data. Example: visit portal.gdc.cancer.gov to browse 995 CoMMpass cases.">GDC</abbr> open access
- **Clinical metadata** — 995 patients, 88 variables each
  - Demographics (age, sex, race), disease staging (<abbr title="International Staging System: classifies myeloma severity. Stage I = low tumour burden, Stage III = high. Example: ISS III patients have 5× higher death rate than Stage I.">ISS</abbr>), outcomes (alive/dead)
  - Size: 1.2 <abbr title="Megabytes">MB</abbr>
- **Treatment records** — 7,184 drug administration records for 994 patients
  - Which drugs, which line of therapy, timing
  - Example: Patient MMRF_2240 received Dexamethasone + Cyclophosphamide as first-line, then switched to Dexamethasone alone as second-line
- **<abbr title="Molecular Signatures Database: curated collections of known gene groups. Hallmark = 50 well-defined biological processes. Example: 'HALLMARK_APOPTOSIS' contains 161 genes involved in programmed cell death.">MSigDB</abbr> gene sets** — 708 predefined gene groups (bundled in package, not downloaded)

## The data we don't have

| Missing data | Why it matters | Why it's missing |
|-------------|----------------|-----------------|
| <abbr title="Fluorescence In Situ Hybridisation: a lab test that detects specific chromosome changes. Example: t(4;14) means chromosomes 4 and 14 swapped pieces — this happens in ~15% of myeloma patients and predicts worse outcomes.">FISH</abbr> markers | Classify patients into high-risk vs standard-risk | Not in <abbr title="Genomic Data Commons">GDC</abbr>; requires <abbr title="Multiple Myeloma Research Foundation: the organisation that runs the CoMMpass study and controls access to detailed clinical data.">MMRF</abbr> Gateway |
| <abbr title="Progression-Free Survival: time from treatment start until the disease gets worse or the patient dies. Example: a PFS of 24 months means the treatment kept the cancer controlled for 2 years.">PFS</abbr> endpoints | Measure treatment effectiveness | Not in <abbr title="Genomic Data Commons">GDC</abbr>; requires <abbr title="Multiple Myeloma Research Foundation">MMRF</abbr> Gateway |
| Treatment response | Did the drug work? (complete response, partial, etc.) | <abbr title="Genomic Data Commons">GDC</abbr> field is all empty |
| Biomarker lab values | Blood test results (creatinine, LDH, etc.) | Stored in nested <abbr title="Genomic Data Commons">GDC</abbr> fields, not flat columns |

- 15 of 250 pipeline targets return empty because of these gaps
- Registration submitted to <abbr title="Multiple Myeloma Research Foundation">MMRF</abbr> Researcher Gateway

## Who survives longer? Staging matters

```{r km-iss}
#| fig-height: 4.5
show_grob(safe_read("vig_km_iss"))
```

- <abbr title="International Staging System: Stage I (albumin ≥ 3.5, B2M < 3.5) = low tumour burden. Stage III (B2M ≥ 5.5) = high tumour burden. Example: Stage I patient typically survives 5+ years; Stage III median ~2-3 years.">ISS</abbr> Stage III patients die at **5× the rate** of Stage I
- The three curves separate early and stay separated — staging predicts long-term outcome
- 994 patients; alive vs dead comparison derived from vital status

## What predicts survival?

- A <abbr title="Cox Proportional Hazards model: a statistical method that estimates how much each factor (age, sex, staging) increases or decreases the risk of death. Example: HR = 2.0 means twice the risk of death compared to the reference group.">Cox regression model</abbr> adjusting for age, sex, and <abbr title="International Staging System">ISS</abbr> stage:

| Factor | <abbr title="Hazard Ratio: the relative risk of death. HR = 1 means no effect. HR = 2 means twice the risk. HR = 0.5 means half the risk. Example: HR = 4.9 for ISS III means these patients die at nearly 5× the rate of ISS I.">HR</abbr> | Meaning |
|--------|---:|---------|
| <abbr title="International Staging System">ISS</abbr> II vs I | 2.4 | 2.4× higher death rate |
| <abbr title="International Staging System">ISS</abbr> III vs I | 4.9 | Nearly 5× higher death rate |
| Male vs female | 1.5 | 50% higher death rate |
| Age | Not significant | After adjusting for <abbr title="International Staging System">ISS</abbr>, age adds little |

- <abbr title="Concordance index: measures how well the model discriminates who dies first. C = 0.5 is random guessing. C = 1.0 is perfect. C = 0.687 means the model correctly ranks 69% of patient pairs. Example: given two patients, the model correctly predicts which dies first 69% of the time.">C-index</abbr> = 0.687 — the model correctly ranks patients 69% of the time

## Which genes differ between alive and dead?

```{r volcano}
#| fig-height: 4.5
show_grob(safe_read("vig_volcano_plot"))
```

- **1,615 genes** significantly differ between alive and dead patients
- <abbr title="Volcano plot: each dot is one gene. X-axis = how much the gene changes (fold change). Y-axis = how confident we are (statistical significance). Genes in the top-left and top-right corners are the most interesting — big change AND high confidence. Example: a gene at x=2, y=10 has 4× expression change with very high confidence.">Volcano plot</abbr>: x-axis = expression change, y-axis = confidence
- Thresholds: |<abbr title="log₂ Fold Change: the log-base-2 of the ratio of expression between groups. log₂FC = 1 means 2× change. log₂FC = 2 means 4× change. Why log₂? Because it makes increases and decreases symmetric: +1 = doubled, -1 = halved. Example: gene with 1000 counts in alive vs 2000 in dead has log₂FC = 1.">log₂FC</abbr>| > 1 (at least 2× change) and adjusted p-value < 0.05

## What gene changes mean in practice

- **Typical**: <abbr title="log₂ Fold Change">log₂FC</abbr> ≈ 0.8
  - A 1.7× expression difference — subtle but real
  - Example: gene X has 1,000 <abbr title="RNA sequencing">RNA</abbr> counts in alive patients, 1,700 in dead
  - This is the kind of change that adds up across many genes
- **Extreme**: <abbr title="log₂ Fold Change">log₂FC</abbr> = 8.0
  - A 256× difference — almost certainly an artefact
  - Usually caused by a gene with very low counts in one group
  - Example: 1 count in alive, 256 in dead — statistically big but biologically noisy
- **The 1,615 genes** at our thresholds:
  - Expect ~80 false positives (5% of 1,615)
  - The rest are real differences worth investigating
  - Median <abbr title="log₂ Fold Change">log₂FC</abbr> among significant genes: ~1.2 (2.3× change)

## Which biological processes are disrupted?

- <abbr title="Gene Set Enrichment Analysis: instead of looking at one gene at a time, GSEA asks 'are genes in this known group (e.g. cell death genes) collectively shifted?' Example: if 100 out of 161 apoptosis genes are slightly upregulated, GSEA detects this even though no single gene is dramatic.">GSEA</abbr> tests whether predefined gene groups are collectively shifted
- We tested 50 <abbr title="Hallmark gene sets: a curated collection of 50 biological processes from MSigDB, each representing a specific, well-defined state or function. They are more specific than broad categories like 'cancer genes'. Example: HALLMARK_HYPOXIA (200 genes activated when oxygen is low) vs HALLMARK_APOPTOSIS (161 genes involved in cell death). Non-Hallmark sets include thousands of less-curated groups.">Hallmark</abbr> pathways from <abbr title="Molecular Signatures Database">MSigDB</abbr>

```{r gsea}
#| fig-height: 3.5
show_grob(safe_read("vig_gsea_dotplot"))
```

- Each dot is one pathway; position shows enrichment strength
- Enriched pathways = biological processes more active in dead patients

## Can we prove causation?

```{r dag}
#| fig-height: 4
show_grob(safe_read("vig_causal_dag_plot"))
```

- A <abbr title="Directed Acyclic Graph: a diagram that maps assumed cause-and-effect relationships. Arrows show 'A causes B'. If there is a hidden common cause of both A and B, we cannot prove A causes B even if they are correlated. Example: ice cream sales and drowning deaths are correlated — but the common cause is hot weather, not ice cream.">DAG</abbr> maps our assumptions about what causes what
- Chromosome changes (<abbr title="Cytogenetics: the study of chromosome structure. In myeloma, specific chromosome rearrangements (like swapping pieces between chromosomes 4 and 14) predict worse outcomes.">cytogenetics</abbr>) → survival: **cannot be proven** because unmeasured tumour biology confounds it
- <abbr title="International Staging System">ISS</abbr> → survival: **can be estimated** if we adjust for age, chromosome changes, and <abbr title="Gene expression: how active each gene is, measured as RNA counts. Example: a gene with high expression (10,000 counts) is very active; one with low expression (10 counts) is nearly silent.">gene expression</abbr>

## Missing data tells a story

| Variable | What "missing" means | <abbr title="Hazard Ratio">HR</abbr> | p-value |
|----------|---------------------|---:|--------:|
| Cause of death | Patient is **alive** | ≈ 0 | 2 × 10⁻²⁶ |

- Patients without a recorded cause of death are alive — the gap in the data IS the answer
- This is <abbr title="Missing Not At Random: the reason the data is missing is related to the missing value itself. Example: cause of death is missing BECAUSE the patient is alive. FISH test is missing BECAUSE the doctor decided the patient was low-risk and didn't order it. Contrast with MCAR (Missing Completely At Random) where missingness is unrelated to the value.">MNAR</abbr> — the missing value is informative, not noise
- Obvious in hindsight, but the statistical test (p < 10⁻²⁶) confirms it
- In myeloma, <abbr title="Fluorescence In Situ Hybridisation">FISH</abbr> testing is expensive — if missingness correlates with better survival, doctors may skip it for low-risk patients

## Summary

| What we measured | Result |
|------------------|--------|
| Patients analysed | 994 of 1,143 enrolled |
| Genes tested | 60,660 |
| Significant genes (alive vs dead) | 1,615 |
| <abbr title="International Staging System">ISS</abbr> III vs I hazard ratio | 4.9× |
| Model discrimination (<abbr title="Concordance index">C-index</abbr>) | 0.687 |
| Pipeline targets | 250 (0 errors) |

- **Website**: [johngavin.github.io/coMMpass-analysis](https://johngavin.github.io/coMMpass-analysis/)
- **Code**: [github.com/JohnGavin/coMMpass-analysis](https://github.com/JohnGavin/coMMpass-analysis)
- **Data**: [GDC portal](https://portal.gdc.cancer.gov/projects/MMRF-COMMPASS) (open access)
- **Study registration**: [NCT01454297](https://clinicaltrials.gov/study/NCT01454297)
