---
title: "05. Survival Analysis"
format: html
vignette: >
  %\VignetteIndexEntry{05. Survival Analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
library(targets)

# Discover targets store
proj_dir <- Sys.getenv("TAR_PROJECT")
tar_store <- NULL
if (proj_dir != "" && dir.exists(file.path(proj_dir, "_targets"))) {
  tar_store <- file.path(proj_dir, "_targets")
} else if (dir.exists("_targets")) {
  tar_store <- "_targets"
} else if (dir.exists("../_targets")) {
  tar_store <- "../_targets"
}

safe_tar_read <- function(name) {
  if (is.null(tar_store)) return(NULL)
  tryCatch(tar_read_raw(name, store = tar_store), error = function(e) NULL)
}
```

## Overview

> See the [Glossary](glossary.html) for term definitions used throughout this project.

- Survival analysis of the CoMMpass cohort stratified by clinical and cytogenetic features
- **Kaplan-Meier curves** with log-rank tests for group comparisons
- **Cox proportional hazards models** for multivariate analysis
- **Forest plots** of hazard ratios with 95% confidence intervals

> **Note:** This vignette was built in CI with `sample_limit=10`. Local builds
> default to 100 samples. Numbers below reflect the CI subset.

## Overall Survival

```{r km-overall, fig.cap = "Overall Kaplan-Meier survival curve for all CoMMpass patients with valid survival data. Shaded region: 95% confidence interval. Time in days from diagnosis."}
safe_tar_read("vig_km_overall")
```

```{r km-overall-text, results='asis'}
safe_tar_read("vig_km_overall_text")
```

## Survival by ISS Stage

- [ISS](https://doi.org/10.1200/JCO.2005.04.242) stratifies by serum albumin and beta-2 microglobulin
- ISS III is expected to have worse survival than ISS I
- See the [EDA vignette](exploratory-analysis.html) for ISS stage distributions

```{r km-iss, fig.cap = "Kaplan-Meier curves by ISS stage (I, II, III). Log-rank p-value tests whether survival distributions differ across stages. ISS III patients are expected to have shorter survival. See Greipp et al. 2005 for ISS definition."}
safe_tar_read("vig_km_iss")
```

## Survival by Cytogenetic Risk Group

Cytogenetic risk per [IMWG 2014 criteria](https://doi.org/10.1200/JCO.2014.55.1519):
**high-risk** = t(4;14) OR t(14;16) OR del(17p); **standard-risk** = all others.

```{r km-risk, fig.cap = "Kaplan-Meier curves by IMWG cytogenetic risk group (high vs standard). High-risk: t(4;14), t(14;16), or del(17p). Standard-risk: all others. Log-rank p-value shown."}
safe_tar_read("vig_km_risk")
```

## Survival by Individual Cytogenetic Markers

Each marker is analyzed separately: patients with vs without the alteration.

```{r km-markers, fig.height = 5, fig.cap = "Kaplan-Meier curves for individual cytogenetic markers. Each panel compares patients with (positive) vs without (negative) the alteration. Markers with fewer than 3 positive or negative patients are skipped."}
safe_tar_read("vig_km_markers")
```

## Cox Proportional Hazards

### Basic Model: Age + Gender

```{r cox-basic}
safe_tar_read("vig_cox_basic_table")
```

### Full Model: Age + Gender + ISS + Cytogenetic Risk

```{r cox-full}
safe_tar_read("vig_cox_full_table")
```

### Forest Plot

```{r forest-plot, fig.cap = "Forest plot of hazard ratios from multivariate Cox regression. Points: HR estimates. Horizontal bars: 95% CI. Dashed line: HR = 1 (no effect). Log scale x-axis. Stars: significance level."}
safe_tar_read("vig_forest_plot")
```

### Proportional Hazards Assumption

```{r ph-test}
safe_tar_read("vig_ph_test_table")
```

## Model Comparison

```{r model-comparison}
safe_tar_read("vig_cox_comparison")
```

## Survival by Gene Expression

Patients are split at the median VST expression of each top DE gene into
"High" and "Low" groups. This connects differential expression results to
clinical outcomes.

> **Note on multiple testing:** With 5 genes tested, a Bonferroni-corrected
> significance threshold is p < 0.01.

```{r km-expression, fig.height = 5, fig.cap = "Kaplan-Meier curves stratified by median-split gene expression. High = above median VST, Low = below median. Log-rank p-value shown. Top DE genes from consensus analysis."}
safe_tar_read("vig_km_by_expression")
```

## Expression by Cytogenetic Subtype

Violin/box plots showing gene expression (VST) stratified by cytogenetic
marker status. This reveals whether specific alterations drive expression
changes in top DE genes.

```{r expr-by-subtype, fig.height = 6, fig.cap = "Gene expression by cytogenetic subtype for top DE genes. Violin + box plots, positive vs negative marker status. Wilcoxon p-value annotated."}
safe_tar_read("vig_expr_by_subtype")
```

## Next Steps

- **Time-varying coefficients**: For covariates violating the PH assumption.
- **Cure models**: If a plateau is observed in KM curves, consider
  mixture cure models.
- See the [cytogenetic landscape](exploratory-analysis.html) for alteration
  frequencies underlying risk groups.
- See the [DE results](differential-expression.html) for gene signatures
  that could inform survival stratification.
