---
title: "05. Survival Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{05. Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)

safe_tar_read <- function(name) {
  tryCatch(
    targets::tar_read(name = name, store = targets::tar_config_get("store")),
    error = function(e) NULL
  )
}
```

## Overview

> See the [Glossary](glossary.html) for term definitions used throughout this project.

- Survival analysis of the CoMMpass cohort stratified by clinical and cytogenetic features
- **Kaplan-Meier curves** with log-rank tests for group comparisons
- **Cox proportional hazards models** for multivariate analysis
- **Forest plots** of hazard ratios with 95% confidence intervals

> **Note:** This vignette was built in CI with `sample_limit=10`. Local builds
> default to 100 samples. Numbers below reflect the CI subset.

## Overall Survival

```{r km-overall, fig.cap = "Overall Kaplan-Meier survival curve for all CoMMpass patients with valid survival data. Shaded region: 95% confidence interval. Time in days from diagnosis."}
km_overall <- safe_tar_read("km_overall")

if (!is.null(km_overall) && !is.null(km_overall$fit)) {
  p <- plot_km(km_overall, title = "Overall Survival")
  print(p)

  median_os <- km_overall$median_survival
  if (!all(is.na(median_os))) {
    cat(paste0("\n**Median overall survival:** ",
               round(median_os[1]), " days (",
               round(median_os[1] / 365.25, 1), " years)\n"))
  }
} else {
  cat("*Overall KM not available. Run `tar_make()` to build survival targets.*\n")
}
```

## Survival by ISS Stage

- [ISS](https://doi.org/10.1200/JCO.2005.04.242) stratifies by serum albumin and beta-2 microglobulin
- ISS III is expected to have worse survival than ISS I
- See the [EDA vignette](exploratory-analysis.html) for ISS stage distributions

```{r km-iss, fig.cap = "Kaplan-Meier curves by ISS stage (I, II, III). Log-rank p-value tests whether survival distributions differ across stages. ISS III patients are expected to have shorter survival. See Greipp et al. 2005 for ISS definition."}
km_iss <- safe_tar_read("km_by_iss")

if (!is.null(km_iss) && !is.null(km_iss$fit)) {
  p <- plot_km(km_iss, title = "Survival by ISS Stage")
  print(p)
} else {
  note <- if (!is.null(km_iss$note)) km_iss$note else
    "ISS data not available"
  cat(paste0("*", note, ".*\n"))
}
```

## Survival by Cytogenetic Risk Group

Cytogenetic risk per [IMWG 2014 criteria](https://doi.org/10.1200/JCO.2014.55.1519):
**high-risk** = t(4;14) OR t(14;16) OR del(17p); **standard-risk** = all others.

```{r km-risk, fig.cap = "Kaplan-Meier curves by IMWG cytogenetic risk group (high vs standard). High-risk: t(4;14), t(14;16), or del(17p). Standard-risk: all others. Log-rank p-value shown."}
km_risk <- safe_tar_read("km_by_risk")

if (!is.null(km_risk) && !is.null(km_risk$fit)) {
  p <- plot_km(km_risk, title = "Survival by Cytogenetic Risk")
  print(p)
} else {
  note <- if (!is.null(km_risk$note)) km_risk$note else
    "Cytogenetic risk data not available"
  cat(paste0("*", note, ".*\n"))
}
```

## Survival by Individual Cytogenetic Markers

Each marker is analyzed separately: patients with vs without the alteration.

```{r km-markers, fig.height = 5, fig.cap = "Kaplan-Meier curves for individual cytogenetic markers. Each panel compares patients with (positive) vs without (negative) the alteration. Markers with fewer than 3 positive or negative patients are skipped."}
km_markers <- safe_tar_read("km_by_markers")

marker_labels <- c(
  t_4_14 = "t(4;14)", t_11_14 = "t(11;14)", t_14_16 = "t(14;16)",
  del_17p = "del(17p)", gain_1q = "gain(1q)"
)

if (!is.null(km_markers) && length(km_markers) > 0) {
  for (m in names(km_markers)) {
    km <- km_markers[[m]]
    label <- if (m %in% names(marker_labels)) marker_labels[m] else m
    if (!is.null(km$fit)) {
      p <- plot_km(km, title = paste0("Survival by ", label))
      print(p)
      cat("\n\n")
    } else if (!is.null(km$note)) {
      cat(paste0("**", label, "**: *", km$note, "*\n\n"))
    }
  }
} else {
  cat("*Individual marker KM data not available. Run `tar_make()` first.*\n")
}
```

## Cox Proportional Hazards

### Basic Model: Age + Gender

```{r cox-basic}
cox_basic <- safe_tar_read("cox_basic")

if (!is.null(cox_basic) && !is.null(cox_basic$hazard_ratios)) {
  knitr::kable(
    cox_basic$hazard_ratios, row.names = FALSE,
    digits = c(0, 3, 3, 3, 4),
    caption = paste0(
      "Cox PH model: age + gender. HR > 1 = increased hazard (worse survival). ",
      "n = ", cox_basic$n, " patients, ", cox_basic$n_events, " events. ",
      "C-index = ", round(cox_basic$concordance, 3), ". ",
      "* p < 0.05, ** p < 0.01, *** p < 0.001."
    )
  )
} else {
  cat("*Basic Cox model not available. Run `tar_make()` first.*\n")
}
```

### Full Model: Age + Gender + ISS + Cytogenetic Risk

```{r cox-full}
cox_full <- safe_tar_read("cox_full")

if (!is.null(cox_full) && !is.null(cox_full$hazard_ratios)) {
  knitr::kable(
    cox_full$hazard_ratios, row.names = FALSE,
    digits = c(0, 3, 3, 3, 4),
    caption = paste0(
      "Cox PH model with clinical and cytogenetic covariates. ",
      "n = ", cox_full$n, " patients, ", cox_full$n_events, " events. ",
      "C-index = ", round(cox_full$concordance, 3), ". ",
      "Covariates: ", paste(cox_full$covariates_used, collapse = ", "), "."
    )
  )
} else {
  cat("*Full Cox model not available. Run `tar_make()` first.*\n")
}
```

### Forest Plot

```{r forest-plot, fig.cap = "Forest plot of hazard ratios from multivariate Cox regression. Points: HR estimates. Horizontal bars: 95% CI. Dashed line: HR = 1 (no effect). Log scale x-axis. Stars: significance level."}
if (!is.null(cox_full) && !is.null(cox_full$hazard_ratios) &&
    nrow(cox_full$hazard_ratios) > 0) {
  p <- plot_forest(cox_full,
                    title = "Multivariate Cox: Hazard Ratios")
  print(p)
} else if (!is.null(cox_basic) && !is.null(cox_basic$hazard_ratios) &&
           nrow(cox_basic$hazard_ratios) > 0) {
  p <- plot_forest(cox_basic,
                    title = "Cox Regression: Hazard Ratios")
  print(p)
} else {
  cat("*Forest plot not available.*\n")
}
```

### Proportional Hazards Assumption

```{r ph-test}
cox_for_ph <- if (!is.null(cox_full)) cox_full else cox_basic

if (!is.null(cox_for_ph) && !is.null(cox_for_ph$ph_test)) {
  ph <- cox_for_ph$ph_test
  ph_table <- as.data.frame(ph$table)
  ph_table$variable <- rownames(ph_table)
  ph_table <- ph_table[, c("variable", "chisq", "p")]
  names(ph_table) <- c("Variable", "Chi-squared", "p-value")

  knitr::kable(
    ph_table, row.names = FALSE, digits = c(0, 2, 4),
    caption = paste0(
      "Proportional hazards assumption test (cox.zph). ",
      "p < 0.05 indicates the PH assumption may be violated for that covariate. ",
      "If violated, time-varying coefficients or stratification should be considered."
    )
  )
} else {
  cat("*PH test not available.*\n")
}
```

## Model Comparison

```{r model-comparison}
if (!is.null(cox_basic) && !is.null(cox_full) &&
    !is.null(cox_basic$concordance) && !is.null(cox_full$concordance)) {
  comp <- data.frame(
    Model = c("Basic (age + gender)",
              paste0("Full (", paste(cox_full$covariates_used, collapse = " + "), ")")),
    N = c(cox_basic$n, cox_full$n),
    Events = c(cox_basic$n_events, cox_full$n_events),
    C_index = c(round(cox_basic$concordance, 3), round(cox_full$concordance, 3)),
    stringsAsFactors = FALSE
  )
  knitr::kable(
    comp, row.names = FALSE,
    caption = paste0(
      "Cox model comparison. C-index: concordance statistic (0.5 = random, ",
      "1.0 = perfect discrimination). Higher C-index indicates better ",
      "prognostic discrimination."
    ),
    col.names = c("Model", "N", "Events", "C-index")
  )
} else {
  cat("*Model comparison not available.*\n")
}
```

## Next Steps

- **Gene expression stratification**: Median-split KM by top DE gene
  expression (requires DE and VST targets).
- **Time-varying coefficients**: For covariates violating the PH assumption.
- **Cure models**: If a plateau is observed in KM curves, consider
  mixture cure models.
- See the [cytogenetic landscape](exploratory-analysis.html) for alteration
  frequencies underlying risk groups.
- See the [DE results](differential-expression.html) for gene signatures
  that could inform survival stratification.
