---
title: "Pipeline Telemetry"
format: html
vignette: >
  %\VignetteIndexEntry{Pipeline Telemetry}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
library(targets)

# Discover target store
store_candidates <- c(
  here::here("_targets"),
  file.path(Sys.getenv("TARGETS_STORE", unset = "_targets")),
  "_targets"
)
for (store in store_candidates) {
  if (dir.exists(store)) {
    Sys.setenv(TAR_STORE = store)
    break
  }
}
```

Pipeline build telemetry from `targets::tar_meta()`. These are metadata
introspection calls (allowed exception to the zero-computation rule).

## Build Timing

Targets ordered by build duration. Long-running targets are candidates
for caching optimization or parallel execution via `crew`.

```{r timing-table}
#| fig.cap: "Target build times. Seconds: wall-clock time per target.
#|   Bytes: serialized object size. Top targets by duration are data
#|   downloads (raw_rnaseq, clinical_data) and DESeq2 paired analysis.
#|   Source: targets::tar_meta()."
meta <- targets::tar_meta(store = Sys.getenv("TAR_STORE", "_targets"))
meta_timed <- meta[!is.na(meta$seconds), c("name", "seconds", "bytes")]
meta_timed <- meta_timed[order(-meta_timed$seconds), ]
meta_timed$bytes_fmt <- ifelse(
  is.na(meta_timed$bytes), "—",
  ifelse(meta_timed$bytes > 1e6,
    paste0(round(meta_timed$bytes / 1e6, 1), " MB"),
    ifelse(meta_timed$bytes > 1e3,
      paste0(round(meta_timed$bytes / 1e3, 1), " KB"),
      paste0(meta_timed$bytes, " B")
    )
  )
)
DT::datatable(
  meta_timed[, c("name", "seconds", "bytes_fmt")],
  colnames = c("Target", "Seconds", "Size"),
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left;",
    "Build timing for all targets. Seconds = wall-clock build time. ",
    "Size = serialized object size on disk. ",
    "Key: raw_rnaseq and clinical_data dominate build time (GDC downloads). ",
    "Source: targets::tar_meta()."
  ),
  options = list(pageLength = 20, dom = "Bfrtip", order = list(list(1, "desc"))),
  rownames = FALSE
)
```

## Build Status Summary

```{r status-summary}
meta <- targets::tar_meta(store = Sys.getenv("TAR_STORE", "_targets"))
n_total <- nrow(meta)
n_error <- sum(!is.na(meta$error))
n_warn <- sum(!is.na(meta$warnings))
n_timed <- sum(!is.na(meta$seconds))
total_secs <- sum(meta$seconds, na.rm = TRUE)
total_bytes <- sum(meta$bytes, na.rm = TRUE)
```

```{r status-table}
DT::datatable(
  data.frame(
    Metric = c("Total targets", "With timing data", "With errors",
               "With warnings", "Total build time", "Total serialized size"),
    Value = c(
      as.character(n_total),
      as.character(n_timed),
      as.character(n_error),
      as.character(n_warn),
      paste0(round(total_secs / 60, 1), " min"),
      paste0(round(total_bytes / 1e6, 1), " MB")
    ),
    stringsAsFactors = FALSE
  ),
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left;",
    "Pipeline health summary. Total build time is sequential sum; ",
    "actual wall-clock is lower with crew parallelism. ",
    "Source: targets::tar_meta()."
  ),
  options = list(dom = "t", paging = FALSE),
  rownames = FALSE
)
```

## Errors and Warnings

```{r errors}
meta <- targets::tar_meta(store = Sys.getenv("TAR_STORE", "_targets"))
issues <- meta[!is.na(meta$error) | !is.na(meta$warnings),
               c("name", "error", "warnings")]
if (nrow(issues) > 0) {
  DT::datatable(
    issues,
    caption = htmltools::tags$caption(
      style = "caption-side: bottom; text-align: left;",
      "Targets with errors or warnings from the last build. ",
      "Investigate these before relying on downstream targets. ",
      "Source: targets::tar_meta()."
    ),
    options = list(dom = "t", paging = FALSE),
    rownames = FALSE
  )
} else {
  cat("No errors or warnings in the last build.\n")
}
```

## Outdated Targets

Targets that need rebuilding due to changed code or upstream dependencies.

```{r outdated}
tryCatch({
  outdated <- targets::tar_outdated(reporter = "silent")
  if (length(outdated) > 0) {
    DT::datatable(
      data.frame(Target = outdated),
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Targets needing rebuild. Run tar_make() to update. ",
        "Source: targets::tar_outdated()."
      ),
      options = list(dom = "t", paging = FALSE),
      rownames = FALSE
    )
  } else {
    cat("All targets are up to date.\n")
  }
}, error = function(e) {
  cat("*Cannot determine outdated targets: ", conditionMessage(e), "*\n")
})
```

## Size Distribution

```{r size-plot, fig.cap="Target serialized sizes (log10 bytes). Comparison: most targets are under 1 MB; a few data objects (expression matrices, DE results) exceed 1 MB. Dashed line = 1 MB threshold. Source: targets::tar_meta()."}
#| fig.height: 5
meta <- targets::tar_meta(store = Sys.getenv("TAR_STORE", "_targets"))
meta_sized <- meta[!is.na(meta$bytes) & meta$bytes > 0, ]
if (nrow(meta_sized) > 0 && requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(
    meta_sized,
    ggplot2::aes(x = log10(bytes), y = stats::reorder(name, bytes))
  ) +
    ggplot2::geom_col(fill = "steelblue", width = 0.6) +
    ggplot2::geom_vline(xintercept = log10(1e6), linetype = "dashed",
                        color = "red", alpha = 0.5) +
    ggplot2::annotate("text", x = log10(1e6), y = 1, label = "1 MB",
                      hjust = -0.2, color = "red", size = 3) +
    ggplot2::labs(
      title = "Target Sizes (serialized)",
      subtitle = paste0(nrow(meta_sized), " targets with stored objects"),
      x = "log10(bytes)", y = NULL
    ) +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(axis.text.y = ggplot2::element_text(size = 5))
} else {
  cat("*No size data available or ggplot2 not installed.*\n")
}
```
