---
title: "Pipeline Telemetry"
format:
  html:
    code-fold: true
    code-summary: "Show code"
vignette: >
  %\VignetteIndexEntry{Pipeline Telemetry}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
in_pkgdown <- nzchar(Sys.getenv("IN_PKGDOWN"))
knitr::opts_chunk$set(
  eval = !in_pkgdown,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
if (!in_pkgdown) library(targets)

proj_dir <- Sys.getenv("TAR_PROJECT")
tar_store <- NULL
if (proj_dir != "" && dir.exists(file.path(proj_dir, "_targets"))) {
  tar_store <- file.path(proj_dir, "_targets")
} else if (dir.exists("_targets")) {
  tar_store <- "_targets"
} else if (dir.exists("../_targets")) {
  tar_store <- "../_targets"
}

safe_tar_read <- function(name) {
  if (!is.null(tar_store)) {
    val <- tryCatch(tar_read_raw(name, store = tar_store), error = function(e) NULL)
    if (!is.null(val) && !(is.character(val) && length(val) == 1 && grepl("not available", val, ignore.case = TRUE))) {
      return(val)
    }
  }
  rds_candidates <- c(
    system.file(paste0("extdata/vignettes/", name, ".rds"), package = "coMMpass"),
    file.path("inst/extdata/vignettes", paste0(name, ".rds")),
    file.path("../inst/extdata/vignettes", paste0(name, ".rds"))
  )
  for (rds in rds_candidates) {
    if (nzchar(rds) && file.exists(rds)) {
      obj <- readRDS(rds)
      if (is.data.frame(obj) && requireNamespace("DT", quietly = TRUE))
        return(DT::datatable(obj, rownames = FALSE))
      if (!is.null(obj)) return(obj)
    }
  }
  invisible(NULL)
}
```

```{r pkgdown-banner}
#| results: asis
#| eval: !expr in_pkgdown
#| echo: false
cat("::: {.callout-note}\n## Online documentation\nThis vignette displays pre-computed results. Run the targets pipeline locally for interactive analysis.\n:::\n")
```

Pipeline build telemetry from `targets::tar_meta()`. These are metadata
introspection calls (allowed exception to the zero-computation rule).

## Build Timing

Targets ordered by build duration. Long-running targets are candidates
for caching optimization or parallel execution via `crew`.

```{r timing-table}
#| fig.cap: "Target build times. Seconds: wall-clock time per target.
#|   Bytes: serialized object size. Top targets by duration are data
#|   downloads (raw_rnaseq, clinical_data) and DESeq2 paired analysis.
#|   Source: targets::tar_meta()."
meta <- targets::tar_meta(store = Sys.getenv("TAR_STORE", "_targets"))
meta_timed <- meta[!is.na(meta$seconds), c("name", "seconds", "bytes")]
meta_timed$seconds <- as.numeric(meta_timed$seconds)
meta_timed <- meta_timed[order(-meta_timed$seconds), ]
meta_timed$bytes_fmt <- ifelse(
  is.na(meta_timed$bytes), "—",
  ifelse(meta_timed$bytes > 1e6,
    paste0(round(meta_timed$bytes / 1e6, 1), " MB"),
    ifelse(meta_timed$bytes > 1e3,
      paste0(round(meta_timed$bytes / 1e3, 1), " KB"),
      paste0(meta_timed$bytes, " B")
    )
  )
)
DT::datatable(
  meta_timed[, c("name", "seconds", "bytes_fmt")],
  colnames = c("Target", "Seconds", "Size"),
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left;",
    "Build timing for all targets. Seconds = wall-clock build time. ",
    "Size = serialized object size on disk. ",
    "Key: raw_rnaseq and clinical_data dominate build time (GDC downloads). ",
    "Source: targets::tar_meta()."
  ),
  options = list(pageLength = 20, dom = "Bfrtip", order = list(list(1, "desc"))),
  rownames = FALSE
)
```

## Build Status Summary

High-level pipeline health metrics: target counts, build time, and storage size.

```{r status-summary}
meta <- targets::tar_meta(store = Sys.getenv("TAR_STORE", "_targets"))
n_total <- nrow(meta)
n_error <- sum(!is.na(meta$error))
n_warn <- sum(!is.na(meta$warnings))
meta$seconds <- as.numeric(meta$seconds)
meta$bytes <- as.numeric(meta$bytes)
n_timed <- sum(!is.na(meta$seconds))
total_secs <- sum(meta$seconds, na.rm = TRUE)
total_bytes <- sum(meta$bytes, na.rm = TRUE)
```

```{r status-table}
DT::datatable(
  data.frame(
    Metric = c("Total targets", "With timing data", "With errors",
               "With warnings", "Total build time", "Total serialized size"),
    Value = c(
      as.character(n_total),
      as.character(n_timed),
      as.character(n_error),
      as.character(n_warn),
      paste0(round(total_secs / 60, 1), " min"),
      paste0(round(total_bytes / 1e6, 1), " MB")
    ),
    stringsAsFactors = FALSE
  ),
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left;",
    "Pipeline health summary. Total build time is sequential sum; ",
    "actual wall-clock is lower with crew parallelism. ",
    "Source: targets::tar_meta()."
  ),
  options = list(dom = "t", paging = FALSE),
  rownames = FALSE
)
```

## Errors and Warnings

Targets that produced errors or warnings during the most recent pipeline run.

```{r errors}
meta <- targets::tar_meta(store = Sys.getenv("TAR_STORE", "_targets"))
issues <- meta[!is.na(meta$error) | !is.na(meta$warnings),
               c("name", "error", "warnings")]
if (nrow(issues) > 0) {
  DT::datatable(
    issues,
    caption = htmltools::tags$caption(
      style = "caption-side: bottom; text-align: left;",
      "Targets with errors or warnings from the last build. ",
      "Investigate these before relying on downstream targets. ",
      "Source: targets::tar_meta()."
    ),
    options = list(dom = "t", paging = FALSE),
    rownames = FALSE
  )
} else {
  cat("No errors or warnings in the last build.\n")
}
```

## Outdated Targets

Targets that need rebuilding due to changed code or upstream dependencies.

```{r outdated}
tryCatch({
  outdated <- targets::tar_outdated(reporter = "silent")
  if (length(outdated) > 0) {
    DT::datatable(
      data.frame(Target = outdated),
      caption = htmltools::tags$caption(
        style = "caption-side: bottom; text-align: left;",
        "Targets needing rebuild. Run tar_make() to update. ",
        "Source: targets::tar_outdated()."
      ),
      options = list(dom = "t", paging = FALSE),
      rownames = FALSE
    )
  } else {
    cat("All targets are up to date.\n")
  }
}, error = function(e) {
  cat("*Cannot determine outdated targets: ", conditionMessage(e), "*\n")
})
```

## Size Distribution

Distribution of target serialized sizes, identifying large objects that dominate storage.

```{r size-plot, fig.cap="Target serialized sizes (log10 bytes). Comparison: most targets are under 1 MB; a few data objects (expression matrices, DE results) exceed 1 MB. Dashed line = 1 MB threshold. Source: targets::tar_meta()."}
#| fig.height: 5
meta <- targets::tar_meta(store = Sys.getenv("TAR_STORE", "_targets"))
meta$bytes <- as.numeric(meta$bytes)
meta_sized <- meta[!is.na(meta$bytes) & meta$bytes > 0, ]
if (nrow(meta_sized) > 0 && requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(
    meta_sized,
    ggplot2::aes(x = log10(bytes), y = stats::reorder(name, bytes))
  ) +
    ggplot2::geom_col(fill = "steelblue", width = 0.6) +
    ggplot2::geom_vline(xintercept = log10(1e6), linetype = "dashed",
                        color = "red", alpha = 0.5) +
    ggplot2::annotate("text", x = log10(1e6), y = 1, label = "1 MB",
                      hjust = -0.2, color = "red", size = 3) +
    ggplot2::labs(
      title = "Target Sizes (serialized)",
      subtitle = paste0(nrow(meta_sized), " targets with stored objects"),
      x = "log10(bytes)", y = NULL
    ) +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(axis.text.y = ggplot2::element_text(size = 5))
} else {
  cat("*No size data available or ggplot2 not installed.*\n")
}
```

## Git Activity

### Commit Velocity

Weekly commit counts colored by conventional-commit type (feat, fix, docs, etc.)
across the last 200 commits.

```{r commit-velocity, fig.cap="Weekly commit velocity colored by change type. Each bar segment represents commits of a given type (feat, fix, docs, ci, etc.) in that ISO week. Source: git log.", fig.height=5}
safe_tar_read("vig_commit_velocity")
```

### Changes by Type

Commits grouped by conventional-commit prefix, with total lines added and removed.

```{r changes-by-type}
summary_data <- safe_tar_read("vig_git_changelog_summary")
if (!is.null(summary_data) && !is.null(summary_data$by_type)) {
  DT::datatable(
    summary_data$by_type,
    caption = htmltools::tags$caption(
      style = "caption-side: bottom; text-align: left;",
      "Commits by conventional-commit type with aggregate line counts. ",
      "Source: git log --numstat (last 200 commits)."
    ),
    rownames = FALSE,
    options = list(dom = "t", paging = FALSE)
  )
} else {
  cat("*Changelog summary not available.*\n")
}
```

### Changes by File Category

How often each file category (R Source, Tests, Vignettes, CI/CD, etc.) was
touched across the last 200 commits.

```{r changes-by-file-category}
if (!is.null(summary_data) && !is.null(summary_data$by_file_category)) {
  DT::datatable(
    summary_data$by_file_category,
    caption = htmltools::tags$caption(
      style = "caption-side: bottom; text-align: left;",
      "Commits touching each file category. A single commit may touch ",
      "multiple categories. Source: git log --numstat (last 200 commits)."
    ),
    rownames = FALSE,
    options = list(dom = "t", paging = FALSE)
  )
} else {
  cat("*File category data not available.*\n")
}
```

### Recent Commits

Full 20-commit changelog with lines added, files changed, and change categories.

```{r recent-commits}
safe_tar_read("vig_git_changelog")
```

## GitHub Activity

### Issues & Pull Requests

Open/closed issue and merged PR counts from the GitHub API.

```{r github-activity}
gh_data <- safe_tar_read("vig_github_activity")
if (!is.null(gh_data) && is.null(gh_data$error)) {
  DT::datatable(
    data.frame(
      Metric = c("Open Issues", "Closed Issues", "Merged PRs"),
      Count = c(gh_data$issues_open, gh_data$issues_closed, gh_data$prs_merged),
      stringsAsFactors = FALSE
    ),
    caption = htmltools::tags$caption(
      style = "caption-side: bottom; text-align: left;",
      "GitHub issue and PR counts. Source: GitHub API via gh::gh()."
    ),
    rownames = FALSE,
    options = list(dom = "t", paging = FALSE)
  )
} else if (!is.null(gh_data$error)) {
  cat(paste0("*GitHub API error: ", gh_data$error, "*\n"))
} else {
  cat("*GitHub activity data not available.*\n")
}
```

### Codebase Metrics

R files, test files, exports, lines of code, and package version.

```{r codebase-metrics}
safe_tar_read("vig_codebase_metrics")
```

## Reproducibility

<details>
<summary>Session Info (click to expand)</summary>

```{r session-info, eval=TRUE}
sessionInfo()
```

</details>
